---
title: "Batch correction: diagnose, correct, verify"
author: "Sergio Ciordia"
package: NADIA
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Batch correction: diagnose, correct, verify}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", crop = NULL,
                      dev = "svg")
```

<style>
.bert-warning {
  width: calc(100% - 4.5em);
  box-sizing: border-box;
  margin: 1.15em 0 1.5em;
  padding: 0.9em 1.1em;
  border: 1px solid #e4c78e;
  border-left: 5px solid #c98718;
  border-radius: 5px;
  background: #fffaf0;
  box-shadow: 0 1px 2px rgba(112, 78, 20, 0.08);
}
.bert-warning p {
  margin: 0;
  padding: 0 !important;
  text-align: justify;
}
.bert-warning strong:first-child {
  color: #7a5310;
}
.batch-argument-table {
  width: calc(100% + 3em);
  margin: 1em 0 1.35em -1.5em;
}
.batch-argument-table table {
  width: 100% !important;
  table-layout: fixed;
}
.batch-argument-table col:first-child {
  width: 29% !important;
}
.batch-argument-table col:nth-child(2) {
  width: 71% !important;
}
.batch-argument-table th:first-child,
.batch-argument-table td:first-child {
  width: 29%;
  padding-right: 1.25em !important;
}
.batch-argument-table th:nth-child(2),
.batch-argument-table td:nth-child(2) {
  width: 71%;
  padding-left: 1.25em !important;
}
.batch-argument-table th,
.batch-argument-table td {
  vertical-align: top;
}
.lightbox-figure {
  width: calc(100% - 1em);
  margin: 1.25em 0 1.5em;
  text-align: center;
}
.lightbox-figure img {
  width: 100%;
  height: auto;
  cursor: zoom-in;
}
#diagram-lightbox {
  display: none;
  position: fixed;
  z-index: 9999;
  inset: 0;
  align-items: center;
  justify-content: center;
  padding: 2vh 2vw;
  background: rgba(10, 18, 25, 0.9);
}
#diagram-lightbox.is-open {
  display: flex;
}
#diagram-lightbox p {
  display: flex;
  align-items: center;
  justify-content: center;
  margin: 0;
}
#diagram-lightbox img {
  max-width: 96vw;
  max-height: 94vh;
  padding: 0.4em;
  background: #ffffff;
  box-shadow: 0 6px 30px rgba(0, 0, 0, 0.35);
  cursor: default;
}
#diagram-lightbox button {
  position: absolute;
  top: 12px;
  right: 18px;
  border: 0;
  background: transparent;
  color: #ffffff;
  font-size: 36px;
  line-height: 1;
  cursor: pointer;
}
body.diagram-lightbox-open {
  overflow: hidden;
}
</style>

# Introduction

Samples prepared on different days, labelled in different batches or acquired in
different run sequences may contain systematic technical differences. Batch
correction aims to remove this unwanted variation while retaining the biological
signal. If the experimental design is confounded or the correction is applied
without adequate checks, biological variation may also be removed.

[`BERT`](https://doi.org/10.1038/s41467-025-62237-4) is fully integrated into
NADIA, and all matrix-level batch correction in this vignette is performed
through it. BERT is designed for the integration of incomplete omic profiles.
Rather than requiring a single complete matrix, it decomposes the integration
into pairwise batch-correction steps
arranged as a binary tree. Each step applies the selected correction method,
such as ComBat or limma, and the corrected branches are then combined. This
hierarchical strategy allows NADIA to correct the normalised protein-abundance
matrix before imputation, without first replacing its missing values.

This vignette therefore follows a simple rule: **diagnose before correcting and
verify afterwards**. Using a simulated batch effect, it shows how to assess the
design and technical variation, choose the variables to remove or protect, apply
and verify BERT, and distinguish matrix correction from covariate adjustment in
the differential abundance model. It also explains why later principal
components and sequential correction require particular care.

The same workflow is then applied to a real two-mix TMT experiment. Its known
three-proteome spike-in design allows uncorrected, corrected and
condition-protected results to be evaluated by variance decomposition, PCA and
differential abundance benchmarking. The example highlights the effects on
sensitivity and specificity, together with the importance of the fold-change
threshold and TMT ratio compression.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("NADIA")
```

```{r library, message = FALSE}
library(NADIA)
data(nadia_dia)
```

Batch correction needs `BERT` and the variance decomposition needs `lme4`; both
are optional dependencies:

```{r deps}
have_bert <- requireNamespace("BERT", quietly = TRUE)
have_lme4 <- requireNamespace("lme4", quietly = TRUE)
c(BERT = have_bert, lme4 = have_lme4)
```

```{r flags, include = FALSE}
run_bc  <- have_bert
run_all <- have_bert && have_lme4
```

# The sample annotation

The example dataset does not contain a recorded batch variable. To demonstrate
the workflow, we create a hypothetical sample annotation with two digestion
days, two injection sequences and four subjects. These variables are arranged so
that each condition is represented at every level of the technical factors.

In a real analysis, these variables should come from the experimental sample
sheet rather than being inferred from the quantitative values. The `Column`
field must match the sample identifiers used by the pipeline:

```{r covariates}
md <- nadia_dia$metadata

covariates <- data.frame(
    Column    = md$Coding,
    Digestion = rep(c("d1", "d2"),             length.out = nrow(md)),
    Injection = rep(c("i1", "i1", "i2", "i2"), length.out = nrow(md)),
    Subject   = rep(paste0("s", 1:4),          length.out = nrow(md)),
    stringsAsFactors = FALSE)

covariates
```

## A simulated batch effect

The annotation defines the groups but does not create a batch effect. We
therefore **introduce a simulated effect** so that its removal can be evaluated
against a known starting point.

The simulated effect is deliberately protein-dependent: in the second digestion
batch, half of the proteins are multiplied by six and the remainder by one
sixth. A uniform shift affecting every protein similarly may be largely removed
by normalisation, whereas protein-specific technical effects can remain and may
require explicit batch correction.

The six-fold change is intentionally large. The biological conditions in this
dataset are already strongly separated, and a smaller simulated batch effect
would be difficult to see in the diagnostic figures. Its magnitude is chosen for
illustration and should not be interpreted as a typical batch effect.

```{r inject}
batched <- nadia_dia
qcols   <- grep("^PG.Quantity_", colnames(batched$protein_quant))
samples <- sub("^PG.Quantity_", "", colnames(batched$protein_quant)[qcols])
d2_cols <- qcols[samples %in% covariates$Column[covariates$Digestion == "d2"]]

set.seed(1)
up     <- sample(c(TRUE, FALSE), nrow(batched$protein_quant), replace = TRUE)
batch_multiplier <- ifelse(up, 6, 1 / 6)
batched$protein_quant[, d2_cols] <-
    batched$protein_quant[, d2_cols] * batch_multiplier

length(d2_cols)   # the six samples digested on day 2
```

All subsequent results are calculated from this modified dataset.

# Step 1: is the design suitable for batch correction?

Before choosing a correction model, cross-tabulate the biological and technical
factors and count the samples in each combination:

```{r design}
table(Digestion = covariates$Digestion, Injection = covariates$Injection)
table(Condition = md$R.Condition,       Digestion = covariates$Digestion)
```

Both tables are complete: every digestion day contains both injection sequences,
and every condition occurs on both digestion days. The design therefore provides
information with which to distinguish the simulated digestion effect from the
biological conditions.

If a condition occurred in only one digestion batch, condition and batch would
be **completely confounded**. Their effects could not be estimated separately:
the correction might fail, or it might remove part of the biological difference
with the batch effect. These tables are therefore an essential first diagnostic,
although they should be followed by the post-correction checks shown below.

::: {.bert-warning}
**Minimum replication required by BERT.** BERT requires at least two samples in
every batch. When `batch_covariates` are supplied, this requirement applies to
each combination of batch and protected-covariate levels. In addition, a protein
must have at least two non-missing quantitative values within each of these
combinations for BERT to estimate its correction. Insufficient replication can
make individual proteins unadjustable and, when it affects many proteins, may
prevent the batch-correction model from being fitted.
:::

# Step 2: how much variance does the batch explain?

Before correcting the batch effect, we first process the modified dataset with
batch correction disabled. This provides the normalised and imputed assay that
will serve as the uncorrected reference. We can then use `pvca_analysis()` to
estimate how much of its variation is associated with the biological condition
and how much is associated with the digestion and injection factors.

```{r process, eval = run_bc}
res <- process_proteomics(batched, covariate_df = covariates, verbose = FALSE)
```

`pvca_analysis()` uses the selected principal components to estimate the share of
variation associated with the specified biological and technical factors. With
the default settings, it reports the following result:

```{r pvca-default, eval = run_all, message = FALSE, warning = FALSE}
pvca_default <- pvca_analysis(
    res$se_proc,
    assay_name         = "Impseqrob_min",
    technical_factors  = c("Digestion", "Injection"),
    biological_factors = "Condition",
    verbose            = FALSE)

pvca_default$variance_components
```

Condition accounts for almost all of the reported variation, while digestion is
absent from the table. This is surprising because the simulated digestion effect
is known to be present.

## Why the default hides it, and what to do

`pca_threshold` determines how many principal components are included: components
are retained until their cumulative explained variance reaches the selected
threshold, which is 60% by default. In this experiment, the biological conditions
are strongly separated and PC1 alone exceeds 60%. The simulated digestion effect
is represented by later components, which are excluded from the default
decomposition.

Raise the threshold and it appears:

```{r pvca-before, eval = run_all, message = FALSE, warning = FALSE}
pvca_before <- pvca_analysis(
    res$se_proc,
    assay_name         = "Impseqrob_min",
    technical_factors  = c("Digestion", "Injection"),
    biological_factors = "Condition",
    pca_threshold      = 0.95,
    variance_threshold = 0,
    verbose            = FALSE)

pvca_before$variance_components
```

After increasing the threshold, digestion accounts for **27.4%** of the weighted
variation and condition for **70.2%**. The batch effect does not need to dominate
the dataset to influence the downstream analysis. These proportions describe
how the retained components are partitioned; they do not define a universal
threshold above which correction is required.

A decomposition that reports no technical contribution is therefore not, by
itself, evidence that the data contain no technical structure. Always check how
many principal components were retained and whether later components contain a
recognisable batch pattern.

```{r pvca-before-plot, eval = run_all, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, fig.cap = "PVCA before correction, using enough principal components to explain 95% of the variance.", fig.alt = "Variance decomposition before batch correction"}
pvca_before$plot
```

# Step 3: correct, and keep the biology

Batch correction is optional and is disabled by default. To apply it within
`process_proteomics()`, set `batch_correct = TRUE` and use `batch_column` to
identify the technical factor that should be removed. Two related arguments,
`covariate_column` and `bio_replicate_column`, describe variables that should be
modelled during differential abundance analysis rather than removed from the
quantitative matrix. The three arguments therefore have distinct roles and
should be defined according to the experimental design:

::: {.batch-argument-table}
| Argument | Effect |
|---|---|
| `batch_column` | identifies the technical factor to be **removed from the quantitative matrix** |
| `covariate_column` | adds one or more variables to the **limma differential abundance model** |
| `bio_replicate_column` | defines a **blocking factor** for samples that are not independent |
:::

Here, `batch_column = "Digestion"` asks BERT/ComBat to estimate and remove the
digestion-day effect. By contrast, `covariate_column = "Injection"` does not
alter the quantitative matrix: it includes injection sequence as a term in the
subsequent limma model. Finally, `bio_replicate_column = "Subject"` uses
`limma::duplicateCorrelation()` to account for the correlation between samples
from the same subject rather than treating all samples as independent.

The diagram below places the three arguments in the complete workflow and shows
typical variables that can be supplied to each one. Click the figure to enlarge
it.

::: {.batch-arguments-figure .lightbox-figure}
```{r batch-arguments-diagram, echo = FALSE, out.width = "100%", fig.cap = "The three arguments act at different stages: batch_column modifies the quantitative matrix, whereas covariate_column and bio_replicate_column act during differential abundance analysis.", fig.alt = "Workflow diagram comparing batch_column, covariate_column and bio_replicate_column. Batch_column removes a technical effect from the normalised matrix before imputation. Covariate_column adds nuisance variables to the limma model without changing the matrix. Bio_replicate_column models correlation between paired, repeated or technical replicate samples."}
knitr::include_graphics("figures/Batch_Correction_Arguments.svg")
```
:::

<div id="diagram-lightbox" role="dialog" aria-modal="true" aria-label="Enlarged diagram">
<button type="button" aria-label="Close enlarged diagram">&times;</button>
<img alt="Enlarged diagram" />
</div>

<script>
(function () {
  function initialiseDiagramLightbox() {
    var figures = document.querySelectorAll('.lightbox-figure img');
    var lightbox = document.getElementById('diagram-lightbox');
    if (!figures.length || !lightbox) return;
    var enlarged = lightbox.querySelector('img');
    var closeButton = lightbox.querySelector('button');
    function closeLightbox() {
      lightbox.classList.remove('is-open');
      document.body.classList.remove('diagram-lightbox-open');
    }
    figures.forEach(function (figure) {
      figure.addEventListener('click', function () {
        enlarged.src = figure.src;
        enlarged.alt = figure.alt || 'Enlarged diagram';
        lightbox.classList.add('is-open');
        document.body.classList.add('diagram-lightbox-open');
        closeButton.focus();
      });
    });
    closeButton.addEventListener('click', closeLightbox);
    lightbox.addEventListener('click', function (event) {
      if (event.target === lightbox) closeLightbox();
    });
    document.addEventListener('keydown', function (event) {
      if (event.key === 'Escape') closeLightbox();
    });
  }
  if (document.readyState === 'loading') {
    document.addEventListener('DOMContentLoaded', initialiseDiagramLightbox);
  } else {
    initialiseDiagramLightbox();
  }
})();
</script>

```{r correct, eval = run_bc, message = FALSE, warning = FALSE, results = "hide"}
res_bc <- process_proteomics(
    batched,
    covariate_df         = covariates,
    batch_correct        = TRUE,
    batch_column         = "Digestion",
    batch_algorithm      = "ComBat",
    batch_ComBat_mode    = 1,
    covariate_column     = "Injection",
    bio_replicate_column = "Subject",
    verbose              = FALSE)
```

```{r correct-assays, eval = run_bc}
SummarizedExperiment::assayNames(res_bc$se_proc)
```

Note where the new `BERT` assay sits: after `cycloess`, before the imputed
matrix. Correction runs on the normalised data, and imputation then runs on its
output.

## Protecting a variable inside ComBat

`batch_column` tells BERT/ComBat **which technical grouping should be removed**.
`batch_covariates` tells it **which biological groupings should be protected
while that correction is estimated**. For example, using
`batch_column = "Batch"` together with `batch_covariates = "Condition"` asks the
model to remove differences between batches without treating condition-related
variation as part of the batch effect.

This still differs from `covariate_column`. A protected `batch_covariates`
variable enters the BERT/ComBat model and influences the corrected matrix.
`covariate_column`, by contrast, is used later in the limma model and never
changes that matrix. For this reason, `batch_correct_proteomics()` warns when
ComBat or limma correction is requested without a protected covariate.

::: {.bert-warning}
**The default does not imply the recommended choice.**
`batch_covariates = NULL` is the default because NADIA cannot determine
automatically which metadata variable represents the biological signal that
should be preserved. In most experiments with a known biological grouping and
sufficient replication, that variable should be supplied explicitly. NADIA
therefore warns when ComBat or limma correction is performed without a
protected covariate.
:::

The appropriate choice depends on both the biological question and the design.
The following guide summarises when a protected covariate should be supplied,
when `NULL` may be reasonable, and when the design should be reconsidered rather
than forcing an unprotected correction. Click the figure to enlarge it.

::: {.batch-covariates-figure .lightbox-figure}
```{r batch-covariates-decision, echo = FALSE, out.width = "100%", fig.cap = "Choosing batch_covariates. A known biological variable should normally be protected when batch and biology can be estimated separately. NULL is not a remedy for confounding or insufficient replication.", fig.alt = "Decision guide for batch_covariates. Use a protected covariate for a known biological variable when the design is estimable. Reassess the design rather than falling back to NULL when batch and biology are confounded or replication is insufficient. NULL may be appropriate when no biological variable needs protection or when all batch-associated variation is deliberately removed."}
knitr::include_graphics("figures/Batch_Covariates_Decision.svg")
```
:::

The following template shows how the four roles can be combined. `Condition` is
already obtained from the `proteomics_data` metadata, whereas the additional
technical and subject variables are supplied through `covariate_df`:

```{r protected-batch-example, eval = FALSE}
# The sample annotation contains Column, Batch, Injection and Subject.
sample_annotation <- read.delim("sample_annotation.tsv")

# Check that Condition is represented at least twice in every batch.
design_check <- merge(
    data.frame(
        Column    = my_data$metadata$Coding,
        Condition = my_data$metadata$R.Condition),
    sample_annotation,
    by = "Column")

batch_condition_counts <- with(
    design_check,
    table(Batch = Batch, Condition = Condition))
batch_condition_counts
stopifnot(all(batch_condition_counts >= 2))

res_bc <- process_proteomics(
    my_data,
    covariate_df         = sample_annotation,
    batch_correct        = TRUE,
    batch_column         = "Batch",       # remove from the matrix
    batch_covariates     = "Condition",   # protect during correction
    covariate_column     = "Injection",   # adjust the limma model
    bio_replicate_column = "Subject")     # paired/repeated samples
```

In this call, only `Batch` is removed from the quantitative matrix. `Condition`
is included in the batch-correction model so that its biological differences are
retained. `Injection` and `Subject` do not participate in BERT/ComBat: they are
used later during differential abundance analysis.

In the simulated example used in this vignette, `Condition` is not protected
because the small batch-by-condition groups do not provide enough usable
observations for BERT to fit the protected model. The unprotected correction is
acceptable here only because the simulated digestion effect is balanced across
conditions and its removal can be evaluated against a known result.

# Step 4: verify

## Did the batch effect go?

Because the batch effect was introduced deliberately, its removal can be measured
directly. For each protein, we calculate the difference between its mean
intensity on the two digestion days. Both the typical absolute difference and
the spread of these differences should decrease after correction:

```{r verify-direct, eval = run_bc}
grp <- covariates$Digestion[
    match(colnames(SummarizedExperiment::assay(res_bc$se_proc, "cycloess")),
          covariates$Column)]

delta <- function(assay_name) {
    m <- SummarizedExperiment::assay(res_bc$se_proc, assay_name)
    rowMeans(m[, grp == "d1", drop = FALSE], na.rm = TRUE) -
    rowMeans(m[, grp == "d2", drop = FALSE], na.rm = TRUE)
}

data.frame(
    assay      = c("cycloess (before)", "BERT (after)"),
    median_abs_difference = c(median(abs(delta("cycloess")), na.rm = TRUE),
                              median(abs(delta("BERT")),     na.rm = TRUE)),
    sd         = c(sd(delta("cycloess"), na.rm = TRUE),
                   sd(delta("BERT"),     na.rm = TRUE)))
```

Both summaries fall markedly, indicating that the simulated between-day effect
has been substantially reduced.

## Did the biology survive?

Removing the technical pattern is only half of the assessment. Repeat the
decomposition on the corrected assay using the same threshold to determine
whether the biological structure remains visible:

```{r pvca-after, eval = run_all, message = FALSE, warning = FALSE}
pvca_after <- pvca_analysis(
    res_bc$se_proc,
    assay_name         = "BERT",
    technical_factors  = c("Digestion", "Injection"),
    biological_factors = "Condition",
    pca_threshold      = 0.95,
    variance_threshold = 0,
    verbose            = FALSE)

pvca_comparison <- merge(
    pvca_before$variance_components,
    pvca_after$variance_components,
    by = "label",
    suffixes = c("_before", "_after"))[, c(1, 2, 4)]

# Use fixed decimal notation so that the proportions can be read directly.
pvca_comparison$weights_before <- formatC(
    pvca_comparison$weights_before, format = "f", digits = 8)
pvca_comparison$weights_after <- formatC(
    pvca_comparison$weights_after, format = "f", digits = 8)

pvca_comparison
```

The estimated digestion contribution falls from 27.4% to a value close to zero,
while the relative share assigned to condition rises from 70.2% to 77.5%. This
is the desired direction: the known technical structure is removed while the
biological conditions remain prominent.

PVCA reports relative shares, so the increase in the condition proportion does
not prove by itself that every biological effect was preserved. It may partly
reflect a smaller total after removal of the batch-associated variance. This is
why the decomposition should be combined with direct checks of known biological
signals and with inspection of the differential abundance results.

## What it changed in the results

```{r de-effect, eval = run_bc}
table(uncorrected = res$DEPs_results$Change)
table(corrected   = res_bc$DEPs_results$Change)
```

In this simulated example, the batch effect increases residual variability and
reduces statistical power. After correction, the number of significant
protein–comparison results rises markedly.

This count cannot establish whether the correction was successful: both improved
power and an over-aggressive correction can increase the number of significant
results. A reduction in the technical contribution together with retention of
the biological structure is reassuring, whereas a simultaneous loss of both is
a warning sign. Whenever known controls or expected changes are available, they
provide a stronger test than the number of significant calls alone.

# Removing two factors, one after the other

`batch_correct_proteomics()` corrects one factor per call. Because each call can
read one assay and write the corrected values to a new assay, corrections can be
applied sequentially. The first call below starts from the normalised
`cycloess` assay; the second uses the first corrected assay as its input:

```{r cascade, eval = run_bc, message = FALSE, warning = FALSE, results = "hide"}
se1 <- batch_correct_proteomics(
    res$se_proc,
    assay_name           = "cycloess",
    batch_column         = "Digestion",
    corrected_assay_name = "BERT",
    verbose              = FALSE)

se2 <- batch_correct_proteomics(
    se1,
    assay_name           = "BERT",       # the already-corrected matrix
    batch_column         = "Injection",
    corrected_assay_name = "BERT_2",
    verbose              = FALSE)
```

```{r cascade-assays, eval = run_bc}
SummarizedExperiment::assayNames(se2)
```

Sequential correction requires care. Each step uses degrees of freedom, and the
second model is fitted to values already modified by the first. The order can
also affect the result when the technical factors are associated. In this
simulated design, digestion and injection are balanced with respect to one
another, as shown in step 1. After the second correction, imputation should be
applied to `BERT_2`, not to the earlier assays.

::: {.bert-warning}
**Risk of over-correction.** Applying BERT sequentially to several technical
factors can overfit the data or remove genuine biological signal because every
step modifies a matrix that has already been corrected. A more conservative
option is to use `batch_column` for the dominant technical effect and include
the remaining factor in `covariate_column`, so that limma adjusts for it without
altering the quantitative matrix again. In this example, that means correcting
`Digestion` and modelling `Injection`. This strategy reduces, but does not
eliminate, the need to check design balance and verify the results.
:::

# Inspecting the correction with PCA

PVCA summarises the contributions numerically, whereas PCA shows how the samples
are arranged. `pca_covariates_plot()` displays the same projection coloured by
each selected covariate, making it easier to identify factors associated with
sample separation:

```{r pca-cov, eval = run_bc, message = FALSE, fig.width = 7, fig.height = 6, fig.cap = "PCA of the simulated example after correction, coloured in turn by each biological or technical variable.", fig.alt = "PCA coloured by condition, digestion, injection and subject"}
pca_grid <- pca_covariates_plot(
    res_bc$se_proc,
    assay_name = "BERT",
    covariates = c("Condition", "Digestion", "Injection", "Subject"),
    scale.     = TRUE,
    verbose    = FALSE)

pca_grid$grid +
    ggplot2::theme(panel.spacing = grid::unit(1.2, "lines"))
```

After a successful correction, samples should retain their biological grouping
while no longer separating systematically by the corrected technical factor. If
strong digestion-related clustering remains, reconsider the annotation, the
model and the selected correction method before proceeding. Sequentially
removing an additional factor is appropriate only when that second factor also
has a justified and independently assessable technical effect.

# Applying the workflow to a real batch effect

The simulated example made it possible to verify removal of an effect whose size
and direction were known. The TMT example provides a complementary test: its two
labelling mixes represent a real technical batch, while the three-proteome
spike-in design provides expected biological changes against which the results
can be evaluated.

## The batch has to be declared, not discovered

`nadia_tmt_report.tsv.gz` contains samples labelled in two TMTpro mixes:
replicates 1–4 of each condition belong to the first mix and replicates 5–8 to
the second. The separate labelling reactions and acquisitions create a plausible
source of technical variation.

The diagram below combines the biological spike-in composition with the channel
allocation across the two TMTpro mixes. Click the figure to enlarge it.

::: {.tmt-design-figure .lightbox-figure}
```{r tmt-design-diagram, echo = FALSE, out.width = "100%", fig.cap = "Three-proteome TMT design. Conditions A, B and D retain an 80% human background while the proportions of *E. coli* and yeast vary. Replicates 1–4 of each condition are assigned to mix 1 and replicates 5–8 to mix 2.", fig.alt = "Experimental design of the TMT spike-in study. Condition A contains 80 percent human, 5 percent E. coli and 15 percent yeast; condition B contains 80, 10 and 10 percent; condition D contains 80 percent human, 20 percent E. coli and no yeast. Mix 1 contains replicates 1 to 4 of A, B and D plus internal standards 1 and 2. Mix 2 contains replicates 5 to 8 plus internal standards 3 and 4."}
knitr::include_graphics("figures/TMT_Batch_Experimental_Design.svg")
```
:::

The Proteome Discoverer report does not contain a column identifying the TMT mix,
so `preprocess_tmt()` cannot add this variable automatically. It must be supplied
to `process_proteomics()` in a data frame whose `Column` values match the sample
identifiers. In a real analysis, the assignments should come from the sample
sheet. For this example, they can be reconstructed from the replicate numbers:

```{r tmt-data}
tmt <- preprocess_tmt(
    system.file("extdata", "nadia_tmt_report.tsv.gz", package = "NADIA"),
    condition_order = c("A", "B", "D"), verbose = FALSE)

tmt_cov <- data.frame(
    Column = tmt$metadata$Coding,
    Mix    = ifelse(tmt$metadata$R.Replicate <= 4, "mix1", "mix2"),
    stringsAsFactors = FALSE)

table(Condition = tmt$metadata$R.Condition, Mix = tmt_cov$Mix)
```

Every condition is represented by four samples in each mix. This balanced design
allows the mix effect to be estimated separately from condition and is important
for interpreting what changes after correction.

## The decomposition finds it

We first process the TMT data without batch correction to establish the
reference assay used for the initial variance decomposition:

```{r tmt-process, eval = run_bc, message = FALSE}
tmt_res <- process_proteomics(tmt, covariate_df = tmt_cov, verbose = FALSE)
```

```{r tmt-pvca, eval = run_all, message = FALSE, warning = FALSE}
tmt_pvca_before <- pvca_analysis(
    tmt_res$se_proc,
    assay_name         = "Impseqrob_min",
    technical_factors  = "Mix",
    biological_factors = "Condition",
    pca_threshold      = 0.95,
    variance_threshold = 0,
    verbose            = FALSE)

tmt_pvca_before$variance_components
```

Before correction, TMT mix accounts for 69.1% of the weighted variation and
condition for 30.5%. Unlike the preceding simulation, no numerical shift was
introduced artificially: this structure is present in the example report.

## Correcting it, and checking against biology

To make the role of `batch_covariates` explicit, we first correct the mix effect
without protecting a biological variable:

```{r tmt-correct, eval = run_bc, message = FALSE, warning = FALSE, results = "hide"}
tmt_bc <- process_proteomics(
    tmt, covariate_df = tmt_cov,
    batch_correct = TRUE, batch_column = "Mix",
    batch_algorithm = "ComBat", batch_ComBat_mode = 1,
    verbose = FALSE)
```

With `batch_covariates = NULL`, no biological variable is explicitly protected.
In most experiments, `Condition` should be supplied so that ComBat does not
attribute condition-associated differences to the batch effect. This example
provides a useful comparison because every condition has four samples in each
mix: `Condition` and `Mix` are orthogonal at the level of sample allocation, so
the protected and unprotected corrections are expected to be similar. We test
that expectation by repeating the correction with `Condition` protected:

```{r tmt-correct-protected, eval = run_bc, message = FALSE, warning = FALSE, results = "hide"}
tmt_bc_prot <- process_proteomics(
    tmt, covariate_df = tmt_cov,
    batch_correct = TRUE, batch_column = "Mix",
    batch_covariates = "Condition",
    batch_algorithm = "ComBat", batch_ComBat_mode = 1,
    verbose = FALSE)
```

The protected and unprotected results are compared with the known spike-in
changes below. Cross-tabulation of the design remains the first diagnostic: it
shows whether batch and condition can be estimated separately and whether the
protected model has sufficient replication. It cannot, however, establish that
the correction preserved the biological signal, so post-correction checks are
still required.

Repeating PVCA on the unprotected corrected assay verifies whether the
mix-associated component was removed:

```{r tmt-pvca-after, eval = run_all, message = FALSE, warning = FALSE}
tmt_pvca_after <- pvca_analysis(
    tmt_bc$se_proc,
    assay_name         = "BERT",
    technical_factors  = "Mix",
    biological_factors = "Condition",
    pca_threshold      = 0.95,
    variance_threshold = 0,
    verbose            = FALSE)

tmt_pvca_comparison <- merge(
    tmt_pvca_before$variance_components,
    tmt_pvca_after$variance_components,
    by = "label",
    suffixes = c("_before", "_after"))[, c(1, 2, 4)]

tmt_pvca_comparison$weights_before <- formatC(
    tmt_pvca_comparison$weights_before, format = "f", digits = 8)
tmt_pvca_comparison$weights_after <- formatC(
    tmt_pvca_comparison$weights_after, format = "f", digits = 8)

tmt_pvca_comparison
```

The estimated contribution of mix falls from 69.1% to a value close to zero,
while the relative share assigned to condition rises from 30.5% to 96.8%. This
confirms removal of the visible mix structure, but it does not by itself show
that the biological signal was preserved. The spike-in expectations below
provide a more direct assessment of the consequences for biological inference.

This report is a three-proteome spike-in: *E. coli* increases from condition A to
D, yeast decreases, and the human background remains constant at 80%. Therefore,
for the `D-A` comparison the expected directions are known: quantified *E. coli*
proteins should increase, yeast proteins should decrease, and human proteins
should remain unchanged.

The organism name is stored in the UniProt `OS=` field of the protein description.
The TMT preprocessor does not create a separate organism column, so it is parsed
here for this diagnostic:

```{r tmt-verify, eval = run_bc}
species <- sub(".*OS=(.+?)\\s+[A-Za-z]+=.*", "\\1",
               tmt$protein_id$PG.ProteinDescriptions)
names(species) <- tmt$protein_id$PG.ProteinGroups

directional_performance <- function(res, comparison) {
    d  <- res$DEPs_results
    d  <- d[d$Comparison == comparison, ]
    sp <- species[match(d$Protein.IDs, names(species))]
    c(`E. coli called up` = mean(d$Change[grepl("^Escherichia", sp)] == "Up"),
      `yeast called down` = mean(d$Change[grepl("^Saccharomyces", sp)] == "Down"),
      `human unchanged`   = mean(d$Change[grepl("^Homo", sp)] == "No Change"))
}

round(100 * cbind(
    uncorrected = directional_performance(tmt_res,     "D-A"),
    corrected   = directional_performance(tmt_bc,      "D-A"),
    protected   = directional_performance(tmt_bc_prot, "D-A")), 1)
```

Correction increases recovery in the expected direction from 78.8% to 97.4% for
*E. coli* and from 84.2% to 95.6% for yeast. However, the percentage of human
proteins correctly left unchanged falls from 89.0% to 38.2%. The batch effect has
been removed and sensitivity has increased, but specificity against the known
constant background has deteriorated. This trade-off would be missed if only the
number of recovered spike-in proteins were reported. It is also measured at the
default fold-change threshold of zero; the section *Significance is not effect
size* below shows how much of it depends on that choice.

Protecting `Condition` produces very similar values: 97.8% rather than 97.4% for
*E. coli*, 96.1% rather than 95.6% for yeast, and 36.6% rather than 38.2% for the
human background. This agreement is consistent with the balanced allocation of
every condition across both mixes. It should not be generalised: when conditions
are unevenly distributed across batches, an unprotected ComBat correction can
remove part of the biological difference together with the batch effect. In
that setting, **the biological condition should be protected whenever the design
allows it to be estimated**.

## Visual assessment with PCA

The PCA projections below show the same samples before and after the unprotected
correction, coloured separately by condition and TMT mix. This assay is shown
because the directional benchmark above found very similar biological
performance for the protected and unprotected corrections in this balanced
design:

```{r tmt-pca-before, eval = run_bc, message = FALSE, fig.width = 8.5, fig.height = 4, fig.cap = "TMT experiment before correction. The same PCA is coloured by biological condition and TMT mix.", fig.alt = "PCA of the TMT experiment before correction, coloured by condition and by TMT mix. The samples separate by mix along the first component."}
tmt_pca_before <- pca_covariates_plot(
    tmt_res$se_proc,
    assay_name = "Impseqrob_min",
    covariates = c("Condition", "Mix"),
    scale. = TRUE,
    verbose = FALSE)$grid

tmt_pca_before +
    ggplot2::theme(panel.spacing.x = grid::unit(2, "lines"))
```

Before correction, PC1 explains 66.4% of the variance and clearly separates the
two mixes. Within each biological condition, the samples form two groups along
PC1 according to their mix, whereas condition is more clearly represented along
PC2. The main source of separation is therefore technical rather than biological.

```{r tmt-pca-after, eval = run_bc, message = FALSE, fig.width = 8.5, fig.height = 4, fig.cap = "TMT experiment after correction. Mix-related separation is no longer visible and condition becomes the main source of structure.", fig.alt = "PCA of the TMT experiment after correction, coloured by condition and by TMT mix. The mix panel has no structure left and the conditions separate along the first component."}
tmt_pca_after <- pca_covariates_plot(
    tmt_bc$se_proc,
    assay_name = "BERT",
    covariates = c("Condition", "Mix"),
    scale. = TRUE,
    verbose = FALSE)$grid

tmt_pca_after +
    ggplot2::theme(panel.spacing.x = grid::unit(2, "lines"))
```

After correction, samples from the two mixes overlap within the condition-related
clusters. PC1 remains the dominant axis but now separates the biological
conditions rather than the TMT mixes. This visual result agrees with the PVCA
comparison, although neither diagnostic alone evaluates false-positive control.

## What it changed in the results

```{r tmt-de-effect, eval = run_bc}
table(uncorrected = tmt_res$DEPs_results$Change)
table(corrected   = tmt_bc$DEPs_results$Change)
```

Across the three comparisons, the number of significant protein–comparison calls
rises from approximately 1,100 to 3,300. The spike-in evaluation shows why this
increase cannot be interpreted as an improvement on its own. Many additional
calls involve human proteins, even though the experimental design specifies a
constant 80% human background.

At the default `logFC_threshold = 0`, a small but consistent shift can become
statistically significant once the residual variance is reduced. Correction
therefore improves detection of the intended *E. coli* and yeast changes, but it
also increases false-positive calls against the known constant background. Both
sensitivity and specificity must be considered when deciding whether the
corrected results are preferable.

## Statistical significance and fold-change magnitude

Statistical significance indicates whether the data support a difference,
whereas the absolute log2 fold change describes how large that estimated
difference is. Here, the magnitude of the log2 fold change is used as the effect
size. The additional calls observed after correction may therefore be
statistically significant even when their estimated abundance changes are
small. The `D-A` comparison illustrates this distinction by comparing the
significant human and *E. coli* proteins:

```{r tmt-effect-size, eval = run_bc}
fold_change_summary <- function(res, organism) {
    d  <- res$DEPs_results
    d  <- d[d$Comparison == "D-A", ]
    sp <- species[match(d$Protein.IDs, names(species))]
    d  <- d[grepl(organism, sp) & d$Change != "No Change", ]
    c(significant      = nrow(d),
      median_abs_logFC = median(abs(d$logFC)),
      p90_abs_logFC    = quantile(abs(d$logFC), 0.9, names = FALSE))
}

round(rbind(
    `human, uncorrected`   = fold_change_summary(tmt_res, "^Homo"),
    `human, corrected`     = fold_change_summary(tmt_bc,  "^Homo"),
    `E. coli, uncorrected` = fold_change_summary(tmt_res, "^Escherichia"),
    `E. coli, corrected`   = fold_change_summary(tmt_bc,  "^Escherichia")), 2)
```

After correction, the significant human proteins have a median absolute log2
fold change of only 0.09, equivalent to a difference of approximately 6% in
intensity (FC = 1.06). The corresponding median for the significant *E. coli*
proteins is 0.84, equivalent to an increase of approximately 79% in intensity
(FC = 1.79) and therefore indicating a much larger abundance change. The human
background is
expected to remain constant in this spike-in experiment, so these small human
changes should not be interpreted as meaningful biological effects.

Batch correction reduces the residual variability and can therefore make small,
consistent differences statistically significant. With the default
`logFC_threshold = 0`, NADIA does not require a minimum fold-change magnitude.
If the analysis should retain only changes that are both statistically supported
and sufficiently large to be biologically relevant, a non-zero
`logFC_threshold` can be specified.

`benchmarking_proteomics()` compares the results with the known spike-in changes.
The expected values and the complete scoring procedure are described in
`vignette("benchmarking")`; here, the `D-A` results are summarised across several
log2 fold-change thresholds. True positives (TP) are *E. coli* and yeast
proteins recovered with the expected direction of change, whereas false
positives (FP) are human proteins called differential despite the constant human
background. The Matthews correlation coefficient (MCC) summarises the complete
confusion matrix in a single score, with values closer to 1 indicating better
overall classification. Click the table to enlarge it.

::: {.tmt-benchmark-table .lightbox-figure}
```{r tmt-threshold-table, echo = FALSE, out.width = "100%", fig.cap = "Benchmarking of the D-A comparison across log2 fold-change thresholds. Sensitivity, specificity and MCC are coloured on a common scale from 0 (red) to 1 (green); the 0.3 threshold is highlighted because it provides the best overall balance in this experiment.", fig.alt = "Table comparing true positives, false positives, sensitivity, specificity and Matthews correlation coefficient for uncorrected, corrected and condition-protected analyses at log2 fold-change thresholds of zero, 0.2, 0.3, 0.5 and 1. True positives are expected E. coli and yeast changes and false positives are unchanged human proteins called differential. Sensitivity, specificity and MCC cells are coloured from red at zero to green at one. The cells for threshold 0.3 are bold and outlined."}
knitr::include_graphics("figures/TMT_Benchmarking_Thresholds.svg")
```
:::

With no fold-change threshold, correction recovers more true positives but also
introduces many false positives, resulting in a lower MCC. Thresholds between
0.2 and 0.5 provide a better balance: at 0.3, for example, the corrected analysis
achieves the highest MCC (0.9167) while recovering 450 true positives. The
protected and unprotected corrections remain very similar in this balanced
design. Nevertheless, protecting the biological condition is the safer choice
when it can be estimated, particularly when conditions are unevenly distributed
across batches.

TMT data are affected by **ratio compression**: interference from co-isolated
precursors makes measured abundance ratios closer to 1 and therefore moves log2
fold changes towards zero. In the `D-A` comparison, the expected changes of +2
for *E. coli* and -3.3 for yeast are measured as approximately +0.84 and -1.62,
respectively. The true spike-in changes are consequently closer to the unchanged
human background than the experimental proportions would suggest.

This compression makes threshold selection a trade-off. A threshold of 1 removes
almost all human false positives but also misses more than half of the expected
spike-in changes, whereas a threshold of 0 counts even very small changes and
partly reflects statistical power. The threshold should therefore balance
sensitivity and specificity while taking the degree of ratio compression into
account; **no single value is appropriate for every experiment**.

# References

- Schumann Y, Schlumbohm S, Neumann JE, *et al.* (2025). [High performance data
  integration for large-scale analyses of incomplete *Omic* profiles using
  Batch-Effect Reduction Trees (BERT)](https://doi.org/10.1038/s41467-025-62237-4).
  *Nature Communications*, 16, 7104.

- Johnson WE, Li C, Rabinovic A (2007). [Adjusting batch effects in microarray
  expression data using empirical Bayes
  methods](https://doi.org/10.1093/biostatistics/kxj037). *Biostatistics*, 8(1),
  118–127. This work introduced the ComBat batch-correction method.

# Session information

```{r session}
sessionInfo()
```
