---
title: "Batch correction: diagnose, correct, verify"
author: "Sergio Ciordia"
package: NADIA
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Batch correction: diagnose, correct, verify}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", crop = NULL)
```

# Introduction

Samples digested on different days, or injected in different sequences, carry
that history in their intensities. Batch correction removes it. Done carelessly
it removes the biology along with it, and nothing warns you — the result simply
looks cleaner and means less.

This vignette follows the discipline that avoids that: **diagnose before
correcting, verify after**. Along the way it shows two things that are easy to
get wrong: the difference between removing a factor and adjusting for it, and the
fact that a large batch effect can be **invisible** in the default variance
decomposition.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("NADIA")
```

```{r library, message = FALSE}
library(NADIA)
data(nadia_dia)
```

Batch correction needs `BERT` and the variance decomposition needs `lme4`; both
are optional dependencies:

```{r deps}
have_bert <- requireNamespace("BERT", quietly = TRUE)
have_lme4 <- requireNamespace("lme4", quietly = TRUE)
c(BERT = have_bert, lme4 = have_lme4)
```

```{r flags, include = FALSE}
run_bc  <- have_bert
run_all <- have_bert && have_lme4
```

# The sample annotation

The example experiment has no batches — it is twelve injections of one digest. So
we attach a plausible technical history: two digestion days crossed with two
injection sequences, and four subjects.

In a real analysis this is your sample sheet, read from a file. The `Column`
column must match the sample names the pipeline uses:

```{r covariates}
md <- nadia_dia$metadata

covariates <- data.frame(
    Column    = md$Coding,
    Digestion = rep(c("d1", "d2"),             length.out = nrow(md)),
    Injection = rep(c("i1", "i1", "i2", "i2"), length.out = nrow(md)),
    Subject   = rep(paste0("s", 1:4),          length.out = nrow(md)),
    stringsAsFactors = FALSE)

covariates
```

## A simulated batch effect

The annotation alone is just labels; the data have no batch effect to find. To
have something to correct we **inject one deliberately**, so that every number
below can be checked against a known truth.

The effect is protein-dependent on purpose. A batch that simply scaled every
protein by a constant would be removed by normalisation and would never reach
this vignette; a real digestion-day effect hits different proteins differently,
and that is what survives normalisation and needs ComBat.

```{r inject}
batched <- nadia_dia
qcols   <- grep("^PG.Quantity_", colnames(batched$protein_quant))
samples <- sub("^PG.Quantity_", "", colnames(batched$protein_quant)[qcols])
d2_cols <- qcols[samples %in% covariates$Column[covariates$Digestion == "d2"]]

set.seed(1)
up     <- sample(c(TRUE, FALSE), nrow(batched$protein_quant), replace = TRUE)
factor <- ifelse(up, 2.5, 1 / 2.5)   # half the proteins up, half down, in d2
batched$protein_quant[, d2_cols] <- batched$protein_quant[, d2_cols] * factor

length(d2_cols)   # the six samples digested on day 2
```

Everything from here on is a real computation on those data.

# Step 1: is the design correctable at all?

This is the step most often skipped, and it costs one line. Before deciding what
to remove, cross the factors and count:

```{r design}
table(Digestion = covariates$Digestion, Injection = covariates$Injection)
table(Condition = md$R.Condition,       Digestion = covariates$Digestion)
```

Both tables are full: every digestion day contains both injection sequences, and
every condition appears in both digestion days. That is what makes the batch
effect separable from the biology.

Had a condition appeared in only one digestion day, the two would be
**confounded**, and no algorithm could tell them apart. ComBat would still run
and still return a matrix — it would simply have removed the biological
difference as if it were a batch effect. This table is your only protection
against that, and it is worth printing every time.

# Step 2: how much variance does the batch explain?

```{r process, eval = run_bc}
res <- process_proteomics(batched, covariate_df = covariates, verbose = FALSE)
```

`pvca_analysis()` decomposes the variance of the principal components over the
factors you name. Run at its defaults, it reports this:

```{r pvca-default, eval = run_all, message = FALSE, warning = FALSE}
pvca_default <- pvca_analysis(
    res$se_proc,
    assay_name         = "Impseqrob_min",
    technical_factors  = c("Digestion", "Injection"),
    biological_factors = "Condition",
    verbose            = FALSE)

pvca_default$variance_components
```

Condition explains essentially everything and the digestion day does not appear
at all — even though we injected a 2.5-fold effect into half the proteins and
know for a fact that it is there.

## Why the default hides it, and what to do

`pca_threshold` controls how many principal components enter the decomposition:
enough to reach that share of the variance, 60 % by default. In this experiment
the conditions differ enormously — many proteins are absent from one condition
entirely — so PC1 alone clears 60 %, and PC1 is pure condition. The batch effect
lives in later components, and the default never looks at them.

Raise the threshold and it appears:

```{r pvca-before, eval = run_all, message = FALSE, warning = FALSE}
pvca_before <- pvca_analysis(
    res$se_proc,
    assay_name         = "Impseqrob_min",
    technical_factors  = c("Digestion", "Injection"),
    biological_factors = "Condition",
    pca_threshold      = 0.95,
    variance_threshold = 0,
    verbose            = FALSE)

pvca_before$variance_components
```

Digestion accounts for about three quarters of the variance across those
components, and condition for barely a fifth. **The batch effect is not merely
present; it is dominant, and it is masking the biology.**

The lesson generalises beyond this dataset: a decomposition that reports nothing
technical is not evidence that there is nothing technical. Check what
`pca_threshold` let in before believing it.

```{r pvca-before-plot, eval = run_all, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, fig.alt = "Variance decomposition before batch correction"}
pvca_before$plot
```

# Step 3: correct, and keep the biology

`process_proteomics()` runs batch correction as an optional step **before**
imputation. Three of its arguments are easy to confuse, and they do different
things:

| Argument | Effect |
|---|---|
| `batch_column` | the factor to be **removed** from the data |
| `covariate_column` | a variable **kept**, as a term in the limma design |
| `bio_replicate_column` | a **blocking** factor, for samples that are not independent |

`batch_column = "Digestion"` erases the digestion-day effect from the matrix.
`covariate_column = "Injection"` does the opposite — it leaves the injection
effect in the data and accounts for it when testing.
`bio_replicate_column = "Subject"` tells limma that samples from the same subject
are correlated, so it estimates that correlation instead of assuming
independence.

```{r correct, eval = run_bc, message = FALSE}
res_bc <- process_proteomics(
    batched,
    covariate_df         = covariates,
    batch_correct        = TRUE,
    batch_column         = "Digestion",
    batch_algorithm      = "ComBat",
    batch_ComBat_mode    = 1,
    covariate_column     = "Injection",
    bio_replicate_column = "Subject",
    verbose              = FALSE)

SummarizedExperiment::assayNames(res_bc$se_proc)
```

Note where the new `BERT` assay sits: after `cycloess`, before the imputed
matrix. Correction runs on the normalised data, and imputation then runs on its
output.

## Protecting a variable inside ComBat

A fourth argument, `batch_covariates`, is passed to ComBat itself so a variable
is *preserved* during correction rather than absorbed into it. When condition and
batch are partly confounded this is what stops ComBat from erasing the biology,
and `batch_correct_proteomics()` warns when it is left empty.

It is not free. Protecting a variable adds parameters to a model fitted per
batch, and with few samples per combination the model becomes singular and the
correction fails outright. With twelve samples, three conditions and two batches,
`batch_covariates = "Condition"` is already too much to ask of these data. The
honest rule: protect the biological variable when the design can afford it, and
when it cannot, say so and treat the result with suspicion.

# Step 4: verify

## Did the batch effect go?

Because we injected the effect, we can measure its removal directly. The mean
difference between the two digestion days, per protein, should collapse:

```{r verify-direct, eval = run_bc}
grp <- covariates$Digestion[
    match(colnames(SummarizedExperiment::assay(res_bc$se_proc, "cycloess")),
          covariates$Column)]

delta <- function(assay_name) {
    m <- SummarizedExperiment::assay(res_bc$se_proc, assay_name)
    rowMeans(m[, grp == "d1", drop = FALSE], na.rm = TRUE) -
    rowMeans(m[, grp == "d2", drop = FALSE], na.rm = TRUE)
}

data.frame(
    assay      = c("cycloess (before)", "BERT (after)"),
    median_abs = c(abs(median(delta("cycloess"), na.rm = TRUE)),
                   abs(median(delta("BERT"),     na.rm = TRUE))),
    sd         = c(sd(delta("cycloess"), na.rm = TRUE),
                   sd(delta("BERT"),     na.rm = TRUE)))
```

The spread of the between-day difference drops by an order of magnitude.

## Did the biology survive?

That is the other half, and the half that matters. Repeat the decomposition on
the corrected assay, at the same threshold:

```{r pvca-after, eval = run_all, message = FALSE, warning = FALSE}
pvca_after <- pvca_analysis(
    res_bc$se_proc,
    assay_name         = "BERT",
    technical_factors  = c("Digestion", "Injection"),
    biological_factors = "Condition",
    pca_threshold      = 0.95,
    variance_threshold = 0,
    verbose            = FALSE)

merge(pvca_before$variance_components, pvca_after$variance_components,
      by = "label", suffixes = c("_before", "_after"))[, c(1, 2, 4)]
```

Digestion falls from roughly 0.74 to numerically zero, and condition rises from
about 0.22 to 0.79. That is the pattern you want: the technical factor gone,
the biological factor **recovered**, not merely preserved.

The failure case looks different — the technical factor falls and the biological
one falls with it. This table is where you would see it, which is the whole
reason for computing the decomposition twice.

## What it changed in the results

```{r de-effect, eval = run_bc}
table(uncorrected = res$DEPs_results$Change)
table(corrected   = res_bc$DEPs_results$Change)
```

Uncorrected, the batch effect inflates the residual variance and buries real
differences. After correction the number of proteins called significant rises
sharply, because the biology was there all along.

This is also why the direction of the check matters. A correction that *increases*
significance while the biological share of the variance rises is doing its job. A
correction that increases significance while the biological share falls is
manufacturing results, and only the decomposition can tell you which one you have.

# Removing two factors, one after the other

`batch_correct_proteomics()` corrects one factor per call, but it reads an assay
and writes another, so calls chain. To remove digestion and then injection, point
the second call at the output of the first:

```{r cascade, eval = run_bc, message = FALSE, warning = FALSE}
se1 <- batch_correct_proteomics(
    res$se_proc,
    assay_name           = "Impseqrob_min",
    batch_column         = "Digestion",
    corrected_assay_name = "BERT",
    verbose              = FALSE)

se2 <- batch_correct_proteomics(
    se1,
    assay_name           = "BERT",       # the already-corrected matrix
    batch_column         = "Injection",
    corrected_assay_name = "BERT_2",
    verbose              = FALSE)

SummarizedExperiment::assayNames(se2)
```

Two caveats. Each correction consumes degrees of freedom, and the second is
fitted on data the first has already altered. And the order is not neutral when
the two factors are not orthogonal — here they are, which is exactly what the
design table in step 1 established.

# Looking at it directly

PVCA summarises; a PCA lets you look. `pca_covariates_plot()` draws the same
projection once per covariate, which is the quickest way to spot a factor that
separates samples:

```{r pca-cov, eval = run_bc, message = FALSE, fig.width = 7, fig.height = 6, fig.alt = "PCA coloured by condition, digestion, injection and subject"}
pca_grid <- pca_covariates_plot(
    res_bc$se_proc,
    assay_name = "BERT",
    covariates = c("Condition", "Digestion", "Injection", "Subject"),
    scale.     = TRUE,
    verbose    = FALSE)

pca_grid$grid
```

Samples grouping by condition and mixed with respect to the technical factors is
the picture you want. Still grouping by digestion day means the correction did not
take, and the cascade above is the next thing to try.

# The same thing on a real batch

Everything so far rested on an effect we planted ourselves. That was deliberate:
planting it is what let us check that what came out was what went in.

But it leaves the last test hanging. The rule stated above — significance rises
*and* the biological share of the variance rises — is a rule about telling a real
correction from a manufactured one. Applied to an invented effect, the only thing
it can be checked against is the invention.

The TMT example report closes that circle. It carries a batch nobody planted, and
a biology that is known in advance.

## The batch has to be declared, not discovered

`nadia_tmt_report.tsv.gz` was labelled in two TMTpro mixes: replicates 1 to 4 of
every condition in the first, replicates 5 to 8 in the second. Two labelling
reactions, two acquisitions — a batch effect in the plainest sense.

It is also nowhere in the file. A Proteome Discoverer report has no column naming
the mix a channel came from, and `preprocess_tmt()` reads nothing but the report.
So this batch is declared exactly as the digestion days were declared at the start
of this vignette, and by the same mechanism: a data frame whose `Column` values
match the sample names, handed to `process_proteomics()`. Where that information
comes from is your problem, not the pipeline's — here it happens to be recoverable
from the replicate number.

```{r tmt-data}
tmt <- preprocess_tmt(
    system.file("extdata", "nadia_tmt_report.tsv.gz", package = "NADIA"),
    condition_order = c("A", "B", "D"), verbose = FALSE)

tmt_cov <- data.frame(
    Column = tmt$metadata$Coding,
    Mix    = ifelse(tmt$metadata$R.Replicate <= 4, "mix1", "mix2"),
    stringsAsFactors = FALSE)

table(Condition = tmt$metadata$R.Condition, Mix = tmt_cov$Mix)
```

Step 1, again, and it passes: every condition appears in both mixes, four samples
each. The batch is separable from the biology. Hold on to the fact that the table
is *balanced* — it decides what the correction is going to buy.

## The decomposition finds it

```{r tmt-process, eval = run_bc, message = FALSE}
tmt_res <- process_proteomics(tmt, covariate_df = tmt_cov, verbose = FALSE)
```

```{r tmt-pvca, eval = run_all, message = FALSE, warning = FALSE}
pvca_analysis(
    tmt_res$se_proc,
    assay_name         = "Impseqrob_min",
    technical_factors  = "Mix",
    biological_factors = "Condition",
    pca_threshold      = 0.95,
    variance_threshold = 0,
    verbose            = FALSE)$variance_components
```

The mix takes the larger share. Nothing was injected to make that happen: it is
what two labelling reactions do to real samples, and it is the reason this step
is worth running on data you did not simulate.

## Correcting it, and checking against biology

```{r tmt-correct, eval = run_bc, message = FALSE, warning = FALSE, results = "hide"}
tmt_bc <- process_proteomics(
    tmt, covariate_df = tmt_cov,
    batch_correct = TRUE, batch_column = "Mix",
    batch_algorithm = "ComBat", batch_ComBat_mode = 1,
    verbose = FALSE)
```

No `batch_covariates` here. The design table above showed the mix and the
condition are fully crossed, so there is no biological signal riding on the batch
for ComBat to mistake for one — which is the situation that argument exists to
protect against.

Now the part the simulated example could not do. This report is a three-proteome
spike-in: *E. coli* was loaded in increasing amounts from condition A to D, yeast
in decreasing amounts, and human protein was held constant. So for the `D-A`
comparison we know the answer before running anything — every *E. coli* protein
should come out up, every yeast protein down.

The species is in the description line, in UniProt's `OS=` field. `preprocess_lfq()`
parses it for you; the TMT reader does not, so we take it here:

```{r tmt-verify, eval = run_bc}
species <- sub(".*OS=(.+?)\\s+[A-Za-z]+=.*", "\\1",
               tmt$protein_id$PG.ProteinDescriptions)
names(species) <- tmt$protein_id$PG.ProteinGroups

found <- function(res, comparison) {
    d  <- res$DEPs_results
    d  <- d[d$Comparison == comparison, ]
    sp <- species[match(d$Protein.IDs, names(species))]
    c(`E. coli called up` = mean(d$Change[grepl("^Escherichia", sp)] == "Up"),
      `yeast called down` = mean(d$Change[grepl("^Saccharomyces", sp)] == "Down"))
}

round(100 * cbind(uncorrected = found(tmt_res, "D-A"),
                  corrected   = found(tmt_bc,  "D-A")), 1)
```

More of the proteins that should have been found are found. The verdict the
simulated example could only assert, this one can check.

## What the correction actually bought

It is worth being precise about *what* improved, because it is not what most
people expect. Compare the effect sizes themselves — the median log2 fold change
per species, which for a spike-in is a quantity we can also predict:

```{r tmt-fold, eval = run_bc}
fold <- function(res) {
    d  <- res$DEPs_results
    d  <- d[d$Comparison == "D-A", ]
    sp <- species[match(d$Protein.IDs, names(species))]
    vapply(list(`E. coli` = "^Escherichia",
                yeast     = "^Saccharomyces",
                human     = "^Homo"),
           function(p) median(d$logFC[grepl(p, sp)]), numeric(1))
}

round(cbind(uncorrected = fold(tmt_res), corrected = fold(tmt_bc)), 2)
```

**They barely move.** The fold changes were right before the correction and are
right after it. What changed is how many of them cleared the significance
threshold.

That is not a disappointment; it is what a balanced design implies, and it is
worth stating as a rule:

> When every condition is equally represented in every batch, the batch cannot
> bias your fold changes. It inflates the residual variance instead, and buries
> real differences under it. Correcting it does not make the answers more
> accurate — it makes more of the true answers reach significance.

The confounded case is the other one, and it is the dangerous one: there the batch
*does* distort the fold changes, and no amount of statistical power rescues them.
Which of the two you are in is decided by the design table in step 1, printed
before any of this begins — which is why that step is the one not to skip.

```{r session}
sessionInfo()
```
