---
title: "Benchmarking with a spike-in: scoring against known truth"
author: "Sergio Ciordia"
package: NADIA
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Benchmarking with a spike-in: scoring against known truth}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", crop = NULL)
```

# Introduction

`vignette("choosing-methods")` had to work by proxy, because in an ordinary
experiment nobody knows which proteins really changed. A spike-in experiment
removes that limitation. Proteins from a foreign organism are added at known
ratios into a constant background, so for every protein there is a right answer:
the spike-in species *should* come out differential, at a ratio you specified,
and the background *should not*.

With real answers available you can count mistakes rather than infer them, and
that is what this vignette does — including a definition of "mistake" that is
less obvious than it looks.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("NADIA")
```

```{r library, message = FALSE}
library(NADIA)
data(nadia_dia)
```

Some metrics need `pROC`, which is optional:

```{r deps}
have_proc <- requireNamespace("pROC", quietly = TRUE)
have_proc
```

# The experiment

The example dataset is a three-organism spike-in: a human background with
*E. coli* and yeast added at different amounts per condition.

```{r species}
spikein <- read.delim(system.file("extdata", "nadia_dia_spikein.tsv.gz",
                                  package = "NADIA"))
species_df <- data.frame(Protein.IDs = spikein$PG.ProteinGroups,
                         Species     = spikein$PG.OrganismId)

table(species_df$Species)
```

`species_df` is deliberately a separate table rather than a column of the
results. The species is a property of the protein, not of any analysis, so it can
be attached to the output of any pipeline you want to score.

# Declaring the truth

`expected_values` says what each species should do in each comparison, as a
log2 fold change. This is the specification of the experiment, taken from how
much of each organism was mixed in — not something derived from the data.

```{r expected}
expected <- data.frame(
    Comparison     = rep(c("B-A", "D-A", "D-B"), each = 2),
    Species        = c("ECOLI", "YEAST"),
    expected_logFC = c(1, -0.58, 2, -3.3, 1, -2.72))

expected
```

Human proteins do not appear, and that is the convention: any species without a
declared expectation is background, and should not change.

# Scoring a pipeline

```{r bench, message = FALSE, warning = FALSE}
res <- process_proteomics(nadia_dia, verbose = FALSE)

bench <- benchmarking_proteomics(res$DEPs_results, expected,
                                 species_df = species_df,
                                 output_dir = NULL, verbose = FALSE)
names(bench)
```

```{r bench-metrics}
bench$metrics_table
```

# How TP, FP, TN and FN are defined

This is the part worth reading slowly, because the natural definition is wrong in
a way that quietly corrupts the specificity.

Two independent things go into every classification.

**`truth` — what the protein *is*.** It is 1 if the protein belongs to a
spike-in species with a declared expectation, 0 if it is background. This comes
from `species_df` and `expected_values`, that is, from how the experiment was
mixed. It is a property of the protein and **no analysis can change it**.

**`predicted` — what the pipeline *said*.** It is 1 only if the protein is
significant (`adj.P.Val < alpha`) **and** its `logFC` has the sign the experiment
specified. Significance alone is not enough.

Crossing them gives the four cells:

| | `truth = 1` (spike-in) | `truth = 0` (background) |
|---|---|---|
| **`predicted = 1`** | **TP** — detected, right direction | **FP** — background wrongly flagged |
| **`predicted = 0`** | **FN** — missed, *or found with the wrong sign* | **TN** — correctly left alone |

The cell that decides the design is the bottom left. A spike-in protein that
comes out **significant with the wrong sign** is an **FN**, not an FP.

The temptation to call it a false positive is understandable — the pipeline did
make a wrong claim about it. But the protein genuinely is differential; the
experiment says so. What failed was the characterisation, which makes it a missed
detection. Filing it under FP would put a spike-in protein into the count of
*background proteins wrongly flagged*, and that count is the entire content of
the specificity. One misfiled protein and specificity stops measuring what its
name says.

Those cases are not hypothetical. In this run there are nine:

```{r classification}
wrong_sign <- merge(
    subset(bench$classified_df, truth == 1 & classification == "FN" &
                                Change != "No Change"),
    expected, by = c("Comparison", "Species"))
wrong_sign <- wrong_sign[sign(wrong_sign$logFC) != sign(wrong_sign$expected_logFC), ]

nrow(wrong_sign)
head(wrong_sign[, c("Protein.IDs", "Comparison", "Species", "logFC",
                    "expected_logFC", "adj.P.Val", "classification")], 4)
```

Yeast proteins that should have dropped by 3.3 and instead rose, significantly.
Every one is an `FN`, and every one keeps `truth = 1`.

The invariant that follows is worth checking, because it is what makes the
specificity trustworthy:

```{r invariant}
table(classification = bench$classified_df$classification,
      truth          = bench$classified_df$truth)
```

Two of the four cells are empty by construction. `TP` and `FN` occur only where
`truth` is 1; `FP` and `TN` only where it is 0. **No prediction can move a protein
between the biological classes** — which is also why `truth` can be handed to
`pROC` untouched, so the AUC always scores the ranking of a fixed, biologically
defined label.

# The confusion plots: the whole picture at once

The tables above are per comparison. The confusion figures show every cell of
every comparison together, and that global view is where the structure of the
experiment becomes visible at a glance.

Start with the breakdown by species, which is the most informative of the two:

```{r conf-species, fig.width = 7.5, fig.height = 5, fig.alt = "Confusion matrix by species and comparison"}
bench$gg_confusion_by_species
```

```{r conf-species-tbl}
bench$confusion_by_species
```

Read down the columns and the definitions appear as a shape. **Every `HUMAN` row
has zeros in TP and FN**; every spike-in row has zeros in FP and TN. That is not
a property of this pipeline — it is the invariant above, drawn. Background
proteins can only ever be TN or FP, and spike-ins only ever TP or FN.

What is left is the real performance, and it is legible per species. In `B-A`,
*E. coli* is recovered 87.9 % of the time but yeast only 42.5 %, because the yeast
ratio in that comparison is a log2 fold change of -0.58 — a factor of 1.5, which
is genuinely hard. Move to `D-A`, where yeast is expected at -3.3, and recovery
jumps to 90.9 %. The benchmark is not scoring the pipeline in the abstract; it is
scoring it against effects of a stated size, and small effects are supposed to be
harder.

The overall view collapses the species and adds the percentages:

```{r conf-overall, fig.width = 7.5, fig.height = 4.5, fig.alt = "Overall confusion matrix by comparison"}
bench$gg_confusion_overall
```

```{r conf-overall-tbl}
bench$confusion_overall
```

The FP column is the one to watch: 1.4 %, 3.0 % and 1.8 % of proteins. Those are
false discovery rates in the region a 5 % threshold promises, which is the basic
sanity check — a pipeline whose FP column ran into double figures would be
failing at multiple-testing correction regardless of how good its sensitivity
looked.

# The metrics heatmap

Six metrics across the three comparisons, on one colour scale:

```{r bench-heatmap, fig.width = 7.5, fig.height = 4, fig.alt = "Heatmap of the benchmark metrics by comparison"}
bench$gg_heatmap
```

The value of the heatmap is that it makes the *pattern of weakness* visible.
Specificity and precision are uniformly high across all three comparisons; the
column that varies is sensitivity, and it varies with the size of the spiked-in
difference. A pipeline with the opposite profile — steady sensitivity, wobbling
specificity — would be a pipeline with a multiple-testing problem, and the two
look nothing alike here.

# The OpDEA metrics

The usual metrics have a weakness this experiment exposes: 1,256 of the 1,997
proteins are background and should not change, so a method that called *nothing*
significant would score 63 % accuracy. The OpDEA benchmarking study
(Peng et al., 2024) selected three metrics that resist that:

```{r opdea, eval = have_proc}
bench$opdea_metrics
```

- **nMCC** — the Matthews correlation coefficient rescaled to `[0, 1]`. It uses
  all four cells of the confusion matrix, so unlike accuracy it cannot be
  inflated by a large, easy majority class.
- **G_mean** — the geometric mean of sensitivity and specificity. Being
  geometric, it collapses if either one is poor; an arithmetic mean would let a
  good sensitivity hide a bad specificity.
- **pAUC** — partial area under the ROC curve, restricted to low false positive
  rates (1 %, 5 %, 10 %) and McClish-corrected so the numbers stay comparable
  across different restrictions.

The three tell a consistent story here: `B-A` is the hard comparison
(nMCC 0.84, G_mean 0.79) and `D-A` the easy one (0.94 both), which matches what
the confusion plots showed about the effect sizes.

# ROC curves, and why the zoom is the one that matters

The ROC curve traces sensitivity against false positive rate as the q-value
threshold sweeps from 0 to 1. The full curve first:

```{r roc, eval = have_proc, fig.width = 6.5, fig.height = 5, fig.alt = "ROC curves for the three comparisons"}
bench$gg_roc
```

Its legend reports the **full AUC** for each comparison, in the order the
comparisons appear in the data — it is not ranked.

The AUCs are high — 0.83 to 0.98 — and that is the number usually quoted. But
consider what the AUC is an average over: **every possible threshold**, including
q = 0.5 and q = 0.9, which no one would ever use. A large part of the area under
that curve is earned in a region of the parameter space that has nothing to do
with how the analysis is actually run.

The zoom restricts the view to the first 10 % of false positive rate:

```{r roc-zoom, eval = have_proc, fig.width = 6.5, fig.height = 5, fig.alt = "ROC curves zoomed into the low false-positive region"}
bench$gg_roc_zoom
```

Two things change in the legend, and both are deliberate.

It now reports the **pAUC at 5 % FPR** instead of the full AUC — the McClish-
corrected area over the region reached at roughly q = 0.05, not over all
thresholds.

And the entries are **sorted by that pAUC, best first**. This is the difference
from the plot above, where the comparisons appear in whatever order the data
supplied them. So in the zoomed figure the legend is itself the ranking: read it
top to bottom and you have the comparisons ordered by how well they perform at
the threshold you actually use. A comparison that led on full AUC can drop down
this list, and when it does, the full AUC was flattering it.

The reason this is the right view is concrete, and it can be checked. The
operating point of the analysis — the actual sensitivity and false positive rate
you get at the threshold you actually used, `adj.P.Val < 0.05` — is already in
`metrics_table`, since `FPR = 1 - Specificity`:

```{r operating-point}
op <- data.frame(
    Comparison = bench$metrics_table$Comparison,
    FPR_at_q005 = round(1 - bench$metrics_table$Specificity, 4),
    TPR_at_q005 = bench$metrics_table$Sensitivity)
op

all(op$FPR_at_q005 <= 0.10)   # inside the zoom window?
```

Between 2.2 % and 4.8 % false positive rate. **All three operating points sit
inside the zoomed region**, and two of the three inside the 5 % band the pAUC
integrates over. The full ROC devotes most of its width to thresholds to the
right of every point you will ever use; the zoom shows the neighbourhood you are
actually standing in.

So when two pipelines are separated by 0.01 of AUC, that difference may live
entirely at q-values nobody applies. A difference in pAUC at 5 % is a difference
where you work. That is why `pAUC_005` carries weight in the ranking below, and
why the zoom is worth reading before the full curve.

# The volcanoes

Shown but not run, to keep this vignette light; `vignette("visualization")`
renders the interactive figures live.

```{r bench-volcano, eval = FALSE}
# One interactive volcano per comparison, points coloured by species
bench$hc_volcano_list[["B-A"]]
```

These are the only Highcharts objects the module produces; everything above is
ggplot2. Colouring by species turns a volcano into a diagnostic: background
proteins should cluster at the origin, and each spike-in species should sit
around its expected fold change.

# Comparing whole pipelines

Scoring one pipeline tells you how it did. The point of a benchmark is to choose
between several, and `benchmarking_multiple()` does that from the outputs of
individual runs.

```{r multiple, message = FALSE, warning = FALSE, eval = have_proc}
pipelines <- list(
    cycloess_combo = list(norm_method = "cycloess", imp_method = "combo"),
    quantile_combo = list(norm_method = "quantile", imp_method = "combo"),
    cycloess_min   = list(norm_method = "cycloess", imp_method = "min"))

invisible(capture.output(
  opdea <- do.call(rbind, lapply(names(pipelines), function(p) {
      r <- do.call(process_proteomics,
                   c(list(nadia_dia, verbose = FALSE), pipelines[[p]]))
      b <- benchmarking_proteomics(r$DEPs_results, expected,
                                   species_df = species_df,
                                   output_dir = NULL, verbose = FALSE)
      cbind(Assay = p, b$opdea_metrics)
  }))))

ranking <- bm_compute_ranking(opdea)
names(ranking)
ranking$mean_ranking
```

Rank 1 is the best pipeline, averaged over the five OpDEA metrics and the three
comparisons. Here the verdict is unanimous — `cycloess` with `combo` wins on
every metric.

That is the same question `vignette("choosing-methods")` had to answer by proxy,
and the two answers agree: `cycloess` came first there too, on replicate
tightness and condition separation alone, with no knowledge of which proteins
had actually been spiked in. The agreement is worth noting in both directions. It
is evidence that the proxy metrics are measuring something real, and it is a
reason to trust them on the experiments where no spike-in is available — which is
most of them.

Agreement is not guaranteed, though, and when the two disagree the spike-in wins:
it is scoring against what actually happened, and the proxy is scoring against an
assumption about what a good analysis looks like. That is why a spike-in is worth
including in a method-development experiment even though it costs a channel.

```{r multiple-plot, eval = have_proc, fig.width = 7, fig.height = 4.5, fig.alt = "Ranking of the pipelines across the OpDEA metrics"}
bm_plot_ranking_bars(ranking, type = "mean")
```

# A note on what this proves

A spike-in tells you which pipeline best recovers *a known ratio of a foreign
organism spiked into a fixed background*. That is a real question and a hard one,
but it is not identical to your biological question. Spike-in proteins are
present in every sample at controlled amounts, whereas the biology you care about
may be genuinely absent from a condition — which is the MNAR case that
`vignette("missing-values")` is about, and the one a spike-in reproduces least
well.

Treat the ranking as strong evidence about the analytical pipeline, not as proof
that its output is right for a different kind of experiment.

```{r session}
sessionInfo()
```
