---
title: "Choosing a normalisation and an imputation method"
author: "Sergio Ciordia"
package: NADIA
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Choosing a normalisation and an imputation method}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", crop = NULL)
```

# Introduction

Thirteen normalisation methods and nineteen imputation methods make 247
pipelines, all of which run without complaint and none of which announces itself
as the right one. The choice is usually made by convention, and convention is a
poor guide here: the methods differ in their assumptions, and your data either
satisfy those assumptions or they do not.

This vignette measures the alternatives instead. It uses **no ground truth** —
that is the hard case, and the ordinary one. If your experiment includes a
spike-in, you have real answers to score against and
`vignette("benchmarking")` is the better tool.

Without ground truth the question has to be rephrased into something answerable:

- **Normalisation** — which method leaves replicates most alike and conditions
  most distinct? That is measurable directly.
- **Imputation** — hide values you *do* have, impute them, and see who
  reconstructs them best. That is measurable by simulation.

Both are proxies. The final section is about where they stop working.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("NADIA")
```

```{r library, message = FALSE}
library(NADIA)
data(nadia_dia)
```

# Normalisation

`nm_prepare_se()` builds the object and `nm_run_normalizations()` fills it with
one assay per method. A handful is enough to show the shape of the answer; drop
the `methods` argument for all twelve benchmark methods.

```{r nm-run, message = FALSE}
se_norm <- nm_prepare_se(nadia_dia, verbose = FALSE)
se_norm <- nm_run_normalizations(
    se_norm,
    methods = c("log2Norm", "cycloess", "Rlr", "quantile"),
    verbose = FALSE)

SummarizedExperiment::assayNames(se_norm)
```

`normalization_metrics()` then scores them. Setting `output_dir = NULL` keeps it
from writing anything to disk:

```{r nm-metrics, message = FALSE, warning = FALSE}
nm <- normalization_metrics(se_norm, output_dir = NULL, verbose = FALSE)
nm$final_rank
```

## Reading that table

Five criteria, combined by averaging their ranks. Rank 1 is best throughout.

The first four ask whether replicates agree, computed **per protein** and then
summarised:

- `Rank_PCV` — pooled coefficient of variation within condition.
- `Rank_PMAD` — pooled median absolute deviation, the robust version.
- `Rank_PEV` — pooled estimate of variance.
- `Rank_Cor` — intragroup correlation, which should be high.

Alone they would be misleading, and dangerously so: a method that flattened every
sample onto the same values would take first place in all four. So the fifth
criterion pulls the other way.

- `Rank_Sep` — separation between conditions, as an F-ratio on PC1.

A method wins by tightening replicates *and* keeping conditions apart. Here
`cycloess` does both and takes the top spot, with `Rlr` close behind. Plain
`log2` is last, which is the expected result: it is the baseline, no
normalisation at all.

The individual rankings are available if the summary hides something:

```{r nm-detail}
nm$pcv_rank
nm$pc1_rank
```

Every diagnostic plot is in the same object — `nm$boxplot`, `nm$density`,
`nm$pca`, `nm$correlation`, `nm$mds`, `nm$qq` and the ranking charts:

```{r nm-plot, fig.width = 7, fig.height = 4.5, fig.alt = "Final ranking of the normalisation methods"}
nm$final_ranking
```

# Imputation

Here a simulation supplies the missing ground truth. `imputation_metrics()`
takes the proteins with no missing values at all, blanks a fraction of them
(`na_prop`, 0.2), re-imputes, and compares against what it hid.

```{r im-metrics, message = FALSE, warning = FALSE}
se_imp <- im_prepare_se(nadia_dia, norm_method = "cycloess", verbose = FALSE)

im <- imputation_metrics(
    se_imp,
    assay_name = "cycloess",
    methods    = c("Impseqrob", "knn", "MinDet", "QRILC", "min"),
    combo_methods = list(
        Impseqrob_min = list(mar_method = "Impseqrob", mnar_method = "min")),
    output_dir = NULL,
    verbose    = FALSE)

im$metrics_table
```

Four metrics, from the NAguideR framework:

- **NRMSE** — normalised root mean squared error. Lower is better.
- **SOR** — sum of ranks, which penalises features left unimputed.
- **PSS** — Procrustes sum of squares, on the overall structure of the data.
- **ACC_OI** — correlation between imputed and true values, computed **only on
  the masked cells**. Computed over whole rows it would sit near 1 for every
  method, since most of each row was never touched, and would distinguish
  nothing.

`ACC_OI` is `NA` for `min`, and that is informative rather than broken: `min`
writes the same constant into every gap in a protein, and the correlation between
a constant and anything is undefined. A method that cannot vary its answer cannot
be scored on how well it varies.

## An important caveat

Read that table again with its construction in mind. The simulation can only hide
values it can see, so the blanks it creates are **missing at random by
definition** — and the ranking that follows measures how well each method handles
random gaps.

That is precisely the case MNAR methods are not built for. `min` comes last by a
wide margin here, but its job is to fill values that dropped below the detection
limit, and this benchmark contains none of those. The same holds with
`pattern = "from_data"`, which copies the observed per-row missingness rates but
still masks observed values.

So the table answers a real question — *if a value is missing at random, who
reconstructs it best?* — and it answers it honestly. It just is not the question
of whether `min` is a good MNAR method, and nothing computed on complete rows can
be. That question needs `vignette("benchmarking")`, or a defensible assumption.

Note where the two-stage `combo` lands: mid-table. It is worse than pure
`Impseqrob` at reconstructing random gaps, because half its values come from the
MNAR method by design. On a benchmark composed entirely of MAR gaps, that is
exactly what should happen.

# Chaining the two decisions

The imputation benchmark runs on a normalised matrix, so the normalisation
choice comes first and feeds the second — and `im_prepare_se()` accepts the
ranking directly, rather than making you copy the winner across by hand:

```{r chain, message = FALSE, warning = FALSE}
se_chained <- im_prepare_se(nadia_dia, pc1_rank = nm$pc1_rank, verbose = FALSE)
SummarizedExperiment::assayNames(se_chained)
```

The assay named after the winner is the one the imputation benchmark then scores,
so the two decisions stay consistent.

# Putting the choice to work

Whatever the tables recommend, it reaches the analysis as two arguments:

```{r apply, message = FALSE}
res <- process_proteomics(nadia_dia,
                          norm_method = "cycloess",
                          imp_method  = "combo",
                          mar_method  = "Impseqrob",
                          mnar_method = "min",
                          verbose     = FALSE)

table(res$DEPs_results$Change)
```

# How much does it matter?

Enough to be worth the ten seconds:

```{r sensitivity, message = FALSE, warning = FALSE}
variants <- list(
    "cycloess + combo" = list(norm_method = "cycloess", imp_method = "combo"),
    "cycloess + knn"   = list(norm_method = "cycloess", imp_method = "knn"),
    "quantile + combo" = list(norm_method = "quantile", imp_method = "combo"),
    "log2 + min"       = list(norm_method = "log2",     imp_method = "min"))

# Some imputation methods print progress on stdout; capture.output keeps it out.
invisible(capture.output(
    counts <- do.call(rbind, lapply(names(variants), function(v) {
        r <- do.call(process_proteomics,
                     c(list(nadia_dia, verbose = FALSE), variants[[v]]))
        tab <- table(factor(r$DEPs_results$Change,
                            levels = c("Up", "Down", "No Change")))
        data.frame(pipeline = v, Up = tab[["Up"]], Down = tab[["Down"]],
                   No_Change = tab[["No Change"]])
    }))))

counts
```

Same data, same test, same threshold, and the answer changes shape entirely. The
count of down-regulated proteins runs from 197 to 2,825 across these four
pipelines — a factor of fourteen — and the total called differentially abundant
from about 1,100 to 3,700.

That spread is the argument for measuring the choice rather than inheriting it.
It is also the argument for stating in any publication which normalisation and
imputation produced the table, since without that the numbers are not
reproducible even from the same raw data.

```{r session}
sessionInfo()
```
