---
title: "Input formats: DIA, TMT and label-free"
author: "Sergio Ciordia"
package: NADIA
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Input formats: DIA, TMT and label-free}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", crop = NULL)
```

# Introduction

NADIA reads three kinds of report, and after the reading step it stops caring
which one it was. `preprocess_spectronaut()`, `preprocess_tmt()` and
`preprocess_lfq()` all return an object of class `proteomics_data`, with the same
three elements, so `process_proteomics()` and everything downstream take any of
them unchanged.

This vignette shows the three preprocessors on real reports and then makes the
convergence explicit. The interesting part is not that they converge — it is
*where they get the experimental design from*, because that differs in each case
and is the one thing you have to get right.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("NADIA")
```

```{r library, message = FALSE}
library(NADIA)
```

# DIA: Spectronaut and DIA-NN

Spectronaut and DIA-NN export a **long** report: one row per protein group *per
run*, with the condition and replicate written in every row.

```{r dia}
dia_file <- system.file("extdata", "nadia_dia_report.tsv.gz", package = "NADIA")
dia <- preprocess_spectronaut(dia_file, condition_order = c("A", "B", "D"),
                              verbose = FALSE)
dia
```

The design comes from the report itself, in the `R.Condition` and `R.Replicate`
columns. `condition_order` does not define the design; it fixes the order the
conditions appear in downstream and selects which ones to keep.

```{r dia-md}
dia$metadata
```

Because the report is run-level and long, several columns hold one value per run
for the same protein — sequence coverage, molecular weight, C-score. The
`agg_*` arguments choose how those collapse to a single number per protein:

```{r dia-agg, eval = FALSE}
dia_median <- preprocess_spectronaut(
    dia_file, condition_order = c("A", "B", "D"),
    agg_coverage_run = "median",   # default "max"
    agg_cscore_runwise = "median", # default "mean"
    verbose = FALSE)
```

These affect only the annotation carried alongside the data, never the
intensities.

# TMT: Proteome Discoverer

A TMT report from Proteome Discoverer is **wide**: one row per protein, one
column per channel. The design is encoded in the column names, as
`Abundance: <condition>_<replicate>`.

```{r tmt}
tmt_file <- system.file("extdata", "nadia_tmt_report.tsv.gz", package = "NADIA")
tmt_cols <- colnames(read.delim(tmt_file, nrows = 1, check.names = FALSE))
grep("^Abundance", tmt_cols, value = TRUE)
```

`preprocess_tmt()` parses those suffixes to build the sample-to-condition map.
Nothing else is consulted, which is worth knowing: **if the channel names are
wrong, the design is wrong**, and there is no annotation file to correct it.

This experiment has four conditions of eight replicates plus four `IS` channels,
an internal standard used to bridge the two TMT mixes. `condition_order` is where
you leave it out, because a standard is not a biological group and including it
would generate contrasts against it:

```{r tmt-run}
tmt <- preprocess_tmt(tmt_file, condition_order = c("A", "B", "C", "D"),
                      verbose = FALSE)
tmt
head(tmt$metadata)
table(tmt$metadata$Condition)
```

# Label-free: Proteome Discoverer

The label-free export is wide too, but differs from TMT in two ways.

First, it carries **per-sample metrics** beyond the abundance — a Mascot score, a
PSM count and a peptide count for every sample:

```{r lfq-cols}
lfq_file <- system.file("extdata", "nadia_lfq_report.tsv.gz", package = "NADIA")
lfq_cols <- colnames(read.delim(lfq_file, nrows = 1, check.names = FALSE))
per_sample <- c("Abundance", "Score Mascot", "# PSMs (by Search Engine)",
                "# Peptides (by Search Engine)")
vapply(per_sample, function(p) sum(startsWith(lfq_cols, paste0(p, ":"))), integer(1))
```

Those are matched to samples **by sample name**, not by column position.

Second, and more important, the design does **not** live in the column names.
It comes from a separate annotation file:

```{r lfq-annot}
annot_file <- system.file("extdata", "nadia_lfq_annotation.tsv", package = "NADIA")
read.delim(annot_file)
```

`Column` names the sample as it appears in the report, `Condition` assigns it to
a group, and the optional `Experiment` column separates independent experiments.
This is the more flexible arrangement: relabelling an experiment means editing a
small text file rather than renaming columns in the report.

```{r lfq-run}
lfq <- preprocess_lfq(lfq_file, annot_path = annot_file, verbose = FALSE)
lfq
lfq$metadata
```

# Where the design comes from

That is the whole difference, and it is worth stating in one place:

| Format | Layout | Design taken from |
|---|---|---|
| Spectronaut / DIA-NN | long, run-level | the `R.Condition` / `R.Replicate` columns |
| TMT (PD) | wide | the `Abundance: <cond>_<rep>` column suffixes |
| LFQ (PD) | wide | a **separate annotation file** (`annot_path`) |

# Naming the conditions

Whatever the format, the condition names end up as **column names of a design
matrix**, and limma builds contrasts from them by pasting names around a minus
sign. `"B-A"` is not a label the package invented for display; it is the
expression `B - A` evaluated against that matrix.

That single fact decides which names are usable. A condition called `Ct-1` would
produce the contrast string `Tr-1-Ct-1`, which cannot be parsed back into two
condition names, and which R cannot interpret as an expression either.

The rule is: **a condition name must be a syntactically valid R name.** In
practice:

| In the name | Verdict | Why |
|---|---|---|
| letters, digits (not first) | fine | `Ctrl`, `Treat24h` |
| `.` full stop | **fine** | `Ct.1` gives the contrast `Tr.1-Ct.1` |
| `_` underscore | **fine** | `Ct_a` gives `Tr_a-Ct_a` |
| `-` hyphen | **rejected** | collides with the contrast separator |
| space | **rejected** | not a syntactic name |
| leading digit | **rejected** | `1Ct` is not a name; R would rewrite it `X1Ct` |

So yes, a full stop is safe, and it is the natural choice if you want to separate
parts of a condition name — `Ct.1`, `Ct.2` rather than `Ct-1`, `Ct-2`.
`make.names()` is the exact test:

```{r names-test}
candidates <- c("Ctrl", "Ct.1", "Ct_a", "Ct-1", "Ct 1", "1Ct")
data.frame(name  = candidates,
           valid = make.names(candidates) == candidates,
           what_r_would_do = make.names(candidates))
```

An invalid name does not slip through to produce a wrong answer. The pipeline
checks before fitting the model and stops with a message that names the problem:

```{r names-error, error = TRUE}
bad <- dia
bad$metadata$R.Condition <- paste0(bad$metadata$R.Condition, "-x")
bad$metadata$Coding <- paste0(bad$metadata$R.Condition, "_",
                              bad$metadata$R.Replicate)
colnames(bad$protein_quant) <- sub("^(PG.Quantity_)([A-Z])_", "\\1\\2-x_",
                                   colnames(bad$protein_quant))

process_proteomics(bad, verbose = FALSE)
```

One caution while you are renaming by hand, because this failure mode *does* pass
unnoticed. Samples are matched to their intensities by **`metadata$Coding`
against the `PG.Quantity_<coding>` column names**. Those two must agree. Break
the pairing and the affected samples are silently dropped — no error, just a
smaller analysis:

```{r names-drop}
half <- dia
half$metadata$R.Condition <- sub("^A$", "A.1", half$metadata$R.Condition)
half$metadata$Coding <- paste0(half$metadata$R.Condition, "_",
                               half$metadata$R.Replicate)
# ... but the PG.Quantity_ columns still say "A_1", "A_2", ...

res_drop <- process_proteomics(half, verbose = FALSE)
ncol(res_drop$se_proc)   # 12 samples went in
unique(as.character(SummarizedExperiment::colData(res_drop$se_proc)$Condition))
```

Four samples gone and condition A with them, from a rename that looked complete.
Note that `R.Condition` on its own is only a label: changing it alone is
harmless, because it is not what the matching uses. It is `Coding` and the column
names that have to move together. If a run comes back with fewer samples than you
sent it, this is the first thing to check.

Two format-specific notes. In **TMT** the condition is read from the
`Abundance: <condition>_<replicate>` suffix; the replicate is matched as the
trailing digits, so an underscore inside the condition name is still parsed
correctly (`Ct_a_1` gives condition `Ct_a`, replicate `1`). In **LFQ** the names
come from the `Condition` column of the annotation file, which is the easiest
place to fix them — no report needs to be touched.

# One contract, three sources

All three carry the same class hierarchy, and share `proteomics_data`:

```{r classes}
vapply(list(DIA = dia, TMT = tmt, LFQ = lfq), function(x) class(x)[1], character(1))
vapply(list(DIA = dia, TMT = tmt, LFQ = lfq),
       function(x) inherits(x, "proteomics_data"), logical(1))
```

And the same three elements:

```{r elements}
lapply(list(DIA = dia, TMT = tmt, LFQ = lfq), names)
```

Which is what lets the pipeline be indifferent to the source:

```{r pipeline}
res_tmt <- process_proteomics(tmt, verbose = FALSE)
res_lfq <- process_proteomics(lfq, verbose = FALSE)

res_tmt$comparisons
res_lfq$comparisons
table(res_lfq$DEPs_results$Change)
```

Not a single argument had to change. The same is true of every vignette that
follows: where they say `nadia_dia`, a TMT or LFQ object would do.

# One caveat about the metadata

The three metadata tables are *not* identical, even though the three objects are
interchangeable. Spectronaut reports identification counts that Proteome
Discoverer does not:

```{r md-cols}
lapply(list(DIA = dia, TMT = tmt, LFQ = lfq), function(x) colnames(x$metadata))
```

Functions that display those counts degrade rather than fail —
`summary_list_widget()` shows whichever columns are present. See
`vignette("results-and-export")`.

```{r session}
sessionInfo()
```
