---
title: "Input formats: DIA, TMT and label-free"
author: "Sergio Ciordia"
package: NADIA
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Input formats: DIA, TMT and label-free}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", crop = NULL)
```

# Introduction

NADIA reads four kinds of report, and after the reading step it stops caring
which one it was. `preprocess_spectronaut()`, `preprocess_diann()`,
`preprocess_tmt()` and `preprocess_lfq()` all return an object of class
`proteomics_data`, with the same three elements, so `process_proteomics()` and
everything downstream take any of them unchanged.

This vignette shows the four preprocessors on real reports and then makes the
convergence explicit. The interesting part is not that they converge — it is
*where they get the experimental design from*, because that differs in each case
and is the one thing you have to get right.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("NADIA")
```

```{r library, message = FALSE}
library(NADIA)
```

# DIA: Spectronaut

Spectronaut exports a **long** report: one row per protein group *per run*, with
the condition and replicate written in every row.

```{r dia}
dia_file <- system.file("extdata", "nadia_dia_report.tsv.gz", package = "NADIA")
dia <- preprocess_spectronaut(dia_file, condition_order = c("A", "B", "D"),
                              verbose = FALSE)
dia
```

The design comes from the report itself, in the `R.Condition` and `R.Replicate`
columns. `condition_order` does not define the design; it fixes the order the
conditions appear in downstream and selects which ones to keep.

```{r dia-md}
dia$metadata
```

Because the report is run-level and long, several columns hold one value per run
for the same protein — sequence coverage, molecular weight, C-score. The
`agg_*` arguments choose how those collapse to a single number per protein:

```{r dia-agg, eval = FALSE}
dia_median <- preprocess_spectronaut(
    dia_file, condition_order = c("A", "B", "D"),
    agg_coverage_run = "median",   # default "max"
    agg_cscore_runwise = "median", # default "mean"
    verbose = FALSE)
```

These affect only the annotation carried alongside the data, never the
intensities.

# DIA: DIA-NN

DIA-NN is the other DIA search engine in common use, and it needs its own
function for a reason that has nothing to do with the acquisition: its
protein-group matrix (`report.pg_matrix.tsv`) is **wide**. One row per protein
group, one column per run, like the Proteome Discoverer exports below and unlike
the Spectronaut report above.

```{r diann}
diann_file <- system.file("extdata", "nadia_diann_report.tsv.gz",
                          package = "NADIA")
diann <- preprocess_diann(diann_file, condition_order = c("A", "B", "D"),
                          verbose = FALSE)
diann
```

## Where the design comes from, and the one thing you must supply

DIA-NN names each intensity column after the raw file it came from:

```
D:\DIA-NN\Projects\20241014_XII_CursoProtQ_DIA_500ng_sample_A1.raw
```

That path says nothing the software can rely on. `sample_A1` reads as condition
A, replicate 1 to a person, and would read the same way to a regular expression
— right here, and wrong on the next report, silently. So the design is never
guessed. You declare it, in one of two ways, and neither needs a separate
annotation file.

**Rename the columns** to the `Abundance: <Condition>_<Replicate>` convention
that Proteome Discoverer already uses. The report shipped with the package has
been renamed this way, which is why the call above needs no extra argument.

**Or name the runs in order**, and leave the report exactly as DIA-NN wrote it:

```{r diann-sample-names, eval = FALSE}
preprocess_diann(
    "report.pg_matrix.tsv",
    condition_order = c("A", "B", "D"),
    sample_names = c("A_1", "A_2", "A_3", "A_4",
                     "B_1", "B_2", "B_3", "B_4",
                     "D_1", "D_2", "D_3", "D_4"))
```

`sample_names` is positional: one label per intensity column, in the order the
columns appear in the file. If you supply neither, the error lists those columns
in that order, so the vector is a copy-paste away.

Either way, `R.FileName` keeps the **original** header, so renaming never costs
you the trail back to the raw file:

```{r diann-md}
diann$metadata
```

## What a protein-group matrix does not carry

This is the sparsest of the four formats. A `pg_matrix` holds intensities and
six annotation columns, and nothing at all per sample — no precursor counts, no
sequence coverage, no scores, no molecular weight. `protein_id` is
correspondingly thin:

```{r diann-cols}
c(spectronaut = ncol(dia$protein_id), diann = ncol(diann$protein_id))
```

Two DIA-NN columns do map onto the wide contract: `N.Sequences` becomes the
identified-peptide family and `N.Proteotypic.Sequences` the one used for
quantification, replicated across runs the same way `preprocess_tmt()` replicates
the Proteome Discoverer globals. Coverage and the C-score are simply left out
rather than filled with `NA`, because nothing downstream reads them and an empty
column is worse than an absent one — it draws a filter box that empties the
table. The precursor counts *are* kept as `NA`, and only because the protein and
quantification tables detect their sample columns by looking for them.

The one per-run number a matrix can still give is how many groups were
quantified in each run, and it is computed rather than read — over the 2,000
groups of this trimmed file, so it is not comparable with the counts in the
Spectronaut metadata above, which the search engine reported over the whole
experiment:

```{r diann-counts}
diann$metadata[, c("Coding", "R.ProteinGroupsIdentified")]
```

## The same runs, searched twice

The DIA-NN report here and the Spectronaut report above are the **same twelve
injections** of the same experiment, put through the two search engines. That
makes the missingness directly comparable, and it is not the same:

```{r diann-vs-spectronaut}
miss <- function(x) {
    q <- as.matrix(x$protein_quant[grep("^PG.Quantity_",
                                        names(x$protein_quant))])
    q[q == 0] <- NA
    c(missing = round(100 * mean(is.na(q)), 1),
      complete = round(100 * mean(rowSums(is.na(q)) == 0), 1))
}
rbind(spectronaut = miss(dia), diann = miss(diann))
```

Which engine is "right" is not the point — the point is that the choice of
search engine moves the missingness by half again, before any imputation method
has been named. That is the decision `vignette("missing-values")` picks up.

A caveat on these two files: each is an independent random sample of 2,000
protein groups from its own full search, so they overlap in only a few hundred
groups. They are faithful for comparing *rates*, not for comparing protein by
protein — that needs the full reports.

# TMT: Proteome Discoverer

A TMT report from Proteome Discoverer is **wide**: one row per protein, one
column per channel. The design is encoded in the column names, as
`Abundance: <condition>_<replicate>`.

```{r tmt}
tmt_file <- system.file("extdata", "nadia_tmt_report.tsv.gz", package = "NADIA")
tmt_cols <- colnames(read.delim(tmt_file, nrows = 1, check.names = FALSE))
grep("^Abundance", tmt_cols, value = TRUE)
```

`preprocess_tmt()` parses those suffixes to build the sample-to-condition map.
Nothing else is consulted, which is worth knowing: **if the channel names are
wrong, the design is wrong**, and there is no annotation file to correct it.

This experiment has three conditions of eight replicates plus four `IS` channels,
an internal standard used to bridge the two TMT mixes: replicates 1 to 4 are the
first mix and 5 to 8 the second. `condition_order` is where you leave the standard
out, because it is not a biological group and including it would generate
contrasts against it:

```{r tmt-run}
tmt <- preprocess_tmt(tmt_file, condition_order = c("A", "B", "D"),
                      verbose = FALSE)
tmt
head(tmt$metadata)
table(tmt$metadata$R.Condition)
```

# Label-free: Proteome Discoverer

The label-free export is wide too, but differs from TMT in two ways.

First, it carries **per-sample metrics** beyond the abundance — a Mascot score, a
PSM count and a peptide count for every sample:

```{r lfq-cols}
lfq_file <- system.file("extdata", "nadia_lfq_report.tsv.gz", package = "NADIA")
lfq_cols <- colnames(read.delim(lfq_file, nrows = 1, check.names = FALSE))
per_sample <- c("Abundance", "Score Mascot", "# PSMs (by Search Engine)",
                "# Peptides (by Search Engine)")
vapply(per_sample, function(p) sum(startsWith(lfq_cols, paste0(p, ":"))), integer(1))
```

Those are matched to samples **by sample name**, not by column position.

Second, and more important, the design does **not** live in the column names.
It comes from a separate annotation file:

```{r lfq-annot}
annot_file <- system.file("extdata", "nadia_lfq_annotation.tsv", package = "NADIA")
read.delim(annot_file)
```

`Column` names the sample as it appears in the report, `Condition` assigns it to
a group, and the optional `Experiment` column separates independent experiments.
This is the more flexible arrangement: relabelling an experiment means editing a
small text file rather than renaming columns in the report.

```{r lfq-run}
lfq <- preprocess_lfq(lfq_file, annot_path = annot_file, verbose = FALSE)
lfq
lfq$metadata
```

# Where the design comes from

That is the whole difference, and it is worth stating in one place:

| Format | Layout | Design taken from |
|---|---|---|
| Spectronaut | long, run-level | the `R.Condition` / `R.Replicate` columns |
| DIA-NN | wide | the `Abundance: <cond>_<rep>` suffixes, or `sample_names` |
| TMT (PD) | wide | the `Abundance: <cond>_<rep>` column suffixes |
| LFQ (PD) | wide | a **separate annotation file** (`annot_path`) |

Only the long format carries its own design. The three wide ones have to be told,
and the difference between them is just how much editing that costs you.

# Naming the conditions

Whatever the format, the condition names end up as **column names of a design
matrix**, and limma builds contrasts from them by pasting names around a minus
sign. `"B-A"` is not a label the package invented for display; it is the
expression `B - A` evaluated against that matrix.

That single fact decides which names are usable. A condition called `Ct-1` would
produce the contrast string `Tr-1-Ct-1`, which cannot be parsed back into two
condition names, and which R cannot interpret as an expression either.

The rule is: **a condition name must be a syntactically valid R name.** In
practice:

| In the name | Verdict | Why |
|---|---|---|
| letters, digits (not first) | fine | `Ctrl`, `Treat24h` |
| `.` full stop | **fine** | `Ct.1` gives the contrast `Tr.1-Ct.1` |
| `_` underscore | **fine** | `Ct_a` gives `Tr_a-Ct_a` |
| `-` hyphen | **rejected** | collides with the contrast separator |
| space | **rejected** | not a syntactic name |
| leading digit | **rejected** | `1Ct` is not a name; R would rewrite it `X1Ct` |

So yes, a full stop is safe, and it is the natural choice if you want to separate
parts of a condition name — `Ct.1`, `Ct.2` rather than `Ct-1`, `Ct-2`.
`make.names()` is the exact test:

```{r names-test}
candidates <- c("Ctrl", "Ct.1", "Ct_a", "Ct-1", "Ct 1", "1Ct")
data.frame(name  = candidates,
           valid = make.names(candidates) == candidates,
           what_r_would_do = make.names(candidates))
```

An invalid name does not slip through to produce a wrong answer. The pipeline
checks before fitting the model and stops with a message that names the problem:

```{r names-error, error = TRUE}
bad <- dia
bad$metadata$R.Condition <- paste0(bad$metadata$R.Condition, "-x")
bad$metadata$Coding <- paste0(bad$metadata$R.Condition, "_",
                              bad$metadata$R.Replicate)
colnames(bad$protein_quant) <- sub("^(PG.Quantity_)([A-Z])_", "\\1\\2-x_",
                                   colnames(bad$protein_quant))

process_proteomics(bad, verbose = FALSE)
```

One caution while you are renaming by hand, because this failure mode *does* pass
unnoticed. Samples are matched to their intensities by **`metadata$Coding`
against the `PG.Quantity_<coding>` column names**. Those two must agree. Break
the pairing and the affected samples are silently dropped — no error, just a
smaller analysis:

```{r names-drop}
half <- dia
half$metadata$R.Condition <- sub("^A$", "A.1", half$metadata$R.Condition)
half$metadata$Coding <- paste0(half$metadata$R.Condition, "_",
                               half$metadata$R.Replicate)
# ... but the PG.Quantity_ columns still say "A_1", "A_2", ...

res_drop <- process_proteomics(half, verbose = FALSE)
ncol(res_drop$se_proc)   # 12 samples went in
unique(as.character(SummarizedExperiment::colData(res_drop$se_proc)$Condition))
```

Four samples gone and condition A with them, from a rename that looked complete.
Note that `R.Condition` on its own is only a label: changing it alone is
harmless, because it is not what the matching uses. It is `Coding` and the column
names that have to move together. If a run comes back with fewer samples than you
sent it, this is the first thing to check.

Two format-specific notes. In **TMT and DIA-NN** the condition is read from the
`Abundance: <condition>_<replicate>` suffix — the same parser serves both; the
replicate is matched as the trailing digits, so an underscore inside the
condition name is still parsed correctly (`Ct_a_1` gives condition `Ct_a`,
replicate `1`), and the same holds for a DIA-NN `sample_names` vector. In **LFQ**
the names come from the `Condition` column of the annotation file, which is the
easiest place to fix them — no report needs to be touched.

# One contract, four sources

All four carry the same class hierarchy, and share `proteomics_data`:

```{r classes}
all4 <- list(DIA = dia, DIANN = diann, TMT = tmt, LFQ = lfq)

vapply(all4, function(x) class(x)[1], character(1))
vapply(all4, function(x) inherits(x, "proteomics_data"), logical(1))
```

And the same three elements:

```{r elements}
lapply(all4, names)
```

Which is what lets the pipeline be indifferent to the source:

```{r pipeline}
res_diann <- process_proteomics(diann, verbose = FALSE)
res_tmt   <- process_proteomics(tmt, verbose = FALSE)
res_lfq   <- process_proteomics(lfq, verbose = FALSE)

res_diann$comparisons
res_tmt$comparisons
res_lfq$comparisons
table(res_lfq$DEPs_results$Change)
```

Not a single argument had to change. The same is true of every vignette that
follows: where they say `nadia_dia`, a DIA-NN, TMT or LFQ object would do.

# One caveat about the metadata

The four metadata tables are *not* identical, even though the four objects are
interchangeable. Spectronaut reports three identification counts that Proteome
Discoverer does not report at all and that a DIA-NN matrix can only supply one
of:

```{r md-cols}
lapply(all4, function(x) colnames(x$metadata))
```

Functions that display those counts degrade rather than fail —
`summary_list_widget()` shows whichever columns are present. See
`vignette("results-and-export")`.

```{r session}
sessionInfo()
```
