---
title: "Missing values: what they mean and how to fill them"
author: "Sergio Ciordia"
package: NADIA
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Missing values: what they mean and how to fill them}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", crop = NULL)
```

# Introduction

A blank in a proteomics matrix can mean two very different things, and the whole
of this vignette follows from the difference.

Sometimes a protein is present and the instrument simply failed to measure it —
a bad chromatographic moment, an interfered precursor, a missed match. The blank
carries no information about the protein's abundance, and something close to a
neighbouring value would be a fair guess. This is **missing at random** (MAR).

Sometimes a protein is absent, or below the detection limit. The blank is
*caused* by the abundance being low, so it is highly informative, and filling it
with a neighbouring value destroys exactly the signal you were looking for. This
is **missing not at random** (MNAR), and in a DIA experiment it is the common
case.

No single imputation method handles both well, which is why NADIA's default is a
two-stage hybrid rather than one of the twenty individual methods.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("NADIA")
```

```{r library, message = FALSE}
library(NADIA)
data(nadia_dia)
```

# Looking at the missingness

The example dataset was deliberately built to keep the missingness of the full
experiment — about 8 % of intensities, roughly a quarter of proteins affected.
That was not free: an earlier version selected the 2,000 best-quantified proteins
and ended up with **no missing values at all**, a dataset on which every
imputation method returns its input unchanged.

```{r pattern}
quant <- as.matrix(nadia_dia$protein_quant[
    grep("^PG.Quantity_", colnames(nadia_dia$protein_quant))])
quant[quant == 0] <- NA

round(100 * mean(is.na(quant)), 1)
table(missing_per_protein = rowSums(is.na(quant)))
```

The shape of that table is the diagnostic, and this one is unusually legible.
Counts fall away from zero as you would expect — 70 proteins with one gap, 67
with two, 99 with three — and then jump to **243 proteins missing exactly four
values**. Four is the group size. That spike is not a coincidence and it is not
random: it is proteins that vanish from one entire condition, which is the
signature of missing *not* at random.

The relationship with intensity settles it:

```{r intensity-vs-missing, fig.width = 6, fig.height = 4, fig.alt = "Mean log2 intensity against number of missing values"}
n_missing  <- rowSums(is.na(quant))
mean_log2  <- rowMeans(log2(quant), na.rm = TRUE)

boxplot(mean_log2 ~ n_missing,
        xlab = "missing values in the protein",
        ylab = "mean log2 intensity of the values present",
        main = "Missingness is concentrated at low abundance")
```

The downward trend is the whole argument. If missingness were random, this plot
would be flat: proteins with gaps would be no dimmer than proteins without. It is
not flat, so the gaps are not random, and an imputation method that assumes they
are will be wrong in a predictable direction.

# The methods on offer

Twenty in total. They divide by the assumption they make:

```{r methods}
mar  <- get(".IMP_METHODS_MAR",  envir = asNamespace("NADIA"))
mnar <- get(".IMP_METHODS_MNAR", envir = asNamespace("NADIA"))

setdiff(mar,  "none")   # assume the value is unrelated to the blank
setdiff(mnar, "none")   # assume the blank means "low"
```

MAR methods reconstruct a value from the structure of the data — nearest
neighbours (`knn`), principal components (`bpca`), sequential regression
(`Impseq`, `Impseqrob`), a random forest (`missForest`), chained equations
(`mice`). They are the right tool for a scattered gap.

MNAR methods draw from the low end instead: the lowest intensity observed anywhere
in the experiment (`min`), half that intensity (`halfmin`, the DIA-NN recipe), the
first percentile of each sample (`MinDet`), a truncated distribution (`QRILC`,
`MinProb`), a down-shifted normal (`PI`, the Perseus recipe). They are the right
tool for a protein that dropped out.

`halfmin` deserves a note, because it is where the log2 scale trips people up.
Every method here operates on log2 values, so halving an intensity means
subtracting 1, not dividing by 2 — `log2(m / 2)` is `log2(m) - 1`. Dividing the
log2 value would land roughly eight times lower than intended, at the square root
of the minimum intensity, and how much lower would depend on where zero happens to
fall on the log scale. Note also that `min` and `halfmin` both rest on a single
number out of every value in the matrix, so one aberrantly low measurement moves
them; `MinDet`, `QRILC` and `MinProb` do not have that fragility.

Applied to the wrong kind of blank, each fails in its own direction. A MAR method
on a dropout invents a mid-range value and erases a real difference; an MNAR
method on a random gap invents a low value and manufactures one.

# No single method serves both

That claim is worth testing rather than asserting, and the test is short. Take
the proteins with no gaps at all, so the true values are known, and hide some of
them in two different ways:

- **MAR-like** — three values per protein, chosen at random.
- **MNAR-like** — the three *lowest* values of each protein, which is what a
  dropout looks like.

Then impute, and compare against the values that were hidden. The measure that
matters is not the error but the **bias** — whether a method's guesses are
systematically too high or too low:

```{r mar-vs-mnar, message = FALSE}
se_c  <- im_prepare_se(nadia_dia, norm_method = "cycloess", verbose = FALSE)
full  <- SummarizedExperiment::assay(se_c, "cycloess")
full  <- full[complete.cases(full), , drop = FALSE]

set.seed(42)
rows <- sample(nrow(full), round(0.25 * nrow(full)))

hide <- function(kind) {
    idx <- integer(0)
    for (i in rows) {
        j <- if (kind == "MNAR") order(full[i, ])[1:3] else sample(ncol(full), 3)
        idx <- c(idx, (j - 1) * nrow(full) + i)
    }
    idx
}

refill <- function(mat, method) {
    s <- SummarizedExperiment::SummarizedExperiment(
        assays  = list(cycloess = mat),
        colData = SummarizedExperiment::colData(se_c)[colnames(mat), , drop = FALSE])
    r <- impute_proteomics(s, "cycloess", imp_method = method, verbose = FALSE)
    SummarizedExperiment::assay(r$se, method)
}

invisible(capture.output(
  scores <- do.call(rbind, lapply(c("MAR", "MNAR"), function(kind) {
    idx <- hide(kind)
    holed <- full; holed[idx] <- NA
    do.call(rbind, lapply(c("Impseqrob", "min"), function(m) {
        est <- refill(holed, m)
        data.frame(hidden = kind, method = m,
                   bias = mean(est[idx] - full[idx]))
    }))
  }))))

reshape(scores, idvar = "method", timevar = "hidden", direction = "wide")
```

Read the two columns.

`Impseqrob`, a MAR method, is **essentially unbiased on random gaps** — its guesses
are off by about 0.002 on a log2 scale, which is nothing. Move the gaps to the
bottom of each protein and the same method comes out **systematically too high**.
It reconstructs a value from the protein's other measurements, and when those
measurements are the ones that did *not* drop out, the answer is always too
optimistic. That is how an on/off signal gets erased.

`min` has the opposite problem, and a far larger one. It is off by about seven
log2 units on random gaps — a factor of a hundred — because it answers "very low"
to a question whose answer was "about average". It improves on the MNAR gaps, but
it is never a safe default for a value that is merely unmeasured.

So neither method is wrong; each is wrong *in the wrong place*. And since a real
experiment contains both kinds of blank at once — thinly spread gaps and whole
groups dropping out, as the table above showed — **no single imputation method can
be the right choice for a whole matrix.**

One honest caveat on this demonstration. Even the MNAR-like masking here hides
values that were actually measured, so a MAR method can still interpolate them
tolerably and its bias comes out modest. A genuine dropout is a value that was
never observable at all, and no simulation built from complete data can reproduce
that. The bias in the table is therefore a *lower bound* on the real thing.

That limitation is structural, and it is the reason the default is what it is.

# The default: `combo`, and why it is the default

If no single method can serve a whole matrix, the way out is not to pick the
least-bad one. It is to **stop treating the matrix as one thing**: decide, blank
by blank, which kind of missingness you are looking at, and send each to the
method built for it. That is `combo`, and it is why it rather than any individual
method is the default.

The classification is per protein and per group, governed by three arguments:

- `prop_na_mnar` (0.51) — a group missing more than this share of its values has
  probably dropped out, so its blanks go to the **MNAR** method.
- `prop_present_mar` (0.5) and `min_present_mar` (1) — how much evidence a group
  must retain for reconstruction to be worth attempting, sending its blanks to
  the **MAR** method.

A group that keeps most of its measurements has scattered gaps; a group that has
lost most of them has dropped out. The threshold is a judgement, not a
measurement, and 0.51 simply says "more than half".

```{r combo}
res <- process_proteomics(nadia_dia, verbose = FALSE)
SummarizedExperiment::assayNames(res$se_proc)
```

You can see the split it made — the two stages are reported separately:

```{r combo-split, message = FALSE}
imp <- impute_proteomics(res$se_proc, "cycloess", verbose = FALSE)
c(MAR_values  = sum(imp$mar_mask,  na.rm = TRUE),
  MNAR_values = sum(imp$mnar_mask, na.rm = TRUE))
```

Both kinds are present, and dropouts outnumber scattered gaps by roughly four to
one — which is what a DIA experiment should look like, and exactly the situation
in which handing the whole matrix to a single MAR method would do the most damage.
A dataset with only one kind of blank would not need `combo`; real data are not
like that.

The assay name records the decision rather than hiding it — `Impseqrob` for the
MAR values, `min` for the MNAR ones — so a result carries its own provenance.
Change either and the name follows:

`impute_proteomics()` returns a list, of which `$se` is the
`SummarizedExperiment` with the new assay added:

```{r combo-variants}
imp_qrilc <- impute_proteomics(res$se_proc,
                               normalized_assay_name = "cycloess",
                               mar_method = "knn", mnar_method = "QRILC",
                               verbose = FALSE)
setdiff(SummarizedExperiment::assayNames(imp_qrilc$se),
        SummarizedExperiment::assayNames(res$se_proc))
```

There is a subtlety in `combo` worth stating, because getting it wrong biases
every result downstream. The MNAR parameters — the minimum, the quantiles of the
low tail — are estimated on the **original** matrix, not on the one the MAR stage
has already filled. Estimating them after the MAR stage would let reconstructed
mid-range values raise the estimate of where the low end is, and every MNAR value
would come out too high.

# `softHybrid`: a weight instead of a rule

`combo` makes a hard decision per group. `softHybrid` refuses to: it computes
both imputations and blends them, weighted by how MNAR-like each value looks.

Two sigmoids set the weight. One reads the missingness rate of the group, the
other the intensity of the values present, since a low-intensity protein with
gaps is more likely to be dropping out than a bright one with the same gaps. The
intensity sigmoid is z-scored, so its steepness `b` means the same thing whatever
the dynamic range of your data.

```{r softhybrid}
imp_soft <- impute_proteomics(
    res$se_proc,
    normalized_assay_name = "cycloess",
    imp_method  = "softHybrid",
    mar_method  = "Impseqrob",
    mnar_method = "min",
    method_args = list(softHybrid = list(a = 8, b = 2, lambda = 0.5)),
    verbose     = FALSE)

grep("softHybrid", SummarizedExperiment::assayNames(imp_soft$se), value = TRUE)
```

Whether the softer boundary is worth it is an empirical question, and
`vignette("choosing-methods")` is how to answer it on your own data.

# `limpa`: not imputing at all

`limpa` belongs in a different category. Rather than filling a value and then
pretending it was measured, it models the detection process and carries the
resulting uncertainty forward as **precision weights**: a protein reconstructed
from little evidence contributes less to the test than one measured cleanly.

That only works if the weights survive to the testing step, so they travel with
the data:

```{r limpa, eval = requireNamespace("limpa", quietly = TRUE)}
res_limpa <- process_proteomics(nadia_dia,
                                imp_method = "limpa",
                                de_method  = "limpa",
                                verbose    = FALSE)

class(S4Vectors::metadata(res_limpa$se_proc)$limpa_elist)
table(res_limpa$DEPs_results$Change)
```

Pairing `imp_method = "limpa"` with `de_method = "limpa"` is the point. Using the
limpa imputation with an ordinary limma test throws the weights away, which
forfeits the only reason to use it.

# Comparing two methods on the same data

Method choice is not cosmetic. Two methods on the same normalised matrix:

```{r compare, message = FALSE}
imp_min <- impute_proteomics(res$se_proc, "cycloess", imp_method = "min",
                             verbose = FALSE)
imp_knn <- impute_proteomics(res$se_proc, "cycloess", imp_method = "knn",
                             verbose = FALSE)

m_min <- SummarizedExperiment::assay(imp_min$se, "min")
m_knn <- SummarizedExperiment::assay(imp_knn$se, "knn")
was_na <- is.na(SummarizedExperiment::assay(res$se_proc, "cycloess"))

data.frame(
    method       = c("min (MNAR)", "knn (MAR)"),
    mean_imputed = c(mean(m_min[was_na]), mean(m_knn[was_na])),
    mean_observed = mean(m_min[!was_na]))
```

The MNAR method fills from the bottom of the distribution and the MAR method
fills near the middle — several log2 units apart, on the same blanks. Everything
downstream inherits that gap.

# A guard you should know about

Some methods leave values behind — a protein too sparse for `knn` to find
neighbours, a column a model would not converge on. Rather than let `NA`s
propagate into limma, where they would silently reduce the residual degrees of
freedom, the pipeline fills any residual with the column minimum:

```{r guard}
anyNA(SummarizedExperiment::assay(res$se_proc, "Impseqrob_min"))
```

The exception is `imp_method = "none"`, which is meant to leave the matrix alone.
The guard is a safety net, not a strategy: a method that trips it frequently is
telling you it is not suited to your data, and `imputation_metrics()` will say so
more precisely.

There is also a pre-filter, `max_na_prop`, which drops proteins emptier than a
given fraction before imputation begins. It is **disabled by default**:

```{r prefilter}
res_all <- process_proteomics(nadia_dia, verbose = FALSE)                 # NULL
res_cut <- process_proteomics(nadia_dia, max_na_prop = 0.8, verbose = FALSE)

c(no_filter = nrow(res_all$se_proc), filtered = nrow(res_cut$se_proc))
```

Disabled is the right default for a package about missing values: the proteins it
would remove are the on/off cases from the start of this vignette, present in one
condition and absent from another, which are often the most interesting rows in
the table. Set it when you would rather not report a fold change built on two
surviving measurements out of twelve.

```{r session}
sessionInfo()
```
