---
title: "An introduction to NADIA"
author: "Sergio Ciordia"
package: NADIA
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{An introduction to NADIA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", crop = NULL)
```

# Introduction

NADIA analyses protein quantification data from bottom-up proteomics, from the
raw search-engine report to the differentially abundant proteins and the figures
that describe them.

The name says what the package is about. **DIA** is data-independent
acquisition, and **NA** is the missing values that dominate it. A DIA experiment
routinely leaves 5-20 % of its quantification matrix empty, and those blanks are
not an accident of the instrument: a protein below the detection limit in one
condition and above it in another produces exactly the pattern that a
differential expression test is supposed to find. Deleting those proteins throws
away the signal; filling them in carelessly manufactures it.

That is why the package treats normalisation and imputation as decisions to be
measured rather than defaults to be accepted. It offers 13 normalisation methods
and 19 imputation methods, and — more to the point — functions that score them
against each other on your own data.

This vignette walks the whole pipeline once, on a DIA experiment. The other eight
vignettes each take one part of it further:

| Vignette | Question it answers |
|---|---|
| `vignette("input-formats")` | My data are TMT or label-free, not DIA |
| `vignette("missing-values")` | Which imputation method, and why |
| `vignette("choosing-methods")` | How do I pick, with no ground truth |
| `vignette("benchmarking")` | I have a spike-in, so I *do* have ground truth |
| `vignette("batch-correction")` | My samples were run in batches |
| `vignette("pattern-profiler")` | Which proteins behave alike across conditions |
| `vignette("visualization")` | I want to adjust the figures |
| `vignette("results-and-export")` | How do I get results out of R |

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("NADIA")
```

```{r library, message = FALSE}
library(NADIA)
```

# The data

The package ships a trimmed Spectronaut experiment: three conditions of four
replicates, 2,000 protein groups. It is available both as a report file, so that
parsing can be shown end to end, and as an already-parsed object.

```{r data}
report <- system.file("extdata", "nadia_dia_report.tsv.gz", package = "NADIA")
dia <- preprocess_spectronaut(report, condition_order = c("A", "B", "D"),
                              verbose = FALSE)
dia
```

`preprocess_spectronaut()` turns the long, run-level report into three tables:
the sample `metadata`, the protein `protein_id` annotation, and the
`protein_quant` matrix. The same object is available without re-reading the file:

```{r data-obj}
data(nadia_dia)
identical(dim(nadia_dia$protein_quant), dim(dia$protein_quant))
```

The missing values are the reason the package exists, so it is worth looking at
them before anything else — and it is worth being precise about where they come
from, because each tool encodes them differently.

A Spectronaut report is **long**: one row per run and protein group. Where a
protein was not quantified in a run, there is no value to write — **the row is
simply absent**. In the full experiment behind this example, 11,679 of the
16 × 10,437 possible combinations have no row, which is 7 % of the grid.
`preprocess_spectronaut()` pivots that long table into the wide `protein_quant`
matrix, and it is the pivot that turns those absences into `NA`:

```{r missing}
quant <- as.matrix(nadia_dia$protein_quant[
    grep("^PG.Quantity_", colnames(nadia_dia$protein_quant))])

round(100 * mean(is.na(quant)), 1)          # percentage of missing intensities
round(100 * mean(rowSums(is.na(quant)) == 0), 1)  # complete proteins
```

Roughly 8 % of the intensities are absent, and about a quarter of the proteins
have at least one gap. Those are the proteins the rest of this vignette has to
decide what to do with.

Other tools encode the same thing differently, which is why the pipeline does not
rely on any single convention. DIA-NN and Proteome Discoverer leave the cell
**empty**, which `readr` parses as `NA` on the way in. Some exports — including
Spectronaut's *pivot* report, which is a different format from the long one read
here — write `NaN`, and R treats that as missing too. A few write a literal
**zero**, which is the dangerous one: a zero is a number, it survives into `log2`
as `-Inf`, and it would be silently treated as a real measurement. That is why
`normalize_proteomics()` converts zeros to `NA` before doing anything else, and
reports how many it converted when `verbose = TRUE`. On this dataset it converts
none, because there are none to convert.

# The pipeline in one call

`process_proteomics()` runs the four steps in order — normalisation, optional
batch correction, imputation, differential expression — and returns everything
downstream needs.

```{r process}
res <- process_proteomics(nadia_dia, verbose = FALSE)
res
```

With no arguments beyond the data it uses cyclic loess normalisation, the
two-stage `combo` imputation and limma. Those defaults are defensible, not
optimal; `vignette("choosing-methods")` shows how to check them against your own
data.

The result carries the `SummarizedExperiment` with one assay per stage:

```{r assays}
SummarizedExperiment::assayNames(res$se_proc)
```

`raw` is the input, `log2` its logarithm, `cycloess` the normalised matrix and
`Impseqrob_min` the imputed one. The last name is not a label but a record: the
`combo` method filled values missing at random with `Impseqrob` and those missing
not at random with `min`, and says so.

# The results

```{r deps}
head(res$DEPs_results, 4)
```

One row per protein and comparison. All three pairwise contrasts were built
automatically:

```{r contrasts}
res$comparisons
table(res$DEPs_results$Change, res$DEPs_results$Comparison)
```

`Change` classifies each protein as `Up`, `Down` or `No Change` using `alpha` and
`logFC_threshold` together. It is a convenience, not a separate test: the
underlying evidence is in `adj.P.Val` and `logFC`.

The three `Missing*` columns are what makes a result interpretable once
imputation is in play. They record the percentage of missing values *before* it,
globally and in each of the two groups being compared. Cross them with `Change`
and the structure of the data becomes visible:

```{r missing-cols}
bins <- cut(res$DEPs_results$MissingGlobal, c(-1, 0, 25, 50, 100),
            labels = c("0%", "1-25%", "26-50%", ">50%"))
table(missing = bins, res$DEPs_results$Change)
```

Read the rows. Proteins with no missing values at all are mostly unchanged, and
when they do move they move up. Proteins missing between a quarter and a half of
their values are almost all called *down* — 619 against 70. That asymmetry is not
a bug in the test; it is the signature of values missing **not** at random. A
protein that drops below the detection limit in one condition leaves blanks
there, and blanks filled from the low end of the distribution become a downward
fold change.

At the extreme the pattern is unambiguous:

```{r onoff}
onoff <- subset(res$DEPs_results,
                (MissingPCT1 == 100 & MissingPCT2 == 0) |
                (MissingPCT1 == 0   & MissingPCT2 == 100))
nrow(onoff)
head(onoff[, c("Protein.IDs", "Comparison", "logFC", "adj.P.Val",
               "MissingPCT1", "MissingPCT2")], 4)
```

Five hundred rows where a protein is entirely absent from one group and entirely
present in the other. Those are on/off signals, and a fold change of -11 is a
statement about the detection limit rather than a measured ratio. The columns let
you separate them from fold changes built on measured values, which no p-value
can do on its own — and they are the reason
`vignette("missing-values")` is worth reading before trusting a method choice.

# Figures

## Interactive

Three of the visualisation modules produce Highcharts objects, one per
comparison:

```{r hc, eval = FALSE}
volcano <- volcano_highchart_list(res$DEPs_results)
boxplot <- boxplot_highchart_list(res$BoxPlot_Input)
pca     <- pca_highchart_list(res$PCA_Input, res$DEPs_results)

volcano[["B-A"]]
```

They are omitted from this vignette's output only to keep it small.
`vignette("visualization")` covers their options, and
`vignette("results-and-export")` shows how to save them.

## Static

The heatmap takes `PCA_Input`, the long table the pipeline already built, and
selects the proteins itself. `mode = "target"` keeps those significant in one
named comparison.

```{r heatmap, fig.width = 6.5, fig.height = 5.5, fig.alt = "Heatmap of the proteins significant in the B-A comparison"}
hm <- proteomics_heatmap(
    res$PCA_Input,
    mode            = "target",
    comparison      = "B-A",
    scale_data      = "row",
    sample_order    = "condition",
    condition_order = c("A", "B", "D"),
    show_row_names  = FALSE,
    heatmap_title   = "Significant in B-A")
hm
```

# What to read next

The single decision with the largest effect on this table is which normalisation
and imputation methods produced it. `vignette("choosing-methods")` measures the
alternatives on this same dataset; `vignette("missing-values")` explains what the
imputation methods actually assume.

```{r session}
sessionInfo()
```
