---
title: "An introduction to NADIA"
author: "Sergio Ciordia"
package: NADIA
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{An introduction to NADIA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", crop = NULL)
```

# Introduction

NADIA provides a complete workflow for the analysis of protein quantification
data from bottom-up proteomics. It starts from quantification reports exported by
Spectronaut, DIA-NN or Proteome Discoverer and produces processed matrices,
differential abundance results, tables and visualisations.

The name is drawn from the package title, *Missing Value-Aware Differential
Abundance Analysis of DIA Proteomics Data*, and it reads on two levels: **NA** is
the way R represents missing values, and **DIA** is data-independent acquisition,
the proteomics context in which the package was first developed. Although NADIA
also processes TMT and label-free (DDA) data, it pays particular attention to the
missing values that routinely appear in DIA experiments.

These values do not always represent random measurement failures. Some may relate
to abundances close to or below the limit of detection, and therefore carry
information relevant to the differential abundance analysis. Removing every
affected protein may discard biological signal, while unsuitable imputation may
attenuate real differences or introduce artificial ones.

For this reason, NADIA treats normalisation and imputation as decisions to be
assessed, rather than as default processing steps. The package includes 13
normalisation methods and 19 imputation strategies, together with functions to
compare them and evaluate their behaviour on the data themselves or against a
known reference.

This vignette walks once through the complete workflow using a DIA experiment:
import, inspection of the missing values, normalisation, imputation, differential
abundance analysis and preparation of the results for visualisation. The
remaining vignettes go further into each module and can be read independently.

| Vignette | Question it answers |
|---|---|
| `vignette("input-formats")` | How do I import DIA, TMT or label-free data? |
| `vignette("missing-values")` | What do the missing values mean and how should I impute them? |
| `vignette("choosing-methods")` | How do I compare the methods when I have no known reference? |
| `vignette("benchmarking")` | How do I evaluate the pipelines with a *spike-in* experiment? |
| `vignette("batch-correction")` | How do I diagnose, correct and verify a batch effect? |
| `vignette("pattern-profiler")` | How do I identify proteins with similar profiles across conditions? |
| `vignette("visualization")` | How do I customise the interactive and static figures? |
| `vignette("results-and-export")` | How do I export, archive and share the results? |

# Installation

Once NADIA is available on Bioconductor, it can be installed with BiocManager:

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("NADIA")
```

After installation, the package is loaded into the R session:

```{r library, message = FALSE}
library(NADIA)
```

# Example data

NADIA ships a trimmed DIA dataset derived from a Spectronaut quantification
report. The experiment used in this vignette contains three conditions — A, B and
D — four replicates per condition and 2,000 protein groups.

The dataset is available in two forms. The first is a compressed quantification
report, which allows import and preprocessing to be shown from the start:

```{r data}
report <- system.file("extdata", "nadia_dia_report.tsv.gz", package = "NADIA")
dia <- preprocess_spectronaut(report, condition_order = c("A", "B", "D"),
                              verbose = FALSE)
dia
```

`preprocess_spectronaut()` turns the long-format report — one row per reported
combination of run and protein group — into a `proteomics_data` object. That
object holds three main elements, which split the report along its two natural
axes, identification and quantification:

- `metadata`, one row per run: file name, condition, replicate, and the
  identification counts the report gives for the run as a whole.
- `protein_id` (2,000 × 52 here), the **identification** side. It lists the
  protein groups the search reported after FDR filtering, with their description,
  gene names and molecular weight, and then, for each run separately, how many
  precursors and how many peptides were identified and assigned to each group,
  along with the sequence coverage and the run-wise C-score.
- `protein_quant` (2,000 × 44), the **quantification** side. Alongside the same
  protein annotation it carries the intensity of each protein group in each run,
  in the `PG.Quantity_*` columns, and next to it how many precursors and peptides
  were actually used to compute that intensity — which is not the same as how
  many were identified.

`protein_quant` is the table everything downstream works on. Normalisation,
imputation and the differential abundance analysis all start from the intensity
matrix it holds, and the peptide counts beside it are carried through as
`UniqPepts`, so the results table can say on how few peptides a protein rests.
`protein_id` is not read by the pipeline at all: it feeds
`protein_list_reactable()` and is what lets a result be traced back to the
identification evidence behind it.

In this example, `condition_order = c("A", "B", "D")` selects the conditions to
keep and sets the order used in the later steps.

The same dataset is also distributed as an already preprocessed object, which
allows the rest of the workflow to be run without reading the report again:

```{r data-obj}
data(nadia_dia)
identical(dim(nadia_dia$protein_quant), dim(dia$protein_quant))
```

`nadia_dia` is the object used throughout the rest of this vignette.

## How missing values are represented

Before processing the data it is worth checking how the missing values are
represented, because that encoding depends on the export format used.

The Spectronaut report used here is in long format: each row corresponds to a
reported combination of run and protein group. Where no quantification exists for
a given combination, that combination has no row at all. When the report is
turned into the wide `protein_quant` matrix, `preprocess_spectronaut()` completes
the grid of protein groups and samples, and represents as `NA` the combinations
that were not present in the report:

```{r missing}
quant <- as.matrix(nadia_dia$protein_quant[
    grep("^PG.Quantity_", colnames(nadia_dia$protein_quant))])

round(100 * mean(is.na(quant)), 1)          # percentage of missing intensities
round(100 * mean(rowSums(is.na(quant)) == 0), 1)  # complete proteins
```

In the full experiment from which this trimmed set was taken, 11,679 of the
16 × 10,437 possible sample and protein group combinations were absent, roughly
7 % of the matrix. In the `nadia_dia` object, around 8 % of the intensities are
missing and about a quarter of the protein groups contain at least one.

These are the observations for which the later stages must decide whether to keep
the missing value, impute it, or exclude the protein group.

Other export formats may encode absence in different ways. An empty cell is
converted directly to `NA` on import, while `NaN` is also recognised as missing
by functions such as `is.na()`. Some reports use a zero as the marker of absence.
In an intensity matrix that zero requires explicit handling, because `log2(0)`
yields `-Inf` rather than a missing value.

For this reason, `normalize_proteomics()` converts zeros to `NA` before applying
the logarithmic transformation, and reports how many it converted when
`verbose = TRUE`. On `nadia_dia` no conversion takes place, because the example
matrix contains no zeros.

At this stage no assumption is yet made about whether each missing value is MAR
or MNAR. `vignette("missing-values")` explains the assumptions used to
distinguish the two and how they are mapped onto the imputation methods.

# The workflow in a single call

`process_proteomics()` coordinates the main stages of the analysis:
normalisation, optional batch-effect correction, imputation of the missing values
and differential abundance analysis. The function returns in memory the objects
and tables needed to inspect, visualise and export the results.

```{r process}
res <- process_proteomics(nadia_dia, verbose = FALSE)
res
```

In this call only the dataset is supplied, so NADIA uses its default
configuration:

- normalisation with `cycloess`;
- the hybrid `combo` imputation strategy;
- differential abundance analysis with `limma`.

No batch correction is applied, because no information on technical factors was
supplied and that stage was not explicitly requested.

With `imp_method = "combo"`, the missing values are assigned to two imputation
branches according to the rules of the hybrid strategy. Since neither
`mar_method` nor `mnar_method` is specified, their defaults are used:
`Impseqrob` for the values classified as MAR and `min` for those classified as
MNAR.

These defaults provide a reasonable starting point, but they are not a
universally optimal choice. `vignette("choosing-methods")` shows how to compare
the alternatives on the data themselves, while `vignette("missing-values")`
explains the assumptions behind the different imputation strategies.

## Where the processed data live

From normalisation onwards, NADIA keeps everything in a
`SummarizedExperiment`, the container most Bioconductor packages use for omics
data. The quantification report goes in as a plain table and comes out as one of
these objects, which means the rest of Bioconductor can be pointed at a NADIA
result without conversion.

The object holds three things at once, indexed together:

- one or more **assays**, matrices of identical shape with protein groups in rows
  and samples in columns;
- **`rowData`**, one row of annotation per protein group;
- **`colData`**, one row of metadata per sample — the conditions and replicates
  the design is built from.

Keeping them in a single object is not a matter of tidiness. Subsetting the
object subsets the annotation with it, so a protein can never end up paired with
another protein's identifier or a sample with another sample's condition. The
alignment is structural rather than something every function has to maintain.

NADIA uses the assay slot to record the pipeline itself: one assay per stage,
each the same matrix at a different point in its processing. The processed object
is in `res$se_proc`:

```{r assays}
SummarizedExperiment::assayNames(res$se_proc)
```

The assays produced in this analysis are:

- `raw`, the input quantification matrix;
- `log2`, the matrix after the logarithmic transformation;
- `cycloess`, the normalised matrix;
- `Impseqrob_min`, the matrix after the hybrid imputation.

The name `Impseqrob_min` encodes the two methods used by `combo`: `Impseqrob` for
the MAR branch first, then `min` for the MNAR branch. This naming helps identify
where the assay came from, although fully reproducing the analysis also requires
keeping the arguments used and the package version.

# Differential abundance results

`res$DEPs_results` holds the main table of the differential abundance analysis:

```{r deps}
head(res$DEPs_results, 4)
```

Each row corresponds to a combination of protein group and comparison. The same
protein group may therefore appear in several rows if it takes part in more than
one comparison.

The pairwise comparisons are built automatically from the conditions and the
order defined during preprocessing:

```{r contrasts}
res$comparisons
table(res$DEPs_results$Change, res$DEPs_results$Comparison)
```

The `Change` column classifies each result as `Up`, `Down` or `No Change`, using
the thresholds set through `alpha` and `logFC_threshold` together:

- `Up`, statistically significant with a positive logFC above the threshold
- `Down`, statistically significant with a negative logFC below the threshold
- `No Change`, not meeting both criteria together

`Change` is an interpretive label, not an additional statistical test. The
underlying evidence remains in `logFC`, `P.Value` and `adj.P.Val`, and those
values should be consulted whenever the magnitude and the uncertainty of a result
need to be interpreted.

## Relationship with the missing values

When imputation has been applied, the statistical result must be interpreted
together with the information about the values that were missing before that
stage.

NADIA keeps four columns, which go from the widest scope to the narrowest:

- `MissGlobal`, the percentage of missing values of the protein group across
  every sample in the experiment, including the conditions that take no part in
  the comparison. It therefore does not depend on the comparison: a given
  protein group carries the same value in all of them.
- `MissComp`, the percentage across the replicates of the two conditions being
  compared, and only those. This is the figure that describes the evidence
  actually available for that contrast.
- `MissCND1` and `MissCND2`, the percentages in each of the two conditions
  separately. In a `B-A` comparison, `MissCND1` corresponds to the numerator
  (B) and `MissCND2` to the denominator (A).

The distinction between the first two matters as soon as an experiment has more
than two conditions, as this one does. `MissGlobal` is a property of the protein
group in the experiment, so it does not move from one comparison to the next;
`MissComp` does, because the replicates it counts change with the contrast. A
protein group that is missing in a single replicate of D shows both behaviours at
once:

```{r missing-scope}
pid <- with(res$DEPs_results, Protein.IDs[MissGlobal != MissComp][1])
subset(res$DEPs_results, Protein.IDs == pid,
       c("Comparison", "MissGlobal", "MissComp", "MissCND1", "MissCND2"))
```

`MissGlobal` describes how well the protein group is covered in the experiment
as a whole, which is useful for deciding how much to trust it at all.
`MissComp` and the two per-condition columns describe the support behind one
particular contrast. None of them determines whether a change is biologically
real, but together they distinguish estimates based mainly on measured
intensities from estimates that depend largely on imputed values.

The relationship between missingness and the differential call can be summarised
directly. `MissComp` is the right column here, because what is being related to
the call is the evidence available for that particular contrast:

```{r missing-cols}
bins <- cut(res$DEPs_results$MissComp, c(-1, 0, 25, 50, 100),
            labels = c("0%", "1-25%", "26-50%", ">50%"))
table(missing = bins, res$DEPs_results$Change)
```

In this example dataset, most results with no missing values are called
`No Change`. Among results with 26–50 % missing values, by contrast, there is a
marked asymmetry: 619 are called `Down` against 70 called `Up`.

This pattern does not prove that all those missing values are MNAR, but it is
consistent with a substantial abundance-dependent component of missingness. When
a protein group stops being quantified in one condition, values imputed from the
lower end of the distribution can produce a large negative change.

Interpretation must always be made in the context of the missingness pattern and
the imputation method used.

## Detection and non-detection patterns between conditions

The most extreme case arises when a protein group is quantified in no replicate
of one condition but in every replicate of the other:

```{r onoff}
onoff <- subset(res$DEPs_results,
                (MissCND1 == 100 & MissCND2 == 0) |
                (MissCND1 == 0   & MissCND2 == 100))
nrow(onoff)
head(onoff[, c("Protein.IDs", "Comparison", "logFC", "adj.P.Val",
               "MissCND1", "MissCND2")], 4)
```

Roughly 500 results in this analysis have 100 % missing values in one group and
0 % in the other. These are detection versus non-detection patterns between
conditions. They may represent biologically relevant changes, but they call for a
different interpretation from the one applied to a ratio computed solely from
observed intensities.

A logFC of around −11, for example, should not be read as a directly measured
abundance ratio. Its magnitude depends largely on the value introduced by the
MNAR method, and it states that one condition lies below the quantifiable range
relative to the other.

The `MissCND1` and `MissCND2` columns make that provenance visible, whereas
the p-value alone says nothing about how many observations were measured and how
many were imputed. `vignette("missing-values")` develops this distinction and
explains how the different imputation strategies affect these results.

# Visualising the results

`process_proteomics()` prepares three tables that can be used directly by the
visualisation modules:

- `res$DEPs_results`, for the *volcano plots*;
- `res$BoxPlot_Input`, to compare the sample distributions across the different
  assays;
- `res$PCA_Input`, for PCA and heatmaps.

The visualisation functions take these tables rather than the complete
`SummarizedExperiment`, which means they can also be applied to results that have
been modified or produced by other workflows, provided they keep the expected
structure.

## Interactive figures

NADIA produces interactive volcano plots, boxplots and PCA projections through
Highcharts. Each function returns a named list whose elements correspond —
depending on the module — to comparisons, assays or protein selection modes:

```{r hc, eval = FALSE}
volcano <- volcano_highchart_list(res$DEPs_results)
boxplot <- boxplot_highchart_list(res$BoxPlot_Input)
pca     <- pca_highchart_list(res$PCA_Input)

volcano[["B-A"]]
```

`volcano[["B-A"]]`, for example, selects the volcano plot for the `B-A`
comparison.

These figures are not rendered in this vignette, to keep the document small.
`vignette("visualization")` shows live versions of the three modules and explains
their main customisation options. `vignette("results-and-export")` describes how
to save each widget as an HTML file.

## Static figure

`proteomics_heatmap()` produces a static heatmap from `res$PCA_Input`. The
function can select which protein groups are shown itself.

In the example below, `mode = "target"` restricts the figure to the protein
groups significant in the comparison named through `comparison`:

```{r heatmap, message = FALSE, warning = FALSE, fig.width = 6.5, fig.height = 5.5, fig.alt = "Heatmap of the proteins significant in the B-A comparison"}
hm <- proteomics_heatmap(
    res$PCA_Input,
    mode            = "target",
    comparison      = "B-A",
    scale_data      = "row",
    sample_order    = "condition",
    condition_order = c("A", "B", "D"),
    show_row_names  = FALSE,
    heatmap_title   = "Significant in B-A")
hm
```

The figure uses the following options:

- `comparison = "B-A"` selects the comparison of interest;
- `scale_data = "row"` standardises the profile of each protein group separately,
  so that colour represents relative changes between samples rather than
  differences in absolute abundance between proteins;
- `sample_order = "condition"` groups the samples by condition;
- `condition_order = c("A", "B", "D")` sets the order of those conditions
  explicitly;
- `show_row_names = FALSE` hides the identifiers, which would otherwise be
  illegible when many rows are drawn.

The heatmap therefore shows the relative patterns of the selected protein groups
across the A, B and D samples. It should not be read as a map of absolute protein
abundances.

Further options for selection, clustering, ordering and customisation are
described in `vignette("visualization")`.

# Next steps

This vignette has shown the complete workflow using NADIA's default
configuration. Those defaults provide a reproducible starting point, but they
should not be read as the best choice for every dataset.

One of the decisions with the largest effect on the final table is which
normalisation and imputation methods are used. `vignette("missing-values")`
explains the assumptions behind the available strategies, while
`vignette("choosing-methods")` shows how to compare them when no known reference
is available. When the experiment contains a spike-in,
`vignette("benchmarking")` evaluates them against the expected changes.

To import your own data, adjust figures or export the results, see the vignettes
linked in the introduction.

The session information used to generate this vignette is shown below:

```{r session}
sessionInfo()
```
