---
title: "Pattern Profiler: grouping proteins by their behaviour"
author: "Sergio Ciordia"
package: NADIA
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Pattern Profiler: grouping proteins by their behaviour}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", crop = NULL)
```

# Introduction

Differential expression answers "which proteins changed", one comparison at a
time. Across three or more conditions a second question appears: **which proteins
changed in the same shape?** Proteins that rise, peak and fall together are
candidates for a shared mechanism, and no pairwise contrast can reveal that.

The Pattern Profiler answers it with Mfuzz's fuzzy c-means clustering. *Fuzzy* is
the important word: instead of assigning each protein to one cluster, it gives a
**membership** between 0 and 1 for every cluster. A protein with a clean profile
scores high in one; an ambiguous one spreads its membership around, and can be
filtered out — which hard clustering has no way to express.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("NADIA")
```

```{r library, message = FALSE}
library(NADIA)
data(nadia_dia)
```

This module needs `Mfuzz`, which is optional. It also needs `Biobase` and `e1071`
**attached**, not merely installed, because Mfuzz calls `exprs()` and `cmeans()`
without qualifying them. `pattern_profiler_analysis()` attaches them on entry and
releases them with `on.exit()`, leaving your search path as it found it — so this
is a note on how it works, not something you have to do.

```{r deps}
have_mfuzz <- all(vapply(c("Mfuzz", "Biobase", "e1071"),
                         requireNamespace, logical(1), quietly = TRUE))
have_mfuzz
```

# Running it

```{r run, eval = have_mfuzz, message = FALSE, warning = FALSE}
res <- process_proteomics(nadia_dia, verbose = FALSE)

pp <- pattern_profiler_analysis(
    res$se_proc, res$DEPs_results,
    assay_name    = "Impseqrob_min",
    filter_mode   = "any",
    alpha         = 0.05,
    auto_select_c = FALSE,   # fixed here to keep the vignette fast
    c             = 3,
    verbose       = FALSE)

c(features_in = pp$n_features_input,
  features_clustered = pp$n_features_final,
  rows_out = pp$n_rows_output,
  clusters = pp$optimal_c)
```

Each protein is reduced to one value per condition — the median across replicates
by default, `aggregate = "mean"` for the alternative — and z-scored, so that the
*shape* of the profile is clustered rather than its absolute level.

```{r long-output, eval = have_mfuzz}
head(pp$long_output, 4)
```

One row per protein **per cluster it belongs to**, with its membership and its
z-scored profile. Which is why 794 proteins produced 948 rows.

# What soft clustering actually buys you

That row count is the whole point, and `summarize_pattern_profiler()` makes it
explicit:

```{r summary, eval = have_mfuzz}
s <- summarize_pattern_profiler(pp$long_output)

c(unique_features   = s$n_unique_features,
  total_rows        = s$n_total_entries,
  in_more_than_one  = s$n_multi_cluster_features)

s$cluster_summary
```

A hundred and twenty-eight proteins do not belong cleanly to a single cluster.
Hard clustering would have assigned each of them to exactly one anyway, silently,
and you would have had no way to tell those apart from the confident assignments.

Note the mean membership per cluster in the table above: cluster 1 averages 0.86,
cluster 2 only 0.48. That is the signature of a cluster that is not really a
pattern so much as a place where leftovers accumulate, and it is worth knowing
before drawing conclusions from it.

`min_membership` is the control:

```{r membership, eval = have_mfuzz}
thresholds <- c(0, 0.25, 0.5, 0.7, 0.9)
data.frame(min_membership = thresholds,
           rows_kept = vapply(thresholds,
                              function(m) sum(pp$long_output$Membership >= m),
                              numeric(1)))
```

A protein whose highest membership is 0.4 across three clusters is not really in
any of them. Whether to draw it is a presentation decision, and this is the number
that lets you make it deliberately.

# Which proteins go in

`filter_mode` decides, and one of its values is easy to misread:

| `filter_mode` | Proteins included |
|---|---|
| `"any"` | significant in **at least one** comparison |
| `"specific"` | significant in the single comparison named by `comparison` |
| `"all"` | **every protein**, with no significance filter at all |

`"all"` does *not* mean "significant in all comparisons". It means no filter.

```{r filter-modes, eval = have_mfuzz, message = FALSE, warning = FALSE}
pp_all <- pattern_profiler_analysis(
    res$se_proc, res$DEPs_results, assay_name = "Impseqrob_min",
    filter_mode = "all", auto_select_c = FALSE, c = 3, verbose = FALSE)

c(any = pp$n_features_input, all = pp_all$n_features_input)
```

Two and a half times as many proteins. If that is what you want it is the right
choice; reaching for it expecting the stricter reading would quietly cluster
everything, most of it noise.

# Choosing the number of clusters

`c` is the one parameter with no defensible default, and `auto_select_c = TRUE`
(the real default) chooses it by sweeping `c_range`:

```{r auto-c, eval = have_mfuzz, message = FALSE, warning = FALSE}
pp_auto <- pattern_profiler_analysis(
    res$se_proc, res$DEPs_results, assay_name = "Impseqrob_min",
    c_range = 2:5, auto_select_c = TRUE, selection_method = "xb",
    verbose = FALSE)

pp_auto$optimal_c
pp_auto$selection_metrics
```

Four indices are computed per candidate `c`:

- **XB** — the Xie-Beni index, compactness against separation. **Lower is
  better**, and `selection_method = "xb"` minimises it.
- **FPC** — fuzzy partition coefficient, how crisp the assignment is. Higher is
  better.
- **AMM** — average maximum membership.
- **Dmin** — minimum distance between centroids.

Here XB is minimised at `c = 2`, and note that it increases monotonically with
`c` in this range. That is common and it is a caution rather than an answer: with
three conditions there is a limit to how many distinguishable shapes exist, and
an index that always prefers fewer clusters is telling you the data do not
strongly support more. `"consensus"` combines the indices for when they disagree,
and `"elbow"` looks for where the improvement flattens.

## Two things that would corrupt this choice

Both are handled, and both are worth knowing about because they fail silently.

A clustering run can fail to converge for a given seed. Such a run is recorded as
`NA`, **not** `0`. A zero would look like a perfect XB score and drag the mean
towards whichever `c` crashed most often — choosing the number of clusters by
which one failed.

And features with zero variance across conditions are dropped before z-scoring.
Dividing by a zero standard deviation gives `NaN`, and a matrix containing `NaN`
does not cluster at all. The count is reported so the drop is never invisible:

```{r constant, eval = have_mfuzz}
c(before = pp$n_features_input, after = pp$n_features_final)
```

# The profiles

Two interactive views. The first draws every protein in a cluster, with the
centroid on top:

```{r profiles, eval = FALSE}
plots <- cluster_profile_highchart_list(pp$long_output, min_membership = 0.5)
plots[[1]]
```

The second overlays only the centroids, which is the summary figure:

```{r centroids, eval = FALSE}
cluster_centroids_highchart(pp$long_output, min_membership = 0.5)
```

Both are omitted from this vignette's output only to keep it small.

# Saving and reloading

The analysis can write its table to disk, and the plotting functions read it
back, so a figure can be redrawn without repeating the clustering:

```{r roundtrip, eval = have_mfuzz, message = FALSE, warning = FALSE}
f <- file.path(tempdir(), "pattern_profiler.tsv")

invisible(pattern_profiler_analysis(
    res$se_proc, res$DEPs_results, assay_name = "Impseqrob_min",
    auto_select_c = FALSE, c = 3, output_file = f, verbose = FALSE))

reloaded <- read_pattern_profiler_data(f)
identical(dim(reloaded), dim(pp$long_output))
unlink(f)
```

# A caveat about the number of conditions

Soft clustering groups *profiles*, and a profile needs enough points to have a
shape. With two conditions every protein is a straight line and there are only
two shapes available, up or down — at which point the clustering has degenerated
into the sign of the fold change, which the differential expression table already
gave you.

Three conditions is the practical minimum. It is why the example dataset keeps
three of its original four.

```{r session}
sessionInfo()
```
