---
title: "Results and export: sharing and preserving an analysis"
author: "Sergio Ciordia"
package: NADIA
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Results and export: sharing and preserving an analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  crop = NULL,
  message = FALSE,
  warning = FALSE
)
options(width = 100)

have_pattern_profiler <- all(vapply(
  c("Mfuzz", "Biobase", "e1071"),
  requireNamespace,
  logical(1),
  quietly = TRUE
))
```

<style type="text/css">
.function-parameters,
.code-details {
  margin: 1em 0 1.4em;
  border: 1px solid #c8d3dc;
  border-radius: 5px;
  background: #f7f9fb;
}
.function-parameters summary,
.code-details summary {
  padding: 0.75em 1em;
  color: #1d3557;
  font-weight: 600;
  cursor: pointer;
}
.function-parameters[open] summary,
.code-details[open] summary {
  border-bottom: 1px solid #c8d3dc;
}
.function-parameters .parameter-content {
  padding: 0.2em 1em 0.7em;
}
.code-details pre,
.code-details table {
  margin: 1em;
}
.export-warning {
  box-sizing: border-box;
  margin: 1.15em 0 1.5em;
  padding: 0.9em 1.1em;
  border: 1px solid #e4c78e;
  border-left: 5px solid #c98718;
  border-radius: 5px;
  background: #fffaf0;
  box-shadow: 0 1px 2px rgba(112, 78, 20, 0.08);
}
.export-warning p {
  margin: 0;
  padding: 0 !important;
}
.export-warning strong:first-child {
  color: #7a5310;
}
</style>

# Introduction

A proteomics analysis should finish with outputs that other people can inspect
and that the analyst can reproduce later. NADIA supports several complementary
routes: selected tables can be exported as TSV or Parquet files, a complete run
can be archived in one `.nadia` file, and interactive tables or figures can be
shared as HTML.

This vignette explains when to use each route, how to create the corresponding
files, and how to verify what has been saved. The examples use the bundled DIA
dataset so that every output can be reproduced without external files.

# Prepare an example analysis

This section creates one NADIA result that will be reused throughout the
vignette. Keeping a single analysis object also makes it easier to compare the
different export formats.

## Install and load NADIA

Install NADIA through Bioconductor if it is not already available, and then load
the package and example data.

```{r installation, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("NADIA")
```

```{r load-package}
library(NADIA)
data(nadia_dia)
```

## Run the processing workflow

`process_proteomics()` returns all analysis components in memory. Because
`export_dir` is not supplied here, this call does not create any files.

```{r run-analysis}
res <- process_proteomics(
  nadia_dia,
  verbose = FALSE
)

export_dir  <- file.path(tempdir(), "nadia_vignette_tables")
nadia_file  <- file.path(tempdir(), "nadia_vignette_analysis.nadia")
parquet_dir <- file.path(tempdir(), "nadia_vignette_parquet")
widget_dir  <- file.path(tempdir(), "nadia_vignette_widget")
```

A Pattern Profiler run is added as well. It is optional throughout this
vignette, but it is what allows the archive to hold the expression patterns
alongside the statistics, and it is used again in the section on functional
input.

```{r run-pattern-profiler, eval = have_pattern_profiler, results = "hide"}
pp <- pattern_profiler_analysis(
  res$se_proc,
  res$DEPs_results,
  assay_name = "Impseqrob_min",
  seed       = 123,
  verbose    = FALSE
)
```

```{r pattern-profiler-fallback, include = FALSE}
if (!have_pattern_profiler) pp <- NULL
```

The returned object contains the processed experiment, differential-abundance
results, plotting tables, comparison names, and the parameters used in the run.

```{r result-components}
component_summary <- data.frame(
  Component = c(
    "se_proc",
    "DEPs_results",
    "BoxPlot_Input",
    "PCA_Input",
    "comparisons",
    "parameters"
  ),
  Size = c(
    paste(dim(res$se_proc), collapse = " x "),
    paste(dim(res$DEPs_results), collapse = " x "),
    paste(dim(res$BoxPlot_Input), collapse = " x "),
    paste(dim(res$PCA_Input), collapse = " x "),
    paste(length(res$comparisons), "comparisons"),
    paste(length(res$parameters), "entries")
  ),
  Purpose = c(
    "Processed assays and sample metadata",
    "Differential-abundance statistics",
    "Long-format intensities for boxplots",
    "Long-format intensities and significance flags for PCA",
    "Available pairwise comparisons",
    "Analysis settings and recorded function call"
  ),
  check.names = FALSE
)

knitr::kable(
  component_summary,
  caption = "Main components returned by process_proteomics().",
  align = c("l", "l", "l")
)
```

The dimensions show why a single export format is not ideal for every task. A
small parameter list is convenient inside R, whereas long plotting tables are
often easier to exchange as files.

# Choose an output for the task

The most appropriate output depends on who will use it and whether they need
selected results or the complete analysis. The following table summarises the
four main options used in this vignette.

```{r export-route-table}
export_routes <- data.frame(
  Output = c(
    "TSV files",
    "Parquet files",
    "One .nadia archive",
    "Interactive HTML"
  ),
  Best_for = c(
    "Inspection in text editors, spreadsheets, and general-purpose software",
    "Typed exchange of large tables and matrices",
    "Preserving and restoring a complete NADIA analysis",
    "Exploring or presenting results without writing R code"
  ),
  Main_advantage = c(
    "Human-readable and widely supported",
    "Preserves column types and is efficient to read",
    "Keeps data, parameters, provenance, and optional clustering together",
    "Provides sorting, filtering, search, and interactive graphics"
  ),
  Main_consideration = c(
    "Several related files must be kept together",
    "Requires software with Parquet support",
    "Requires NADIA or DuckDB-compatible tools to inspect its contents",
    "May require an accompanying dependency folder unless saved as one file"
  ),
  check.names = FALSE
)

knitr::kable(
  export_routes,
  col.names = c("Output", "Best use", "Main advantage", "Main consideration"),
  caption = "Choosing an output according to its intended use.",
  align = c("l", "l", "l", "l")
)
```

These options are complementary. For example, a `.nadia` archive can preserve
the full analysis while a TSV summary and an HTML table are supplied to
collaborators for immediate review.

# Export tabular results

`process_proteomics()` can write the processed matrices and plotting inputs as
part of the analysis call. File export is activated only when `export_dir` is
provided, which prevents an ordinary in-memory analysis from creating files
unexpectedly.

<details class="function-parameters">
<summary>Main export arguments of <code>process_proteomics()</code></summary>
<div class="parameter-content">

- `export_dir` gives the destination directory. Its default value, `NULL`,
  disables every file export.
- `export_format` accepts `"tsv"`, `"parquet"`, or `"both"`. Parquet export
  through `process_proteomics()` requires the `arrow` package.
- `export_normalized` and `export_imputed` control the processed intensity
  matrices.
- `export_volcano`, `export_boxplot`, and `export_pca` control the tables used by
  the corresponding visualisation functions.

</div>
</details>

## Create a set of TSV files

The next call writes all default outputs to a temporary directory. In a real
analysis, replace `export_dir` with a permanent project directory.

```{r export-tsv}
unlink(export_dir, recursive = TRUE)

invisible(process_proteomics(
  nadia_dia,
  export_dir = export_dir,
  export_format = "tsv",
  verbose = FALSE
))

exported_files <- list.files(export_dir, full.names = TRUE)

file_contents <- ifelse(
  grepl("matrix_log2.*Imp", basename(exported_files)),
  "Imputed intensity matrix",
  ifelse(
    grepl("matrix_log2", basename(exported_files)),
    "Normalised intensity matrix",
    ifelse(
      grepl("VolcanoPlot", basename(exported_files)),
      "Differential-abundance results",
      ifelse(
        grepl("BoxPlot", basename(exported_files)),
        "Long-format boxplot input",
        "Long-format PCA input"
      )
    )
  )
)

export_manifest <- data.frame(
  File = basename(exported_files),
  Contents = file_contents,
  Size = sprintf("%.1f KB", file.info(exported_files)$size / 1024),
  check.names = FALSE
)

knitr::kable(
  export_manifest,
  caption = "Files created by the default TSV export.",
  align = c("l", "l", "r")
)
```

The filenames record the normalisation and imputation methods, reducing the
risk of mixing tables from different runs. The plotting inputs are intentionally
long: each row represents an observation required by its corresponding plot,
whereas the two matrix files keep one protein per row.

Individual switches can be disabled when only a subset is needed. This example
exports the final imputed matrix and the differential-abundance table, but omits
the normalised matrix and the boxplot and PCA inputs.

```{r selective-export, eval = FALSE}
process_proteomics(
  nadia_dia,
  export_dir = "results/selected_tables",
  export_format = "tsv",
  export_normalized = FALSE,
  export_imputed = TRUE,
  export_volcano = TRUE,
  export_boxplot = FALSE,
  export_pca = FALSE,
  verbose = FALSE
)
```

## Choose between TSV and Parquet

TSV is convenient when people need to open or inspect a table directly.
Parquet is preferable when preserving column types and reading large datasets
efficiently are more important than human readability.

```{r format-comparison}
format_guide <- data.frame(
  Property = c(
    "Readable as plain text",
    "Preserves column types",
    "Convenient for spreadsheets",
    "Efficient for repeated programmatic access"
  ),
  TSV = c("Yes", "No; types are inferred when imported", "Yes", "Moderate"),
  Parquet = c("No", "Yes", "Requires Parquet support", "Yes"),
  check.names = FALSE
)

knitr::kable(
  format_guide,
  caption = "Practical differences between TSV and Parquet exports.",
  align = c("l", "l", "l")
)
```

Use `export_format = "both"` when the same results must support manual review
and an automated downstream workflow.

```{r export-both, eval = FALSE}
process_proteomics(
  nadia_dia,
  export_dir = "results/tsv_and_parquet",
  export_format = "both",
  verbose = FALSE
)
```

<div class="export-warning">
<p><strong>Metrics exports.</strong> The metric functions
<code>normalization_metrics()</code>, <code>imputation_metrics()</code>, and
<code>benchmarking_multiple()</code> also use an output directory to activate
file creation. Their export flags control what is written once a directory is
supplied; with <code>output_dir = NULL</code>, the results remain in memory.</p>
</div>

# Archive a complete analysis

Separate tables are useful for exchange, but they do not by themselves retain
the complete relationship between preprocessing, processing, parameters, and
provenance. `write_nadia()` stores these elements in a single DuckDB database
using the `.nadia` extension.

## Write and read a `.nadia` file

The required input is the preprocessing object. The processing result and a
Pattern Profiler result can be included when available.

<details class="function-parameters">
<summary>Main arguments of <code>write_nadia()</code></summary>
<div class="parameter-content">

- `file` is the path of the archive to create.
- `preprocessing` is the `proteomics_data` object returned by a NADIA
  preprocessing function.
- `result` is the optional output from `process_proteomics()`.
- `pattern_profiler` optionally adds the output from
  `pattern_profiler_analysis()`.
- `overwrite = TRUE` explicitly replaces an existing archive.

</div>
</details>

```{r write-nadia, eval = requireNamespace("duckdb", quietly = TRUE)}
unlink(c(nadia_file, paste0(nadia_file, ".wal")))

write_nadia(
  nadia_file,
  preprocessing    = nadia_dia,
  result           = res,
  pattern_profiler = pp,
  verbose          = FALSE
)

db <- read_nadia(nadia_file)

archive_summary <- data.frame(
  Property = c(
    "Archive size",
    "Samples",
    "Proteins",
    "Normalised assay",
    "Imputed assay",
    "Pattern Profiler stored"
  ),
  Value = c(
    sprintf("%.1f MB", file.size(nadia_file) / 1024^2),
    db$meta[["n_samples"]],
    db$meta[["n_proteins"]],
    db$meta[["norm_assay"]],
    db$meta[["imputed_assay"]],
    db$meta[["has_pattern_profiler"]]
  ),
  check.names = FALSE
)

knitr::kable(
  archive_summary,
  caption = "Summary of the example .nadia archive.",
  align = c("l", "l")
)
```

The archive contains the full processing run used here, together with the
Pattern Profiler result. Clustering can be supplied to `write_nadia()`, as
above, or added to an existing archive later with
`nadia_add_pattern_profiler()`.

## Restore and verify the result

`read_nadia()` exposes the archive metadata and canonical tables.
`nadia_result()` reconstructs the `proteomics_result` expected by NADIA plotting
and reporting functions, while `nadia_preprocessing()` restores the original
preprocessing object.

```{r verify-roundtrip, eval = requireNamespace("duckdb", quietly = TRUE)}
restored <- nadia_result(db)

roundtrip_check <- data.frame(
  Component = c(
    "Differential-abundance results",
    "PCA input",
    "Final imputed assay"
  ),
  Identical = c(
    identical(restored$DEPs_results, res$DEPs_results),
    identical(restored$PCA_Input, res$PCA_Input),
    identical(
      SummarizedExperiment::assay(restored$se_proc, "Impseqrob_min"),
      SummarizedExperiment::assay(res$se_proc, "Impseqrob_min")
    )
  ),
  check.names = FALSE
)

knitr::kable(
  roundtrip_check,
  caption = "Round-trip verification after restoring the archive.",
  align = c("l", "c")
)
```

All three checks return `TRUE`, confirming that the tables, their column types,
and the final assay are reproduced exactly in this example.

## Inspect stored tables and provenance

The archive avoids unnecessary duplication. Values that can be reconstructed
exactly, such as the log2 assay and `PCA_Input`, are exposed through SQL views
rather than stored as additional copies. For imputation methods that leave
observed values unchanged, only the replaced cells and their MAR or MNAR branch
need to be stored.

```{r nadia-views, eval = requireNamespace("duckdb", quietly = TRUE)}
archive_views <- subset(nadia_tables(nadia_file), type == "view")

view_description <- c(
  v_boxplot_input = "Boxplot input",
  v_deps_results = "Differential-abundance results",
  v_deps_pattern_profiler = "Differential abundance joined to the clusters",
  v_matrix_imputed = "Final imputed matrix",
  v_matrix_norm = "Normalised matrix",
  v_metadata = "Sample metadata",
  v_pattern_profiler = "Cluster memberships and standardised profiles",
  v_pca_input = "PCA input",
  v_protein_id = "Protein identification table",
  v_protein_quant = "Protein quantification table"
)

archive_views$Contents <- unname(view_description[archive_views$name])

knitr::kable(
  archive_views[, c("name", "rows", "Contents")],
  col.names = c("View", "Rows", "Reconstructed contents"),
  caption = "Canonical views available in the example archive.",
  align = c("l", "r", "l"),
  row.names = FALSE
)
```

The views provide the same analysis tables that users normally access in R.
The archive also records package versions, function calls, source-file
information, and analysis parameters.

```{r nadia-provenance, eval = requireNamespace("duckdb", quietly = TRUE)}
provenance_parameters <- subset(
  db$parameters,
  name %in% c(
    "norm_method", "imp_method", "mar_method", "mnar_method",
    "de_method", "alpha", "logFC_threshold"
  )
)

knitr::kable(
  provenance_parameters[, c("step", "name", "value")],
  col.names = c("Stage", "Parameter", "Recorded value"),
  caption = "Selected parameters stored in the .nadia archive.",
  align = c("l", "l", "l"),
  row.names = FALSE
)
```

These values identify the main analysis choices rather than merely describing
the exported table. The recorded call retains more complex arguments that
cannot always be represented by a single scalar value.

<details class="code-details">
<summary>Show the recorded function calls</summary>

```{r nadia-calls, eval = requireNamespace("duckdb", quietly = TRUE)}
db$calls
```

</details>

## Query or extract the archive

`nadia_connect()` opens a DBI connection for targeted SQL queries. The example
below counts the three differential-abundance categories in each comparison
without reconstructing the complete result in R.

```{r nadia-sql, eval = requireNamespace("duckdb", quietly = TRUE)}
con <- nadia_connect(nadia_file)

sql_summary <- DBI::dbGetQuery(
  con,
  paste(
    'SELECT "Comparison", "Change", COUNT(*) AS "Proteins"',
    'FROM v_deps_results',
    'GROUP BY "Comparison", "Change"',
    'ORDER BY "Comparison", "Change"'
  )
)

DBI::dbDisconnect(con, shutdown = TRUE)

knitr::kable(
  sql_summary,
  caption = "Protein counts queried directly from the .nadia archive.",
  align = c("l", "l", "r")
)
```

The counts summarise each comparison while retaining the adjusted-P-value and
fold-change classification used by the original analysis.

For exchange with software that does not read `.nadia` files,
`nadia_export_parquet()` extracts every canonical view as a Parquet file. This
export is performed by DuckDB directly and therefore does not require `arrow`.

```{r nadia-to-parquet, eval = requireNamespace("duckdb", quietly = TRUE)}
unlink(parquet_dir, recursive = TRUE)
invisible(nadia_export_parquet(db, parquet_dir, verbose = FALSE))

parquet_manifest <- data.frame(
  File = list.files(parquet_dir),
  check.names = FALSE
)

knitr::kable(
  parquet_manifest,
  caption = "Parquet tables extracted from the .nadia archive.",
  align = "l"
)
```

<div class="export-warning">
<p><strong>Working location.</strong> Do not keep an open <code>.nadia</code>
database inside a folder that is being synchronised by iCloud Drive, Dropbox,
or OneDrive. Write and use the archive in a local directory, close it, and then
copy the completed file to the synchronised location.</p>
</div>

## Combine differential abundance with clustering

A functional or enrichment analysis needs two things at once: the statistics
that say which proteins changed, and the cluster that says which expression
pattern they follow. `deps_with_clusters()` builds that table in memory, and
`nadia_deps_with_clusters()` reads the equivalent table from an archive. Both
return the columns of `DEPs_results` followed by `Cluster`, `Membership`, and
`ClusterRank`.

<details class="function-parameters">
<summary>Main arguments of <code>deps_with_clusters()</code></summary>
<div class="parameter-content">

- `assignment` selects the shape of the result. `"primary"`, the default,
  keeps the cluster of highest membership and returns one row per protein and
  comparison. `"all"` keeps every cluster a protein belongs to.
- `min_membership` applies an additional membership threshold. It can only be
  stricter than the one used during the analysis, because weaker associations
  were discarded when the clustering result was built.
- `comparison` restricts the result to selected comparisons.
- `significant_only` keeps only the proteins classified as `Up` or `Down`,
  using the classification made during the analysis.

</div>
</details>

```{r functional-input, eval = have_pattern_profiler}
functional_input <- deps_with_clusters(
  res$DEPs_results,
  pp,
  assignment = "primary"
)

knitr::kable(
  head(functional_input[
    !is.na(functional_input$Cluster),
    c("Protein.IDs", "Comparison", "logFC", "adj.P.Val", "Change",
      "Cluster", "Membership")
  ], 6),
  caption = "Differential-abundance results with the dominant cluster of each protein.",
  align = c("l", "l", "r", "r", "l", "r", "r"),
  row.names = FALSE,
  digits = 4
)
```

Soft clustering allows a protein to belong to several clusters at once, so the
two values of `assignment` produce tables of different lengths. The row counts
below show the difference, and confirm that the complete set of tested proteins
is retained in both cases.

```{r assignment-modes, eval = have_pattern_profiler}
soft <- deps_with_clusters(res$DEPs_results, pp, assignment = "all")

assignment_summary <- data.frame(
  Table = c(
    "Differential-abundance results",
    "assignment = \"primary\"",
    "assignment = \"all\""
  ),
  Rows = c(nrow(res$DEPs_results), nrow(functional_input), nrow(soft)),
  Proteins_without_cluster = c(
    NA_integer_,
    sum(is.na(functional_input$Cluster)),
    sum(is.na(soft$Cluster))
  ),
  check.names = FALSE
)

knitr::kable(
  assignment_summary,
  col.names = c("Table", "Rows", "Proteins without a cluster"),
  caption = "Effect of the assignment mode on the number of rows.",
  align = c("l", "r", "r")
)
```

The join is a left join from the differential-abundance results, so every
protein that was tested is retained even when it has no cluster. This matters
for enrichment analysis: the complete set of tested proteins defines the
background against which any over-representation is measured, and removing part
of it changes every result computed from it. A protein receives `NA` when it was
not selected for clustering, or when the clustering did not retain it.

Inside the archive the same table is the view `v_deps_pattern_profiler`, so it
is also available to SQL clients and is included in the Parquet export.

```{r functional-input-sql, eval = have_pattern_profiler && requireNamespace("duckdb", quietly = TRUE)}
con <- nadia_connect(nadia_file)

cluster_counts <- DBI::dbGetQuery(
  con,
  paste(
    'SELECT "Cluster", "Change", COUNT(*) AS "Proteins"',
    'FROM v_deps_pattern_profiler',
    'WHERE "ClusterRank" = 1 AND "Cluster" IS NOT NULL',
    'AND "Comparison" = \'B-A\'',
    'GROUP BY "Cluster", "Change"',
    'ORDER BY "Cluster", "Change"'
  )
)

DBI::dbDisconnect(con, shutdown = TRUE)

knitr::kable(
  cluster_counts,
  caption = "Differential-abundance categories per cluster in comparison B-A.",
  align = c("r", "l", "r")
)
```

The same table read back through R gives an identical result.

```{r functional-input-roundtrip, eval = have_pattern_profiler && requireNamespace("duckdb", quietly = TRUE)}
identical(nadia_deps_with_clusters(nadia_file), functional_input)
```

Reduced to its `FeatureID` and `Cluster` columns, this table is also accepted
directly as the `row_annotation` argument of `proteomics_heatmap()`, which
annotates each row of a heatmap with the cluster it belongs to.

# Explore results with interactive tables

NADIA provides interactive tables for differential-abundance results, protein
annotations, quantification values, and sample metadata. Use a
`*_reactable()` function when only the table is required, or a `*_widget()`
function when a complete page with additional controls and Excel export is
needed.

```{r interactive-table-guide}
interactive_tables <- data.frame(
  Purpose = c(
    "Differential-abundance results",
    "Protein identification and annotation",
    "Raw and processed quantification",
    "Sample metadata and identification counts"
  ),
  Function = c(
    "results_list_reactable() / results_list_widget()",
    "protein_list_reactable() / protein_list_widget()",
    "quant_list_widget()",
    "summary_list_widget()"
  ),
  Required_input = c(
    "DEPs_results",
    "protein_id",
    "protein_quant and a processed matrix",
    "metadata"
  ),
  Display_unit = c(
    "One protein per comparison",
    "One protein with sample-level identification metrics",
    "One protein with raw and processed sample values",
    "One sample"
  ),
  check.names = FALSE
)

knitr::kable(
  interactive_tables,
  col.names = c("Table contents", "Function", "Required input", "Displayed unit"),
  caption = "Interactive tables available for the main NADIA outputs.",
  align = c("l", "l", "l", "l")
)
```

## Differential-abundance results

`results_list_widget()` creates a complete results page with sorting, search,
filters, and Excel export. Supplying `protein_quant` adds protein descriptions
and the number of quantified peptides to the statistical results.

```{r results-widget, eval = FALSE}
results_page <- results_list_widget(
  res$DEPs_results,
  protein_quant = nadia_dia$protein_quant,
  comparisons = "B-A",
  alpha = 0.05,
  page_size = 15
)

results_page
```

The complete interactive table may contain thousands of rows. A compact static
summary helps establish what the user will find when the page is opened.

```{r results-summary-table}
change_counts <- as.data.frame(
  with(res$DEPs_results, table(Comparison, Change)),
  stringsAsFactors = FALSE
)
names(change_counts) <- c("Comparison", "Classification", "Proteins")
change_counts <- change_counts[
  order(
    change_counts$Comparison,
    factor(
      change_counts$Classification,
      levels = c("Up", "Down", "No Change")
    )
  ),
]
row.names(change_counts) <- NULL

knitr::kable(
  change_counts,
  caption = "Number of proteins in each differential-abundance category.",
  align = c("l", "l", "r")
)
```

For example, comparison `B-A` contains both up- and down-regulated proteins,
while the remaining proteins are labelled `No Change` under the selected
adjusted-P-value and fold-change thresholds. The widget allows any of these
groups to be isolated without modifying the underlying result.

## Protein, quantification, and sample tables

The remaining widgets focus on the inputs and processed measurements rather
than differential-abundance statistics. Their examples are grouped below to
keep the main workflow concise.

<details class="code-details">
<summary>Show code for the protein, quantification, and sample tables</summary>

```{r other-widgets, eval = FALSE}
# Protein annotations and sample-level identification metrics.
protein_page <- protein_list_widget(
  nadia_dia$protein_id,
  metadata = nadia_dia$metadata
)

# Add the final processed matrix to the raw quantification table.
imputed_matrix <- SummarizedExperiment::assay(
  res$se_proc,
  "Impseqrob_min"
)
matrix_data <- data.frame(
  ProteinGroups = rownames(imputed_matrix),
  imputed_matrix,
  check.names = FALSE
)

quantification_page <- quant_list_widget(
  nadia_dia$protein_quant,
  matrix_data = matrix_data,
  metadata = nadia_dia$metadata
)

# One row per sample, with the metadata fields that are available.
sample_page <- summary_list_widget(nadia_dia$metadata)
```

</details>

`quant_list_widget()` matches the processed matrix to the quantification table
through `ProteinGroups`; a matrix without this identifier cannot be joined
reliably. `summary_list_widget()` adapts to the available metadata, so fields
that are absent from a particular input format are omitted rather than treated
as an error.

## Export a table to Excel without internet access

Each complete `*_widget()` page includes an **Export to Excel** button. The
workbook is generated in the browser with formatted headers, suitable column
widths, and numeric cells stored as numbers.

```{r offline-assets}
offline_assets <- data.frame(
  Library = c("ExcelJS", "PapaParse"),
  Role = c(
    "Creates the formatted .xlsx workbook",
    "Parses tabular text during browser-side export"
  ),
  Included_with_NADIA = c("Yes", "Yes"),
  check.names = FALSE
)

knitr::kable(
  offline_assets,
  col.names = c("Library", "Role", "Included with NADIA"),
  caption = "JavaScript libraries used by the offline Excel export.",
  align = c("l", "l", "c")
)
```

Both libraries and their licence files are shipped with NADIA rather than
downloaded from a content-delivery network. The Excel button therefore remains
available when the saved page is opened offline, provided its local dependency
folder accompanies the HTML file.

# Save interactive HTML

An interactive result can be saved either as a complete page plus a dependency
folder or, for an individual `htmlwidget`, as one self-contained HTML file. The
choice affects portability and which page controls are retained.

## Save a complete widget page

Objects returned by `results_list_widget()`, `protein_list_widget()`,
`quant_list_widget()`, and `summary_list_widget()` are complete HTML pages.
`htmltools::save_html()` saves the page and copies its JavaScript and CSS
dependencies into a neighbouring `lib` directory.

```{r save-widget-page}
unlink(widget_dir, recursive = TRUE)
dir.create(widget_dir, recursive = TRUE)

results_page <- results_list_widget(
  res$DEPs_results,
  protein_quant = nadia_dia$protein_quant,
  comparisons = "B-A",
  page_size = 15,
  height = 600
)

page_file <- file.path(widget_dir, "results_B-A.html")
htmltools::save_html(results_page, page_file)

saved_page <- data.frame(
  Item = c("HTML page", "Dependency directory"),
  Created = c(
    file.exists(page_file),
    dir.exists(file.path(widget_dir, "lib"))
  ),
  check.names = FALSE
)

knitr::kable(
  saved_page,
  caption = "Files created for a complete interactive results page.",
  align = c("l", "c")
)
```

The HTML file and `lib` directory form one deliverable and must be copied
together. This option preserves the full NADIA toolbar, including its filters
and offline Excel button.

## Save an individual widget as one file

`results_list_reactable()` and the Highcharts plotting functions return actual
`htmlwidget` objects. They can be embedded into a single HTML file with
`htmlwidgets::saveWidget(..., selfcontained = TRUE)`.

```{r save-self-contained-widget}
result_table <- results_list_reactable(
  res$DEPs_results,
  protein_quant = nadia_dia$protein_quant,
  comparisons = "B-A",
  page_size = 15,
  height = 600
)

standalone_file <- file.path(widget_dir, "results_B-A_standalone.html")
htmlwidgets::saveWidget(
  result_table,
  standalone_file,
  selfcontained = TRUE
)

standalone_summary <- data.frame(
  File = basename(standalone_file),
  Size = sprintf("%.1f MB", file.size(standalone_file) / 1024^2),
  Dependency_directory_required = "No",
  check.names = FALSE
)

knitr::kable(
  standalone_summary,
  col.names = c("File", "Size", "Dependency directory required"),
  caption = "Self-contained HTML created from the Reactable table.",
  align = c("l", "r", "c")
)
```

The self-contained table retains Reactable features such as sorting and search,
but it does not include the additional toolbar supplied by
`results_list_widget()`. Embedding all data and libraries also increases the
size of the HTML file.

The same approach can be used for a Highcharts figure by selecting one chart
from the named list before saving it.

```{r save-highchart, eval = FALSE}
volcanoes <- volcano_highchart_list(
  res$DEPs_results,
  comparisons = "B-A",
  alpha = 0.05,
  lfc_thr = 1
)

htmlwidgets::saveWidget(
  volcanoes[["B-A"]],
  "volcano_B-A.html",
  selfcontained = TRUE
)
```

# Preserve reproducibility

Exported values are interpretable only when the analysis settings and source
data can be identified. NADIA returns the processing parameters in memory and
stores them automatically in a `.nadia` archive.

```{r processing-parameters}
selected_parameters <- data.frame(
  Parameter = c(
    "norm_method", "imp_method", "mar_method", "mnar_method",
    "de_method", "alpha", "logFC_threshold"
  ),
  Value = vapply(
    res$parameters[c(
      "norm_method", "imp_method", "mar_method", "mnar_method",
      "de_method", "alpha", "logFC_threshold"
    )],
    as.character,
    character(1)
  ),
  row.names = NULL,
  check.names = FALSE
)

knitr::kable(
  selected_parameters,
  caption = "Main processing parameters for the example analysis.",
  align = c("l", "l")
)
```

These values explain how the final assay and differential-abundance categories
were produced. They should accompany any exported table that may be interpreted
outside the original R session.

```{r reproducibility-checklist}
reproducibility_items <- data.frame(
  Keep = c(
    "Original proteomics report and sample metadata",
    "Analysis script or R Markdown document",
    "A .nadia archive or the exported result tables",
    "Processing parameters and software versions",
    "Any HTML dependency directory used for sharing"
  ),
  Reason = c(
    "Preserves the source measurements and experimental design",
    "Records the sequence of preprocessing and analysis steps",
    "Preserves the numerical outputs that were interpreted",
    "Identifies the exact methods, thresholds, and computational environment",
    "Keeps multi-file interactive pages functional offline"
  ),
  check.names = FALSE
)

knitr::kable(
  reproducibility_items,
  caption = "Recommended material to retain with a NADIA analysis.",
  align = c("l", "l")
)
```

Together, these items preserve both the evidence and the decisions that produced
the reported result. A `.nadia` archive reduces the number of files that must be
coordinated, but it should complement rather than replace the original report
and the analysis script.

```{r cleanup, include = FALSE}
unlink(
  c(
    export_dir,
    nadia_file,
    paste0(nadia_file, ".wal"),
    parquet_dir,
    widget_dir
  ),
  recursive = TRUE
)
```

# Session information

The package versions used to build this vignette are recorded below to support
reproducibility.

```{r session-information}
sessionInfo()
```
