---
title: "Results and export: getting the analysis out of R"
author: "Sergio Ciordia"
package: NADIA
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Results and export: getting the analysis out of R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", crop = NULL)
```

# Introduction

An analysis that only exists inside an R session is not finished. Collaborators
want to browse the results, reviewers want the numbers, and you want to be able to
reproduce the table a year from now.

This vignette covers the three ways out: **files** for archiving and for other
software, **interactive tables** for people who do not use R, and
**self-contained HTML** for sending a figure or a table to someone by email.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("NADIA")
```

```{r library, message = FALSE}
library(NADIA)
data(nadia_dia)
```

```{r run, message = FALSE}
res <- process_proteomics(nadia_dia, verbose = FALSE)
```

# Files

`process_proteomics()` writes nothing unless you give it somewhere to write. That
is deliberate: a function that creates files as a side effect of being called is
a nuisance in a script and a hazard in a package.

```{r export-dir, message = FALSE}
out <- file.path(tempdir(), "nadia_results")

invisible(process_proteomics(nadia_dia, export_dir = out,
                             export_format = "tsv", verbose = FALSE))

list.files(out)
```

`export_format` takes `"tsv"`, `"parquet"` or `"both"`. Parquet needs the `arrow`
package and is worth it for large matrices — it is typically several times
smaller and much faster to read back, and it preserves column types, which TSV
does not.

```{r parquet, eval = requireNamespace("arrow", quietly = TRUE), message = FALSE}
out_pq <- file.path(tempdir(), "nadia_parquet")
invisible(process_proteomics(nadia_dia, export_dir = out_pq,
                             export_format = "both", verbose = FALSE))

sizes <- file.info(list.files(out_pq, full.names = TRUE))["size"]
head(sizes[order(rownames(sizes)), , drop = FALSE], 6)
```

The individual export switches — `export_normalized`, `export_imputed`,
`export_volcano`, `export_boxplot`, `export_pca` — turn off anything you do not
need.

Two functions in the package have a trap worth knowing: `normalization_metrics()`,
`imputation_metrics()` and `benchmarking_multiple()` default to
`export_tables = TRUE`, so they *do* write files unless you pass
`output_dir = NULL`. Every example and vignette in this package passes it.

```{r no-write}
before <- length(list.files(tempdir()))
nm <- normalization_metrics(
    nm_run_normalizations(nm_prepare_se(nadia_dia, verbose = FALSE),
                          methods = c("log2Norm", "cycloess"), verbose = FALSE),
    output_dir = NULL, verbose = FALSE)
identical(before, length(list.files(tempdir())))
```

# Interactive tables

Four tables, each with sorting, filtering, search and an Excel button. The
`*_reactable()` form returns a `reactable` for embedding; the `*_widget()` form
returns a complete page with the search bar, filter panel and buttons around it.

## The results

```{r results-table, eval = FALSE}
results_list_widget(res$DEPs_results,
                    protein_quant = nadia_dia$protein_quant,
                    alpha = 0.05, lfc_thr = 1)
```

Passing `protein_quant` is what makes the expandable detail row work: clicking a
protein shows its intensities across samples, which is the quickest way to check
whether a fold change rests on real measurements or on imputed ones.

```{r results-reactable}
class(results_list_reactable(res$DEPs_results))
```

## The protein annotation

```{r protein-table}
class(protein_list_reactable(nadia_dia$protein_id,
                             metadata = nadia_dia$metadata))
```

## The quantification matrix

`quant_list_widget()` puts the raw quantification beside the processed matrix.
The processed matrix has to carry a `ProteinGroups` column, since a bare matrix
has no column to join on:

```{r quant-table}
mat <- SummarizedExperiment::assay(res$se_proc, "Impseqrob_min")
mat_df <- data.frame(ProteinGroups = rownames(mat), mat, check.names = FALSE)

class(quant_list_widget(nadia_dia$protein_quant,
                        matrix_data = mat_df,
                        metadata    = nadia_dia$metadata))
```

## The sample summary

```{r summary-table}
class(summary_list_widget(nadia_dia$metadata))
```

This one adapts to what it is given. Spectronaut metadata carry identification
counts that Proteome Discoverer does not export, so with a TMT or LFQ experiment
the table shows the columns that exist rather than failing:

```{r summary-degradation}
tmt <- preprocess_tmt(
    system.file("extdata", "nadia_tmt_report.tsv.gz", package = "NADIA"),
    condition_order = c("A", "B"), verbose = FALSE)

colnames(tmt$metadata)                        # no identification counts
class(summary_list_widget(tmt$metadata))      # still works
```

# The Excel button works offline

Each widget carries an "Export to Excel" button that builds a formatted `.xlsx`
in the browser — bold header, coloured fill, sensible column widths, numeric
columns typed as numbers rather than text.

It runs on two JavaScript libraries, **ExcelJS** and **PapaParse**, and both are
**shipped inside the package** rather than loaded from a CDN. So the button works
with no internet connection, and `htmlwidgets::saveWidget(selfcontained = TRUE)`
embeds them along with everything else:

```{r js-libs}
list.files(system.file("js", package = "NADIA"), recursive = TRUE)
```

Both are MIT-licensed, and their terms travel with them in those directories.

# Saving a widget to a file

```{r save-widget, message = FALSE}
w <- results_list_widget(res$DEPs_results)
f <- file.path(tempdir(), "results.html")
htmltools::save_html(w, f)

round(file.size(f) / 1024)   # KB
```

`htmltools::save_html()` writes the page and copies its libraries into a `lib/`
directory beside it — so the HTML and that folder travel together. For a single
file that can be emailed, use `htmlwidgets::saveWidget()` with
`selfcontained = TRUE`, which base64-embeds everything:

```{r save-selfcontained, eval = FALSE}
htmlwidgets::saveWidget(w, file.path(tempdir(), "results_standalone.html"),
                        selfcontained = TRUE)
```

The same applies to the Highcharts figures:

```{r save-hc, eval = FALSE}
v <- volcano_highchart_list(res$DEPs_results)
htmlwidgets::saveWidget(v[["B-A"]], file.path(tempdir(), "volcano_B-A.html"),
                        selfcontained = TRUE)
```

# What to keep

The one thing a results file cannot tell you on its own is how it was made, and
the parameters are recorded for exactly that reason:

```{r parameters}
unlist(res$parameters[c("norm_method", "imp_method", "mar_method",
                        "mnar_method", "de_method", "alpha",
                        "logFC_threshold")])
```

Given `vignette("choosing-methods")` — where four combinations of two arguments
moved the count of down-regulated proteins by a factor of fourteen on this same
dataset — this list is not bookkeeping. It is the difference between a result
someone can reproduce and one they cannot.

```{r cleanup, include = FALSE}
unlink(c(out, file.path(tempdir(), "nadia_parquet"), f), recursive = TRUE)
```

```{r session}
sessionInfo()
```
