---
title: "Visualisation: exploring and presenting results"
author: "Sergio Ciordia"
package: NADIA
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Visualisation: exploring and presenting results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  crop = NULL,
  message = FALSE,
  warning = FALSE
)

options(width = 100)

have_heatmap <- all(vapply(
  c("ComplexHeatmap", "tidyHeatmap", "circlize"),
  requireNamespace,
  logical(1),
  quietly = TRUE
))

have_pattern_profiler <- all(vapply(
  c("Mfuzz", "Biobase", "e1071"),
  requireNamespace,
  logical(1),
  quietly = TRUE
))

# highcharter 0.9.5 declares every Highcharts module for every widget. Some of
# those modules contain literal HTML closing tags, which can be exposed as text
# when Pandoc embeds the dependencies in a self-contained BiocStyle vignette.
# The examples below need only the core and highcharts-more modules. Exporting
# remains available in ordinary NADIA use and is documented later in the text.
prepare_vignette_highchart <- function(widget) {
  widget <- highcharter::hc_exporting(widget, enabled = FALSE)

  highcharts_dir <- system.file(
    "htmlwidgets/lib/highcharts",
    package = "highcharter"
  )

  slim_dependency <- htmltools::htmlDependency(
    name = "highcharts",
    version = "9.3.1.9000",
    src = c(file = highcharts_dir),
    script = c("highcharts.js", "highcharts-more.js", "custom/reset.js"),
    all_files = FALSE
  )

  widget$dependencies <- c(widget$dependencies, list(slim_dependency))
  widget
}
```

<style type="text/css">
.visualisation-choice-table {
  width: calc(100% - 4.5em);
  box-sizing: border-box;
  margin: 1em 0 1.5em;
}
.visualisation-choice-table table {
  width: 100% !important;
  max-width: none;
  margin-left: 0 !important;
  margin-right: 0 !important;
  table-layout: fixed;
  font-size: 0.95em;
  line-height: 1.3;
}
.visualisation-choice-table th,
.visualisation-choice-table td {
  padding: 0.5em 0.7em !important;
  vertical-align: top !important;
}
.visualisation-choice-table th:nth-child(1),
.visualisation-choice-table td:nth-child(1) {
  width: 29%;
}
.visualisation-choice-table th:nth-child(2),
.visualisation-choice-table td:nth-child(2) {
  width: 17%;
}
.visualisation-choice-table th:nth-child(3),
.visualisation-choice-table td:nth-child(3) {
  width: 18%;
}
.visualisation-choice-table th:nth-child(4),
.visualisation-choice-table td:nth-child(4) {
  width: 36%;
}
.visualisation-warning {
  width: calc(100% - 4.5em);
  box-sizing: border-box;
  margin: 1.15em 0 1.5em;
  padding: 0.9em 1.1em;
  border: 1px solid #e4c78e;
  border-left: 5px solid #c98718;
  border-radius: 5px;
  background: #fffaf0;
  box-shadow: 0 1px 2px rgba(112, 78, 20, 0.08);
}
.visualisation-warning p {
  margin: 0 0 0.75em;
  padding: 0 !important;
  text-align: justify;
}
.visualisation-warning p:last-child {
  margin-bottom: 0;
}
.visualisation-warning strong:first-child {
  color: #7a5310;
}
.interactive-chart-compact {
  width: 90%;
  margin: 1em auto 1.6em;
}
.interactive-chart-compact .html-widget {
  width: 100% !important;
}
.boxplot-chart .html-widget {
  height: 560px !important;
}
.boxplot-chart {
  margin-bottom: 2.5em;
}
.wide-data-table {
  width: calc(100% - 4.5em);
  box-sizing: border-box;
  margin: 1em 0 1.5em;
}
.wide-data-table table {
  width: 100% !important;
  max-width: none;
  margin-left: 0 !important;
  margin-right: 0 !important;
  table-layout: fixed;
}
.wide-data-table th,
.wide-data-table td {
  padding: 0.5em 0.7em !important;
  vertical-align: top !important;
}
.pca-outline-table th:nth-child(1),
.pca-outline-table td:nth-child(1) {
  width: 18%;
}
.pca-outline-table th:nth-child(2),
.pca-outline-table td:nth-child(2) {
  width: 38%;
}
.pca-outline-table th:nth-child(3),
.pca-outline-table td:nth-child(3) {
  width: 44%;
}
.heatmap-subset-table th:nth-child(1),
.heatmap-subset-table td:nth-child(1) {
  width: 34%;
}
.heatmap-subset-table th:nth-child(2),
.heatmap-subset-table td:nth-child(2),
.heatmap-subset-table th:nth-child(3),
.heatmap-subset-table td:nth-child(3) {
  width: 12%;
}
.heatmap-subset-table th:nth-child(4),
.heatmap-subset-table td:nth-child(4) {
  width: 42%;
}
.palette-reference-table {
  margin: 1em 0 1.5em;
}
.palette-reference-table table {
  margin-left: 0 !important;
  margin-right: auto !important;
}
@media screen and (max-width: 767px) {
  .visualisation-choice-table {
    overflow-x: auto;
  }
  .wide-data-table {
    overflow-x: auto;
  }
  .visualisation-choice-table table {
    min-width: 760px;
  }
  .wide-data-table table {
    min-width: 680px;
  }
  .visualisation-choice-table,
  .visualisation-warning,
  .wide-data-table,
  .interactive-chart-compact {
    width: 100%;
  }
}
</style>

# Introduction

A useful proteomics figure should answer a specific question. A volcano plot
summarises differential abundance, a boxplot compares sample distributions, a
principal component analysis (PCA) reveals the main sources of variation, and a
heatmap displays coordinated protein profiles across individual samples.
Pattern Profiler adds a different view by grouping proteins with similar
relative trajectories across ordered conditions. These views are complementary;
none is a substitute for the others or for the underlying statistical model.

This vignette presents a complete visualisation workflow for NADIA results. It
shows how to choose and prepare each plot, how to interpret the example output,
which customisation options most often matter, and how to export the result.
The interactive functions use **highcharter**, whereas the publication-oriented
heatmap uses **tidyHeatmap** and **ComplexHeatmap**.

# Setup and example data

This section creates one reproducible analysis that will be reused throughout
the vignette. Keeping the data and thresholds constant makes it possible to
compare what each visualisation adds to the interpretation.

## Installation

NADIA can be installed with `BiocManager`. The heatmap and Pattern Profiler
packages are optional dependencies and are installed separately when those
visualisations are required.

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install(c("NADIA", "ComplexHeatmap", "Mfuzz", "Biobase"))
install.packages(c("tidyHeatmap", "circlize", "e1071"))
```

## Processing the example dataset

The `nadia_dia` dataset contains three conditions (`A`, `B`, and `D`) with four
replicates per condition. `process_proteomics()` returns the processed
experiment and the long-format tables expected by the plotting and clustering
functions.

```{r load-data}
library(NADIA)
data(nadia_dia)

res <- process_proteomics(nadia_dia, verbose = FALSE)
```

The sample design should always be checked before a plot is interpreted.

```{r example-design}
sample_design <- unique(
  res$PCA_Input[c("SampleID", "Condition", "Replicate")]
)

design_summary <- as.data.frame(table(sample_design$Condition))
names(design_summary) <- c("Condition", "Samples")

knitr::kable(
  design_summary,
  row.names = FALSE,
  caption = "Number of samples in each condition of the example experiment."
)
```

The Highcharts functions operate on ordinary data frames, including the
long-format output produced by Pattern Profiler. The clustering analysis itself
uses the processed `SummarizedExperiment` together with `DEPs_results` to select
and aggregate protein profiles. Compatible plotting tables can therefore be
imported from another workflow, whereas a new Pattern Profiler run also needs
the corresponding processed experiment.

```{r visualisation-inputs}
input_summary <- data.frame(
  Object = c("DEPs_results", "BoxPlot_Input", "PCA_Input", "se_proc"),
  Rows = c(
    nrow(res$DEPs_results),
    nrow(res$BoxPlot_Input),
    nrow(res$PCA_Input),
    nrow(res$se_proc)
  ),
  Columns = c(
    ncol(res$DEPs_results),
    ncol(res$BoxPlot_Input),
    ncol(res$PCA_Input),
    ncol(res$se_proc)
  ),
  Used_by = c(
    "Volcano plots and Pattern Profiler",
    "Boxplots",
    "PCA and heatmaps",
    "Pattern Profiler"
  )
)

knitr::kable(
  input_summary,
  row.names = FALSE,
  caption = "Objects returned by process_proteomics() for visualisation."
)
```

# Choosing the appropriate visualisation

Each plot starts from a different analytical question. The following table is a
compact guide to choosing a view before changing colours or other presentation
details.

::: {.visualisation-choice-table}

| Question | Recommended view | Main unit shown | Important limitation |
|:--|:--|:--|:--|
| Which proteins differ between two conditions? | Volcano plot | One protein per comparison | The plot displays model results; it does not refit the model |
| Are sample-level intensity distributions comparable? | Boxplot | One distribution per sample | Similar distributions do not prove that unwanted variation has been removed |
| What are the dominant directions of sample variation? | PCA | One point per sample | Separation depends on the proteins and scaling used |
| Which proteins share a relative abundance pattern? | Heatmap | One row per protein | Row-scaled colours cannot compare absolute abundance between proteins |
| Which proteins follow similar trajectories across ordered conditions? | Pattern Profiler | One standardised profile per protein | Clusters depend on condition order, the chosen number of clusters, and membership threshold |

:::

Several Highcharts functions return **named lists**: volcano plots are named by
comparison, boxplots by assay, PCA plots by subset, and Pattern Profiler profile
plots by cluster. Select one element to display a chart; printing the list
itself is not useful. The Pattern Profiler centroid function instead returns one
chart containing all selected clusters.

```{r named-lists, eval=FALSE}
volcanoes <- volcano_highchart_list(res$DEPs_results)
names(volcanoes)
volcanoes[["B-A"]]

boxplots <- boxplot_highchart_list(res$BoxPlot_Input)
names(boxplots)
boxplots[["Impseqrob_min"]]

pcas <- pca_highchart_list(res$PCA_Input, modes = c("all", "any"))
names(pcas)
pcas[["all"]]
```

## Interactive and static output

Interactive charts are particularly useful during exploration: points can be
identified from tooltips, axes can be zoomed, and groups can be hidden from the
legend. A static heatmap is preferable when dimensions, typography, and vector
output must remain fixed for a manuscript.

One volcano plot, one boxplot, one PCA plot, and two compact Pattern Profiler
charts are rendered below. Additional variants are shown as code without
execution to keep the installed vignette at a reasonable size. The download
module is omitted from these five embedded widgets; charts created in a regular
R session retain their export menu.

## Highcharts licence

::: {.visualisation-warning}
**Important licensing note.** NADIA uses the MIT-licensed R package
**highcharter** as an interface to the Highcharts JavaScript library.
Highcharts is distributed under separate terms, and the licence of highcharter
or NADIA does not grant a Highcharts licence.

At the time this vignette was revised, Highcharts permitted qualifying personal
and educational use under its non-commercial terms, while organisational,
production, and commercial uses could require a commercial licence. Users
should check the current [Highcharts licence](https://shop.highcharts.com/license-eula)
for their intended use. The static heatmap workflow does not use Highcharts.
:::

# Volcano plots

A volcano plot combines effect size and statistical evidence for one comparison.
It is most useful for identifying proteins that are both sufficiently different
in abundance and supported by the differential-abundance model.

## Summarising the selected comparison

Before drawing the figure, it is helpful to count the proteins assigned to each
category. The table below applies the same thresholds as the plot: adjusted
*p*-value below 0.05 and absolute log2 fold change of at least 1.

```{r volcano-summary}
de_ba <- res$DEPs_results[
  res$DEPs_results$Comparison == "B-A",
  ,
  drop = FALSE
]

volcano_class <- ifelse(
  !is.na(de_ba$adj.P.Val) & de_ba$adj.P.Val < 0.05 & de_ba$logFC >= 1,
  "Up",
  ifelse(
    !is.na(de_ba$adj.P.Val) & de_ba$adj.P.Val < 0.05 & de_ba$logFC <= -1,
    "Down",
    "Not significant"
  )
)

volcano_counts <- as.data.frame(table(
  factor(volcano_class, levels = c("Up", "Down", "Not significant"))
))
names(volcano_counts) <- c("Classification", "Proteins")

n_up <- volcano_counts$Proteins[volcano_counts$Classification == "Up"]
n_down <- volcano_counts$Proteins[volcano_counts$Classification == "Down"]
n_not_significant <- volcano_counts$Proteins[
  volcano_counts$Classification == "Not significant"
]

knitr::kable(
  volcano_counts,
  row.names = FALSE,
  caption = "Protein classification for B-A at adjusted p < 0.05 and |log2FC| >= 1."
)
```

Under these criteria, the example contains `r n_up` up-regulated proteins,
`r n_down` down-regulated proteins, and `r n_not_significant` proteins that do
not pass both thresholds. This numerical summary prevents the apparent density
of points from being mistaken for a count.

## Creating and reading the plot

`volcano_highchart_list()` creates one chart for every requested comparison.
Here, the six most statistically significant proteins that also pass both
thresholds are labelled automatically.

::: {.interactive-chart-compact}

```{r volcano-plot}
volcanoes <- volcano_highchart_list(
  res$DEPs_results,
  comparisons = "B-A",
  alpha = 0.05,
  lfc_thr = 1,
  p_col = "adj.P.Val",
  show_top_genes = 6,
  title = "Differential abundance: {comparison}"
)

prepare_vignette_highchart(volcanoes[["B-A"]])
```

:::

The x-axis is the log2 fold change. Positive values indicate greater abundance
in `B` relative to `A`, whereas negative values indicate greater abundance in
`A`. The y-axis is `-log10(adj.P.Val)`, so stronger statistical evidence appears
higher in the chart. A protein is coloured as up- or down-regulated only when it
passes both the horizontal significance threshold and the vertical effect-size
threshold.

The `lfc_thr` argument controls the guides and classification **in the figure**.
It does not rerun differential-abundance analysis or alter `res$DEPs_results`.
For the plot and the result table to communicate the same decision rule, use the
same `alpha` and fold-change threshold in both stages and record those values.

By default, `p_col = "adj.P.Val"` controls both the y-axis and classification.
Raw `P.Value` can be displayed for diagnostic work, but adjusted values are the
appropriate default when many proteins are tested simultaneously.

## Labelling selected proteins and multiple comparisons

Automatic labels are convenient for exploration, while a manually selected set
is usually clearer in a report. `highlight_genes` matches the `Gene.Names`
column and may highlight a protein whether or not it passes the thresholds.

```{r volcano-options, eval=FALSE}
# Label proteins selected for their biological relevance.
volcano_highchart_list(
  res$DEPs_results,
  comparisons = "B-A",
  alpha = 0.05,
  lfc_thr = 1,
  highlight_genes = c("lamB", "yciU", "gltA")
)[["B-A"]]

# Build several charts with identical settings.
volcanoes_all <- volcano_highchart_list(
  res$DEPs_results,
  comparisons = c("B-A", "D-A", "D-B"),
  alpha = 0.05,
  lfc_thr = 1,
  title = "Differential abundance: {comparison}"
)

names(volcanoes_all)
```

# Boxplots of sample distributions

Sample-level boxplots provide a compact quality-control view of the intensity
distribution. They are useful before and after processing because shifts in
location, spread, or shape can reveal samples that deserve closer inspection.

## Comparing assays numerically

The example output contains the initial log2 intensities and the final processed
assay. The following summary calculates the median for every sample and then
reports the range of those sample medians within each assay.

```{r boxplot-summary}
sample_medians <- aggregate(
  Intensity ~ Assay + Column + Condition,
  data = res$BoxPlot_Input,
  FUN = median,
  na.rm = TRUE
)

median_ranges <- do.call(
  rbind,
  lapply(split(sample_medians$Intensity, sample_medians$Assay), function(x) {
    data.frame(
      Minimum = min(x),
      Maximum = max(x),
      Range = max(x) - min(x)
    )
  })
)

median_ranges$Assay <- rownames(median_ranges)
rownames(median_ranges) <- NULL
median_ranges <- median_ranges[c("Assay", "Minimum", "Maximum", "Range")]
median_ranges[-1] <- lapply(median_ranges[-1], round, digits = 3)

knitr::kable(
  median_ranges,
  row.names = FALSE,
  caption = "Range of sample medians before and after processing."
)
```

In this dataset, the range of sample medians decreases after processing. This is
consistent with improved alignment of the sample distributions, but it is not
proof that every technical effect has been removed. Distributional similarity
should be considered together with the experimental design, PCA, and the
normalisation diagnostics described in `vignette("choosing-methods")`.

## Creating and reading the boxplot

The chart below shows the final assay. Each box spans the first to the third
quartile, the central line is the median, and the whiskers extend to the most
extreme values within 1.5 interquartile ranges. Hovering over a box reports these
statistics and the number of finite protein intensities in that sample.

::: {.interactive-chart-compact .boxplot-chart}

```{r boxplot-plot}
set.seed(42)

boxplots <- boxplot_highchart_list(
  res$BoxPlot_Input,
  assays = "Impseqrob_min",
  color_by = "Condition",
  group_order = c("A", "B", "D"),
  horizontal = FALSE,
  show_outliers = TRUE,
  title = "Processed intensity distributions",
  subtitle = "Assay: {assay}",
  height = 560
)

prepare_vignette_highchart(boxplots[["Impseqrob_min"]])
```

:::

The medians are closely aligned, while each sample retains a broad intensity
range. Alignment is expected after the processing used in this example;
complete collapse of the boxes, however, would be suspicious because biological
and sampling variation should remain.

The points beyond the whiskers are displayed with `show_outliers = TRUE`. They
highlight the tails of each sample distribution but are not automatically
erroneous measurements. In much larger experiments, set
`show_outliers = FALSE` when the additional points obscure the boxes or make
the interactive widget unnecessarily heavy.

```{r boxplot-options, eval=FALSE}
# Display both stages as separate list elements.
boxplots_both <- boxplot_highchart_list(
  res$BoxPlot_Input,
  assays = c("log2", "Impseqrob_min"),
  group_order = c("A", "B", "D"),
  show_outliers = TRUE,
  outlier_size = 2,
  box_width = 16
)

boxplots_both[["log2"]]
boxplots_both[["Impseqrob_min"]]
```

# Principal component analysis

PCA reduces thousands of protein measurements to a small number of directions
that capture the largest variation among samples. It is a diagnostic view of
similarity, separation, and possible outliers; it is not a significance test.

## Comparing protein subsets

The protein subset is part of the PCA question. NADIA supports all proteins,
proteins significant in any comparison, or proteins significant in one named
comparison. The table below compares the first two choices using the same
centering and scaling settings.

```{r pca-summary}
pca_scores_all <- build_pca_scores(
  res$PCA_Input,
  mode = "all",
  center = TRUE,
  scale. = TRUE
)

pca_scores_any <- build_pca_scores(
  res$PCA_Input,
  mode = "any",
  center = TRUE,
  scale. = TRUE
)

pca_summary <- data.frame(
  Subset = c("All proteins", "Significant in any comparison"),
  Proteins = c(
    length(unique(res$PCA_Input$FeatureID)),
    length(unique(res$PCA_Input$FeatureID[res$PCA_Input$sig_any %in% TRUE]))
  ),
  PC1 = c(unique(pca_scores_all$PC1_Perc), unique(pca_scores_any$PC1_Perc)),
  PC2 = c(unique(pca_scores_all$PC2_Perc), unique(pca_scores_any$PC2_Perc))
)

names(pca_summary)[3:4] <- c("PC1 variance (%)", "PC2 variance (%)")

knitr::kable(
  pca_summary,
  row.names = FALSE,
  digits = 2,
  caption = "Protein subsets and variance represented by the first two principal components."
)
```

With all proteins, PC1 explains `r unique(pca_scores_all$PC1_Perc)`% of the
variance. Restricting the analysis to proteins already selected by differential
abundance raises this to `r unique(pca_scores_any$PC1_Perc)`%. The stronger
separation in the selected subset is expected because the same data were used to
select proteins that differ between conditions. It should not be presented as
independent validation of those differences.

## Creating and reading the PCA plot

The baseline plot should generally use all proteins. Sample labels are enabled
here because the experiment contains only 12 samples. The 95% confidence
ellipses provide a normal-theory summary of the location and covariance of each
condition.

::: {.interactive-chart-compact}

```{r pca-plot}
condition_colours <- c(
  A = "#0072B2",
  B = "#D55E00",
  D = "#009E73"
)

pcas <- pca_highchart_list(
  res$PCA_Input,
  modes = "all",
  color_by = "Condition",
  group_order = c("A", "B", "D"),
  palette = condition_colours,
  addEllipses = TRUE,
  ellipse_type = "confidence",
  ellipse_level = 0.95,
  show_labels = TRUE,
  point_size = 5,
  center = TRUE,
  scale. = TRUE
)

prepare_vignette_highchart(pcas[["all"]])
```

:::

Points close together have similar multivariate protein profiles under the
chosen preprocessing and scaling. Separation along PC1 or PC2 indicates a major
source of variation, but the plot alone does not establish whether that source
is biological, technical, or confounded with the experimental design. An
isolated replicate should therefore be traced back to its sample-level quality
metrics rather than removed from the plot by inspection alone.

The ellipses are descriptive overlays, not significance tests or decision
boundaries. With only four replicates per condition, their covariance estimates
are uncertain; they should be used for visual orientation rather than as proof
of group separation.

`scale. = TRUE` gives each protein unit variance before PCA and prevents
high-variance proteins from dominating solely because of their scale. With
`scale. = FALSE`, proteins with greater between-sample variance contribute more
strongly. The choice should be recorded because it can materially change the
projection.

## Subset modes, ellipses, and sample filtering

`pca_highchart_list()` uses a comparison name directly as a mode. For example,
`modes = "B-A"` selects proteins significant in that comparison. When
`filter_samples_to_comparison = TRUE`, samples from condition `D` are also
excluded; when it is `FALSE`, all samples are projected using the proteins
selected from `B-A`.

```{r pca-options, eval=FALSE}
pca_ba <- pca_highchart_list(
  res$PCA_Input,
  modes = "B-A",
  alpha = 0.05,
  filter_samples_to_comparison = TRUE,
  group_order = c("A", "B"),
  palette = condition_colours[c("A", "B")],
  addEllipses = FALSE,
  show_labels = TRUE
)

pca_ba[["B-A"]]
```

Two group outlines are available:

::: {.wide-data-table .pca-outline-table}

| `ellipse_type` | Geometry | Appropriate interpretation |
|:--|:--|:--|
| `"convex"` | Smallest convex polygon containing the observed samples | Describes the observed extent of a group |
| `"confidence"` | Covariance ellipse scaled by a chi-squared quantile | Normal-theory summary whose shape is unstable in small groups |

:::

At least three samples are needed to draw either outline. With four replicates
per condition, as in this example, the covariance estimate behind a confidence
ellipse is especially uncertain. The outline should support visual orientation,
not serve as evidence of significant group separation.

# Heatmaps

A heatmap shows the relative abundance profile of many proteins across samples.
It is particularly effective for assessing whether selected proteins form
coherent patterns and whether replicates display similar profiles. Unlike
Pattern Profiler, it preserves the individual sample columns and does not assign
proteins to profile clusters.

## Selecting proteins for the heatmap

Showing every quantified protein is possible but rarely legible. NADIA provides
three filtering modes, and `feature_ids` can be used when a focused, explicitly
defined panel is needed.

::: {.wide-data-table .heatmap-subset-table}

```{r heatmap-subsets}
n_all <- length(unique(res$PCA_Input$FeatureID))
n_any <- length(unique(
  res$PCA_Input$FeatureID[res$PCA_Input$sig_any %in% TRUE]
))
n_ba <- length(unique(
  res$PCA_Input$FeatureID[
    !is.na(res$PCA_Input[["adjP_B-A"]]) &
      res$PCA_Input[["adjP_B-A"]] < 0.05
  ]
))

heatmap_subsets <- data.frame(
  Request = c(
    'mode = "all"',
    'mode = "any"',
    'mode = "target", comparison = "B-A"'
  ),
  Proteins = c(n_all, n_any, n_ba),
  Samples = c(12, 12, 8),
  Interpretation = c(
    "All quantified proteins",
    "Significant in at least one comparison",
    "Significant in B-A; only A and B samples"
  )
)

knitr::kable(
  heatmap_subsets,
  row.names = FALSE,
  caption = "Available protein-filtering modes for the example heatmap."
)
```

:::

Unlike a comparison-specific PCA, `mode = "target"` in
`proteomics_heatmap()` automatically restricts the samples to the conditions in
the comparison. The `B-A` heatmap therefore contains eight samples rather than
all twelve.

## Creating and reading a focused heatmap

The complete `B-A` set contains `r n_ba` proteins, which is too dense for row
labels in a vignette. The following example displays the 30 proteins with the
smallest adjusted *p*-values among those passing the 0.05 threshold.

```{r select-heatmap-proteins}
de_ba_ordered <- de_ba[
  !is.na(de_ba$adj.P.Val) & de_ba$adj.P.Val < 0.05,
  ,
  drop = FALSE
]
de_ba_ordered <- de_ba_ordered[order(de_ba_ordered$adj.P.Val), , drop = FALSE]
ba_top_ids <- head(unique(de_ba_ordered$Protein.IDs), 30)
```

```{r heatmap-plot, eval=have_heatmap, fig.width=8, fig.height=7, fig.align="center", fig.cap="Row-scaled abundance profiles for the 30 proteins with the smallest adjusted p-values in the B-A comparison.", fig.alt="Row-scaled heatmap of the 30 most statistically significant proteins in the B-A comparison"}
proteomics_heatmap(
  res$PCA_Input,
  mode = "target",
  comparison = "B-A",
  feature_ids = ba_top_ids,
  scale_data = "row",
  sample_order = "condition",
  condition_order = c("A", "B"),
  cluster_rows = TRUE,
  show_row_names = TRUE,
  palette_value = "brewer:RdBu",
  reverse_palette = TRUE,
  heatmap_title = "Top B-A protein profiles"
)
```

Each row is transformed to a z-score because `scale_data = "row"`. Warm and
cool colours therefore represent values above and below that protein's own mean;
they do **not** indicate that one protein is more abundant than another. Row
clustering groups proteins with similar relative profiles, while the explicit
condition order keeps the sample columns arranged as `A` followed by `B`.

Use `scale_data = "none"` when absolute log2 intensity differences between
proteins are the subject of the figure. In that case, a small number of
high-abundance proteins may dominate the colour range. Column scaling answers a
different question and is rarely the default for protein-profile heatmaps.

## Annotations and repeated heatmaps

Categorical protein annotations can be supplied as a data frame containing a
`FeatureID` column. Rows may be split by one annotation and ordered by another.
The following complete example is not rendered because the preceding figure
already shows the same proteins.

```{r heatmap-annotations, eval=FALSE}
direction <- ifelse(
  de_ba$logFC[match(ba_top_ids, de_ba$Protein.IDs)] > 0,
  "Higher in B",
  "Higher in A"
)

protein_annotation <- data.frame(
  FeatureID = ba_top_ids,
  Direction = direction
)

proteomics_heatmap(
  res$PCA_Input,
  mode = "target",
  comparison = "B-A",
  feature_ids = ba_top_ids,
  row_annotation = protein_annotation,
  row_annotation_cols = "Direction",
  row_annotation_palette = list(
    Direction = c("Higher in B" = "#D55E00", "Higher in A" = "#0072B2")
  ),
  split_rows_by = "Direction",
  scale_data = "row",
  sample_order = "condition",
  condition_order = c("A", "B")
)

# Build a named list containing several heatmaps.
heatmaps <- proteomics_heatmap_list(
  res$PCA_Input,
  modes = c("any", "B-A", "D-A"),
  scale_data = "row",
  sample_order = "condition",
  condition_order = c("A", "B", "D")
)

names(heatmaps)
```

# Protein pattern clusters (optional)

Pattern Profiler groups proteins that follow similar relative trajectories
across ordered conditions. It complements a heatmap by turning recurring
profile shapes into explicit clusters and by assigning each protein a fuzzy
membership between 0 and 1. This approach is most informative with at least
three ordered conditions, such as time points, doses, or disease stages.

This section requires the optional **Mfuzz**, **Biobase**, and **e1071**
packages. It focuses on visualising an existing clustering result; see
`vignette("pattern-profiler")` for cluster-number selection, membership
diagnostics, and data storage.

## Building and summarising the example clusters

For a concise and reproducible visual example, the analysis below fixes the
number of clusters at two. In a new study, the candidate solutions should first
be compared with `auto_select_c = TRUE` and interpreted using the diagnostics
described in the dedicated Pattern Profiler vignette.

```{r pattern-profiler-run, eval=have_pattern_profiler, results="hide"}
pp_vis <- pattern_profiler_analysis(
  se_proc = res$se_proc,
  DEPs_results = res$DEPs_results,
  assay_name = "Impseqrob_min",
  filter_mode = "any",
  alpha = 0.05,
  condition_order = c("A", "B", "D"),
  aggregate = "median",
  auto_select_c = FALSE,
  c = 2,
  min_membership = 0.25,
  seed = 42,
  verbose = FALSE
)
```

A stricter membership threshold is useful for displaying the clearest examples
of each pattern. The table reports how many proteins remain at 0.70 and the
median standardised profile represented by each cluster.

```{r pattern-profiler-summary, eval=have_pattern_profiler}
display_membership <- 0.70
pp_display <- pp_vis$long_output[
  pp_vis$long_output$Membership >= display_membership,
  ,
  drop = FALSE
]

pattern_summary <- do.call(
  rbind,
  lapply(split(pp_display, pp_display$Cluster), function(x) {
    data.frame(
      Cluster = paste("Cluster", unique(x$Cluster)),
      Proteins = length(unique(x$FeatureID)),
      `Median membership` = median(x$Membership),
      A = median(x$A),
      B = median(x$B),
      D = median(x$D),
      check.names = FALSE
    )
  })
)

pattern_summary[c("Median membership", "A", "B", "D")] <- lapply(
  pattern_summary[c("Median membership", "A", "B", "D")],
  round,
  digits = 3
)

knitr::kable(
  pattern_summary,
  row.names = FALSE,
  caption = paste0(
    "High-membership proteins and median z-score profiles at membership >= ",
    display_membership, "."
  )
)
```

The two clusters describe opposing relative trajectories. One pattern remains
comparatively high in `A` and `B` before decreasing in `D`, whereas the other
rises across the ordered conditions. Because the condition values are
within-protein z-scores, they describe profile shape rather than absolute
protein abundance.

## Comparing cluster centroids

`cluster_centroids_highchart()` places the representative trajectory of every
cluster in one chart. The median is used here because it is less sensitive than
the mean to proteins near a cluster boundary.

::: {.interactive-chart-compact}

```{r pattern-centroids, eval=have_pattern_profiler}
centroid_plot <- cluster_centroids_highchart(
  pp_vis$long_output,
  conditions = pp_vis$conditions,
  min_membership = display_membership,
  centroid_summary = "median",
  title = "Pattern Profiler cluster centroids",
  height = 440
)

prepare_vignette_highchart(centroid_plot)
```

:::

The separation between the centroid lines shows when the dominant patterns
diverge. Hovering over a marker reports the condition and median z-score. The
centroids summarise many proteins and should therefore be read together with
the cluster sizes and memberships in the preceding table.

## Inspecting representative protein profiles

A cluster-level chart reveals whether individual profiles are concentrated
around their centroid or whether the summary hides substantial heterogeneity.
To keep the interactive vignette responsive, the next chart displays 100
proteins sampled evenly across the membership range retained for cluster 1.

::: {.interactive-chart-compact}

```{r pattern-profiles, eval=have_pattern_profiler}
cluster_one <- pp_display[pp_display$Cluster == 1, , drop = FALSE]
cluster_one <- cluster_one[order(cluster_one$Membership, decreasing = TRUE), ]
representative_positions <- unique(round(seq(
  from = 1,
  to = nrow(cluster_one),
  length.out = min(100, nrow(cluster_one))
)))
representative_ids <- cluster_one$FeatureID[representative_positions]
representative_profiles <- cluster_one[
  cluster_one$FeatureID %in% representative_ids,
  ,
  drop = FALSE
]

profile_plot <- cluster_profile_highchart(
  representative_profiles,
  cluster = 1,
  conditions = pp_vis$conditions,
  min_membership = display_membership,
  centroid_summary = "median",
  line_opacity = 0.35,
  title = "Cluster 1: 100 representative protein profiles",
  height = 440
)

prepare_vignette_highchart(profile_plot)
```

:::

The thin lines are individual standardised protein profiles and the thicker
line is their median. A narrow band around the centroid indicates a coherent
pattern; marked deviations identify proteins whose assignment deserves closer
inspection. Showing only high-membership representatives makes the dominant
shape clear, but it does not describe the full uncertainty of the fuzzy
clustering.

The complete interactive list can be generated when detailed exploration is
more important than vignette size. With hundreds of series, these charts are
best inspected in an R session rather than embedded in the installed vignette.

```{r pattern-profiles-all, eval=FALSE}
all_profile_plots <- cluster_profile_highchart_list(
  pp_vis$long_output,
  conditions = pp_vis$conditions,
  min_membership = 0.70,
  centroid_summary = "median"
)

all_profile_plots[["Cluster_1"]]
```

# Colours and consistent presentation

Colour should encode the same meaning across related figures. Reusing a fixed
condition palette makes samples easier to follow and prevents colours from being
reassigned when a subset omits one condition. Pattern Profiler cluster numbers
are arbitrary labels, but a fixed cluster palette should still be reused between
the centroid and individual-profile charts from the same analysis.

```{r colour-examples, eval=FALSE}
condition_colours <- c(
  A = "#0072B2",
  B = "#D55E00",
  D = "#009E73"
)

pca_highchart_list(
  res$PCA_Input,
  modes = "all",
  group_order = c("A", "B", "D"),
  palette = condition_colours
)[["all"]]

boxplot_highchart_list(
  res$BoxPlot_Input,
  assays = "Impseqrob_min",
  group_order = c("A", "B", "D"),
  palette = unname(condition_colours)
)[["Impseqrob_min"]]
```

Palette names are resolved through either `paletteer` or `RColorBrewer`,
depending on their format. An explicit vector is the safest choice for a series
of publication figures because the mapping is visible in the script.

::: {.palette-reference-table}

| Function | Palette argument | Example |
|:--|:--|:--|
| `volcano_highchart_list()` | `palette` or `colors = list(up, down, ns)` | `palette = "ggsci::default_jco"` |
| `boxplot_highchart_list()` | Paletteer name, `"brewer:Name"`, or colour vector | `palette = "brewer:Dark2"` |
| `pca_highchart_list()` | Paletteer name, `"brewer:Name"`, or preferably a named vector | `palette = condition_colours` |
| Pattern Profiler Highcharts | Paletteer name, `"brewer:Name"`, or colour vector | `palette = c("#E63946", "#457B9D")` |
| `proteomics_heatmap()` | Separate value and annotation palettes | `palette_value = "brewer:RdBu"` |

:::

For volcano plots, the explicit colour-list names are lower case:

```{r volcano-colours, eval=FALSE}
volcano_highchart_list(
  res$DEPs_results,
  comparisons = "B-A",
  colors = list(
    up = "#D55E00",
    down = "#0072B2",
    ns = "#BDBDBD"
  )
)[["B-A"]]
```

# Exporting figures and data

Export should preserve both the visual result and the data used to create it.
Interactive widgets can be shared as HTML, while heatmaps can be written as
raster or vector graphics together with their displayed matrix.

## Interactive charts

Every NADIA Highchart includes a download menu for PNG, SVG, and PDF output.
The menu is disabled only in the five embedded examples above to keep this
self-contained vignette compact and robust. For a fully interactive copy, save
the original chart or selected list element as an HTML widget.

```{r export-interactive, eval=FALSE}
htmlwidgets::saveWidget(
  volcanoes[["B-A"]],
  file = file.path(tempdir(), "volcano_B-A.html"),
  selfcontained = TRUE
)

htmlwidgets::saveWidget(
  centroid_plot,
  file = file.path(tempdir(), "pattern_centroids.html"),
  selfcontained = TRUE
)
```

A self-contained widget is convenient for sharing, but it remains subject to
the Highcharts licence described above. The file also contains the plotted data,
so it should not be used to distribute confidential results without review.

## Static heatmaps

For `proteomics_heatmap()`, `export_path` writes the displayed matrix and
metadata to a TSV file, whereas `export_file` writes the figure. The plot format
is inferred from the extension (`.png`, `.svg`, or `.pdf`).

```{r export-heatmap, eval=FALSE}
proteomics_heatmap(
  res$PCA_Input,
  mode = "target",
  comparison = "B-A",
  feature_ids = ba_top_ids,
  scale_data = "row",
  sample_order = "condition",
  condition_order = c("A", "B"),
  export_path = file.path(tempdir(), "heatmap_B-A_data.tsv"),
  export_file = file.path(tempdir(), "heatmap_B-A.png"),
  plot_width = 8,
  plot_height = 7,
  export_dpi = 300
)
```

When `proteomics_heatmap_list()` is used, the mode is appended to each export
name. For example, a base name of `heatmap.png` produces files such as
`heatmap_any.png` and `heatmap_B-A.png`.

The companion vignette `vignette("results-and-export")` describes additional
ways to save complete NADIA analyses, result tables, and interactive objects.

# Practical recommendations

The most defensible visualisation workflow begins with the analytical question
and ends with a reproducible export. The following practices help keep figures
clear and their interpretation proportionate to the evidence:

- inspect the sample design and the number of observations before plotting;
- state the adjusted *p*-value and fold-change thresholds used in volcano plots;
- compare boxplots before and after processing, without treating similar boxes
  as proof that all technical variation has disappeared;
- begin PCA with all proteins, then label analyses of selected proteins as
  supervised or selection-dependent views;
- use row-scaled heatmaps for relative profiles and unscaled heatmaps for
  absolute log2 abundance;
- use Pattern Profiler when conditions have a meaningful order, evaluate the
  number of clusters and membership threshold, and interpret its z-scores as
  relative trajectories rather than absolute abundance;
- reuse explicit colours and sample orders across related figures;
- avoid dense labels when tooltips or a focused protein panel communicate the
  result more clearly; and
- save the plotted data and the analysis parameters together with the figure.

No single figure validates an analysis. Agreement among the experimental
design, quality-control summaries, model results, and several complementary
visualisations provides a much stronger basis for interpretation.

# Session information

The package versions used to build this vignette are recorded below to support
reproducibility.

```{r session-information}
sessionInfo()
```
