---
title: "Visualization: tuning the figures"
author: "Sergio Ciordia"
package: NADIA
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Visualization: tuning the figures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", crop = NULL)
```

# Introduction

The visualisation modules are independent of the pipeline: they take data frames,
not the objects that produced them, so any of them can be pointed at results
computed elsewhere. This vignette covers the arguments that matter and the two or
three places where a default deserves a second thought.

The split is deliberate. **Highcharts** for the figures you explore — volcano,
boxplot, PCA — where hovering a point to read its identity is the whole value.
**ggplot2 and ComplexHeatmap** for the figures you publish, where a static file
at a fixed size is what a journal wants.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("NADIA")
```

```{r library, message = FALSE}
library(NADIA)
data(nadia_dia)
```

```{r run, message = FALSE}
res <- process_proteomics(nadia_dia, verbose = FALSE)
```

**One figure per module is rendered live** below — a volcano, a boxplot and a
PCA. Hover a point to read its identity, drag to zoom, click a legend entry to
hide a series. The variations on each are shown as code without being run: the
Highcharts library costs about 2 MB once, and each rendered figure carries its
own data, which for 2,000 proteins is another 500 KB or so. Three is what fits
comfortably; the rest would take this document past the size a package should
ship.

The `*_highchart_list()` functions return a **named list**, one element per
comparison or assay, so the examples subset it — `[["B-A"]]`, `[["any"]]` — to
render a single figure. Printing the list itself renders nothing useful.

## A note on Highcharts licensing

NADIA uses **highcharter** (MIT) to build its interactive figures. highcharter is
an R interface to the **Highcharts** JavaScript library, which is distributed
under its own separate terms: free for personal and non-commercial use, but
requiring a paid licence for commercial and governmental use.

NADIA does not bundle Highcharts — it is shipped by highcharter — and NADIA's own
GPL-3 licence neither grants nor implies a licence for it. Users are responsible
for ensuring that their use of Highcharts complies with the applicable
[Highcharts licence](https://www.highcharts.com/license). Highsoft offers a
discount to highcharter users.

This is worth reading before you build a figure for a grant report or a
commercial deliverable. The static figures in the second half of this vignette —
ggplot2 and ComplexHeatmap — carry no such restriction, and every analysis in
NADIA can be completed without producing a single interactive figure.

# The three inputs

Each interactive module consumes one of the three tables the pipeline returns,
and none of them needs the `SummarizedExperiment`:

```{r inputs}
str(res$DEPs_results,  max.level = 1, give.attr = FALSE, list.len = 4)
str(res$BoxPlot_Input, max.level = 1, give.attr = FALSE, list.len = 4)
str(res$PCA_Input,     max.level = 1, give.attr = FALSE, list.len = 4)
```

# Volcano plots

```{r volcano}
v <- volcano_highchart_list(res$DEPs_results,
                            alpha   = 0.05,
                            lfc_thr = 1,
                            palette = "RColorBrewer::Set1")
v[["B-A"]]
```

`lfc_thr` here is a **drawing** threshold — where the vertical guides go and how
points are coloured. It is independent of the `logFC_threshold` used when the
model was fitted, which is what decided the `Change` column. Setting it to 1 in
the plot does not re-test anything, and a point drawn outside the guides is not
thereby significant.

Two arguments label points. `show_top_genes` takes the *n* most significant:

```{r volcano-top, eval = FALSE}
volcano_highchart_list(res$DEPs_results, show_top_genes = 10)[["B-A"]]
```

`highlight_genes` takes a list you choose, which is usually what a figure for a
paper needs — the proteins the paper is about, whether or not they top the table:

```{r volcano-highlight, eval = FALSE}
volcano_highchart_list(res$DEPs_results,
                       highlight_genes = c("ALB", "NUDT4B", "EPPK1"))[["B-A"]]
```

`p_col` switches the y axis between `adj.P.Val` (the default, and the honest
choice) and `P.Value`.

# Boxplots

The input holds one row per protein, sample and assay, so a single call can show
the effect of a processing step:

```{r boxplot-assays}
unique(res$BoxPlot_Input$Assay)
```

```{r boxplot}
boxplot_highchart_list(res$BoxPlot_Input,
                       assays     = c("log2", "Impseqrob_min"),
                       color_by   = "Condition",
                       horizontal = TRUE)[["Impseqrob_min"]]
```

Comparing `log2` against the processed assay is the standard check that
normalisation did what it claims: the medians should line up afterwards without
the distributions collapsing onto each other.

`group_order` fixes the order of the boxes, which otherwise follows the factor
levels; `show_outliers = FALSE` drops the individual outlier points, worth doing
when there are thousands of them and the browser starts to labour.

# PCA

```{r pca}
pca_highchart_list(res$PCA_Input,
                   modes        = c("all", "any"),
                   ellipse_type = "confidence",
                   addEllipses  = TRUE)[["any"]]
```

Three arguments deserve attention.

**`modes`** — note the plural — selects which proteins enter the projection.
`"all"` uses every protein; `"any"` restricts to those significant in at least
one comparison; and **the name of a comparison**, such as `"B-A"`, restricts to
the proteins significant in that one. Note that the third form is a comparison
name, not a keyword: there is no `"specific"` mode here, unlike the heatmap and
the Pattern Profiler, where `filter_mode = "specific"` is spelled that way.

The first two answer different questions: `"all"` asks whether the samples
separate at all, `"any"` asks how well they separate *given* the proteins the
analysis already selected. The second will always look better, and that is not
evidence of anything.

**`ellipse_type`** is the one to think about:

| Value | What it draws | What it means |
|---|---|---|
| `"convex"` (default) | the convex hull of the group | exactly where the samples are, no more |
| `"confidence"` | a `t`-based confidence ellipse | an inference about the population |

A hull is a description and cannot be wrong. A confidence ellipse is a claim, and
with four replicates per group it is a claim resting on three degrees of freedom
— it will be large, and its size should not be read as a measurement.

**`filter_samples_to_comparison`** drops samples not involved in a comparison. For
`"B-A"` it plots only the B and A samples, which is often what you meant when the
experiment has more than two groups.

```{r pca-filtered, eval = FALSE}
pca_highchart_list(res$PCA_Input, modes = "B-A",
                   filter_samples_to_comparison = TRUE)[["B-A"]]
```

`scale. = TRUE` (the default) gives every protein equal weight. Turn it off and
the projection is dominated by the most abundant proteins, since their variance
in log2 space is largest in absolute terms.

# Heatmaps

The heatmap is static, built on ComplexHeatmap through tidyHeatmap, and takes the
same long table as the PCA:

```{r heatmap, fig.width = 7, fig.height = 5.5, fig.alt = "Heatmap of proteins significant in B-A"}
proteomics_heatmap(
    res$PCA_Input,
    mode            = "target",
    comparison      = "B-A",
    scale_data      = "row",
    sample_order    = "condition",
    condition_order = c("A", "B", "D"),
    cluster_rows    = TRUE,
    show_row_names  = FALSE,
    heatmap_title   = "Significant in B-A")
```

`scale_data = "row"` z-scores each protein, which is almost always what you want:
without it the picture is a map of abundance, and the few most abundant proteins
set the colour scale for everything else.

`sample_order` is either `"clustering"` (the default, which lets the samples
arrange themselves and so doubles as a check that replicates group together) or
`"condition"`, which imposes your order and makes the figure comparable across
panels.

```{r heatmap-list, eval = FALSE}
# One heatmap per mode, all with the same settings
proteomics_heatmap_list(res$PCA_Input, modes = c("all", "any"),
                        scale_data = "row")
```

# Palettes

Every module accepts `palette`, and it is resolved by `paletteer`, so it takes
the `package::palette` form — a bare `"Set1"` is not enough, because paletteer
needs to know which package that name belongs to. The alternative is `colors`,
which takes the hex values directly.

```{r palettes, eval = FALSE}
volcano_highchart_list(res$DEPs_results,
                       palette = "ggsci::nrc_npg")[["B-A"]]

volcano_highchart_list(res$DEPs_results,
                       colors = list(Up = "#D55E00", Down = "#0072B2",
                                     `No Change` = "#BBBBBB"))[["B-A"]]
```

The explicit form is worth using for anything that will be published: it is the
only one that guarantees the same protein keeps the same colour across figures
made at different times, and the three shown here are colour-blind safe.

# Exporting

Static plots go through their own machinery:

```{r export, eval = FALSE}
proteomics_heatmap(res$PCA_Input, mode = "any",
                   export_path = tempdir(),
                   export_file = "heatmap.png",
                   plot_width = 10, plot_height = 8, export_dpi = 300)
```

Interactive ones carry export buttons — PNG, SVG, PDF — in the chart itself, and
`vignette("results-and-export")` covers saving a whole widget to a self-contained
HTML file.

```{r session}
sessionInfo()
```
